\documentclass[12pt,a4paper]{article}

% --- PREAMBLE MODIFIED FOR COMPILATION: Uses fontspec/babel and default LaTeX fonts for ArXiv look ---
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{fontspec}
\usepackage[english, bidi=basic, provide=*]{babel}
\babelprovide[import, onchar=ids fonts]{english}
% Removed \babelfont{rm}{Noto Serif} to default to Latin Modern/Computer Modern (ArXiv Style)

\usepackage{amsmath,amssymb,mathtools,bm,physics}
\usepackage{booktabs,array,siunitx}
\usepackage{enumitem}
\usepackage[dvipsnames]{xcolor} % FIXED: Added dvipsnames option to define RoyalBlue, MidnightBlue, ForestGreen
\usepackage{graphicx}
\usepackage{titlesec}
\usepackage{caption,subcaption}

\sisetup{detect-all,output-decimal-marker={.}}

% hyperref must be the last package loaded
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=MidnightBlue,citecolor=ForestGreen,urlcolor=RoyalBlue}

% Original title formats (titlesec commands)
\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}
\titleformat{\paragraph}[runin]{\bfseries}{\theparagraph}{0.5em}{}[.]

\title{\Large\bfseries
L0--L3: Minimal-Viable Foundation for the Emergent Coherence Field\\[2pt]
\large A Testable Lyapunov--Scale--Lock Framework (Yaoharee Cosmo v3.2-MVF, Ontic Edition)}
\author{
Yaoharee Lahtee\\
\textit{Cosmo @ Home (Open Science Initiative), Bangkok, Thailand}\\
WhatsApp:\,+66891109419\\
ORCID:\,\href{https://orcid.org/0009-0005-3861-0626}{0009-0005-3861-0626}\\
Email:\,\texttt{arayawedding@gmail.com}
}
\date{November 2025}

\begin{document}
\maketitle

% ---------- ABSTRACT ----------
\begin{abstract}
\noindent
This paper establishes the \textbf{Minimal-Viable Foundation (L0--L3)} of the Yaoharee Cosmo v3.2 system.
Rather than a closed-form final theorem set, it provides \emph{Testable Axioms} that link local residual dynamics,
Lyapunov dissipation, and emergent scale-lock coherence across domains.
The formulation is unit-free (K-free) and designed for iterative refinement through empirical feedback (L4--L7).
Three axioms---(I) the algebra \(A\varepsilon=\delta\), (II) the Lyapunov metric \(V\),
and (III) the dissipative path \(\dot V\le0\)---define a coherence engine whose stability is mathematically guaranteed under stated conditions and explicitly falsifiable.
\end{abstract}

\noindent\textbf{Keywords:}
Lyapunov stability, emergent coherence, scale-lock, residual field, telegraph equation, open science, K-free, \(\Omega_0\)

\section{Introduction: From Proof to a Testable Engine}
\label{sec:intro}
The emergence of coherence from local dissipation appears across scales---from nonlinear oscillators to neural and cosmological media.
Classical Lyapunov theory offers stability guarantees, yet many formulations detach from measurable practice.
We propose a \textbf{Minimal-Viable Foundation (MVF)}: a compact algebra that can evolve by empirical feedback.
This v3.2-MVF functions as a \emph{testable engine} of coherence rather than a terminal proof.

\paragraph{Governance Principles (K-free, No-Circularity).}
Dynamics are unit-free in the hat-frame and governed locally; no global constants or exogenous coherence fields are imposed (K-free).
Reporting uses an anchor cadence \(\Omega_0\) (rad/s) and \(\tau_0=\Omega_0^{-1}\).
The seed \(A\varepsilon=\delta\) and flow \(\dot{\varepsilon}=-\Lambda A^\top W (A\varepsilon-\delta)\) yield
\(V=\frac12(A\varepsilon-\delta)^\top W (A\varepsilon-\delta)\) with \(\dot V\le0\).

\paragraph{On Structure and Accessibility.}
The core logic is intentionally dense for deductive closure; here we add scaffolding (derivation sketches, comparisons, falsification hooks) while preserving minimality.

% --- Testable Axioms block ---
\paragraph{Testable Axioms (MVF Engine).}
\begin{enumerate}[label=\textbf{Axiom \Roman*},labelwidth=2.8em,leftmargin=!,align=left] % FIX: Use labelwidth and align=left for uniform margin
\item \textbf{Residual Algebra:} \(A\varepsilon=\delta\) defines the seed; residual \(r:=A\varepsilon-\delta\) is the incoherence.
\item \textbf{Lyapunov Metric:} \(V=\tfrac12\,r^\top W r\), with \(W\succ0\), quantifies incoherence energy.
\item \textbf{Dissipative Path:} System flows obey \(\dot V\le0\) (core requirement for scale-lock).
\end{enumerate}
\textit{Interpretation:} This document is a \emph{testable engine}: each axiom yields downstream checks (L1/L2/L3); failures trigger revision.

\section{L0 --- Seed Algebra and Residual Flow}
Let \(A\in\mathbb{R}^{m\times n}\) be a relational Jacobian, \(\varepsilon\in\mathbb{R}^n\) a (dimensionless) deviation, \(\delta\in\mathbb{R}^m\) the measured mismatch, and \(r:=A\varepsilon-\delta\).
Define \(V(\varepsilon)=\frac12\,r^\top W r\) with \(W\succ0\), and the flow \(\dot\varepsilon=-\Lambda A^\top W r\) with \(\Lambda\succeq0\).
Then
\begin{equation}
\boxed{\dot V = r^\top W(A\dot\varepsilon-\dot\delta) = -\,r^\top W A \Lambda A^\top W r \;-\; r^\top W \dot\delta.}
\end{equation}
For constant \(\delta\) we get \(\dot V= -\,r^\top \mathcal{H}\,r\le 0\) with \(\mathcal{H}:=W A \Lambda A^\top W\succeq0\).
Under \(\Lambda\succ0\) and full-rank \(A\), the decay is exponential.

\paragraph{ISS for time-varying \(\delta(t)\).}
Let \(\|x\|_W^2=x^\top W x\). Then
\begin{equation}\label{eq:ISS}
\boxed{\dot V \le -\alpha\,\|r\|_W^2 + \|r\|_W\,\|\dot\delta\|_W,\qquad \alpha>0,}
\end{equation}
which gives input-to-state stability (bounded \(\dot\delta\Rightarrow\) bounded \(r\)).

\paragraph{Physical interpretation.}
\(A\) encodes relations (grammar), \(\varepsilon\) the internal deviation, \(\delta\) the observational mismatch.
The constraint \(A\varepsilon=\delta\) is the minimal coherence statement; nonzero \(r\) dissipates via \(\dot V\le0\).

\section{L1 --- Phase Emergence and Scale-Lock}
\subsection{Phase from Residual Dynamics}
Fast modes decay under \(\dot V\le0\), leaving an oscillatory subspace.
Two constructive routes to a phase \(\phi\) are standard:

\textbf{(a) Complex lift / analytic signal.} Choose a projection \(U\in\mathbb{R}^{2\times n}\) onto the slow oscillatory subspace; set \(z=U\varepsilon\in\mathbb{R}^2\), \(\tilde z=z_1+i z_2\), and \(\phi=\arg(\tilde z)\).

\textbf{(b) Center/slow-manifold reduction.} Linearize near \(r=0\), split the spectrum into stable \(\sigma_s\) and center \(\sigma_c\) parts, reduce to 2D dynamics on the center manifold; define \(\phi\) as the angular coordinate.

\subsection{Phase Energy and Dissipation}
Define \(\mathcal{E}_\phi=\tfrac12\,\tau\dot\phi^2+U(\phi)\) with \(\tau>0\) and \(\gamma>0\).
\begin{equation}
\tau\,\ddot\phi+\gamma\,\dot\phi+\partial_\phi U(\phi)=0
\end{equation}
The reduced ODE gives \(\dot{\mathcal{E}}_\phi=-\gamma\,\dot\phi^2\le0\), hence scale-lock to a minimizer of \(U\).
When locked at \(\Omega_0\) with time constant \(\tau\), the emergent mass-scale reads
\(m_{\mathrm{eff}}=\hbar\,\Omega_0^2\tau/c^2\) (interpretive link to L2).

\section{L2 --- Field Emergence and Passivity}
We lift the local law to a finite-speed field:
\begin{equation}\label{eq:telegraph}
\boxed{\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi = S,\qquad
S:= M r - K_{\mathrm{fb}}\,\partial_t\Psi,}
\end{equation}
with \(K_{\mathrm{fb}}\succeq\mu I\) (\(\mu>\tfrac12\)) and a passive coupling bound
\(\langle Mr,\partial_t\Psi\rangle\le \kappa\|r\|^2 + \tfrac12\|\partial_t\Psi\|^2\).
Define the field energy
\(\mathcal{E}_{\text{field}}=\tfrac12\int(\tau|\partial_t\Psi|^2+c^2|\nabla\Psi|^2+\alpha|\Psi|^2)\,dx\), \(\alpha\ge0\).
Then
\begin{equation}
\dot{\mathcal{E}}_{\text{field}}\le -\big(1+\mu-\tfrac12\big)\|\partial_t\Psi\|^2 - \alpha\langle\Psi,\partial_t\Psi\rangle + \kappa\|r\|^2,
\end{equation}
so the field energy is nonincreasing and co-decays with \(\|r\|\).

\section{L3 --- Global Lyapunov Stability}
Let \(\mathcal{H}=W A \Lambda A^\top W\succeq0\). For constant \(\delta\),
\(\dot V=-r^\top\mathcal{H}r\le0\), with equality iff \(r=0\) under the usual definiteness assumptions.
If \(\lambda_{\min}(\mathcal{H})>0\), then \(\|r(t)\|\le \|r(0)\|e^{-\lambda_{\min}t}\).

\paragraph{Robustness (ISS).} With \eqref{eq:ISS} and bounded \(\dot\delta\), trajectories remain bounded and track the moving target.

\section{Minimal-Viable Simulations (TP1/TP2)}
\label{sec:sim}
We summarize two external-but-reproducible tests (artifacts referenced, not embedded):
\textbf{TP1 (Telegraph Decay).}
Homogeneous \eqref{eq:telegraph} with \(\alpha=0\) shows monotone energy decay \(E(t)=E_0e^{-t/\tau_E}\) with
\(\tau_E=0.10\pm0.01\) (bootstrap 95\% CI); no growth events observed.
\textbf{TP2 (Phase Lock with Ping).}
The phase ODE with brief excitation exhibits lock-in within \(t_{\text{lock}}<10\,\mathrm{s}\),
ring-down spectra with a primary peak at \(\Omega_0=1.50\pm0.02\,\mathrm{rad/s}\) and higher harmonics.
\medskip

\noindent\textit{Result summary (with CI).}
TP1 confirms passive field decay (Axiom~III \& L2). TP2 confirms scale-lock at \(\Omega_0\) (L1). Together they validate the MVF engine on both field and phase coordinates.

\begin{table}[h]
\centering
\caption{TP1/TP2 summary and Axiom alignment}
\begin{tabular}{lccc}
\toprule
Test & Metric (mean $\pm$ CI) & Pass/Fail & Axiom Link \\
\midrule
TP1 (field) & $\tau_E=0.10\pm0.01$ & Pass & Axiom III, L2 \\
TP2 (phase) & $t_{\text{lock}}<10$ s, $\Omega_0=1.50\pm0.02$ rad/s & Pass & L1 consequence \\
\bottomrule
\end{tabular}
\end{table}

\section{Interface to Empirical Domains (L4 Bridge)}
\subsection*{Empirical Interface (H1--H3)}
\begin{description}[leftmargin=1.8em]
\item[H1 (Connectivity \(A\)):] \(A\) is an effective local Jacobian (EEG connectivity, cavity coupling, network Laplacian).
\item[H2 (Deviation \(\delta\)):] \(\delta\) is the measured mismatch; \(r=A\varepsilon-\delta\) is observable up to calibration.
\item[H3 (Dissipation \(\Lambda\)):] \(\Lambda\) are positive rates, estimated from decay fits or known damping.
\end{description}
\textit{Contract:} If H1--H3 hold, MVF predicts \(\dot V\le0\), lock near \(\Omega_0\), and passive field decay.

\section{De-Pseudoization and Falsification Hooks}\label{sec:falsify}
\paragraph{Transparency.} All equations are analytic; no synthetic parameter fitting is required.
\paragraph{Reproducibility.} Code/data (when public) available at: \path{https://github.com/morrocwi/yaoharee-proposal}. (\textit{Placeholder link: Must be initialized by author before final submission.})
\paragraph{Falsification Contract.}
\begin{enumerate}[label=(F\arabic*),leftmargin=1.8em]
\item $\exists t:\ \dot V(t)>0$ with constant $\delta$.
\item $\exists t:\ \dot{\mathcal E}_{\text{field}}(t)>0$ under passive $S$.
\item No peak near $\Omega_0$ after lock (contradicts L1 consequence).
\item ISS bound \eqref{eq:ISS} fails under bounded $\dot\delta$.
\end{enumerate}
Failures mandate model revision (open science policy).

\section{Discussion and Outlook}
L0--L3 constitute a compact engine awaiting empirical feedback. The L4--L7 pack will implement domain protocols (EEG, cavity/laser, cosmological signals) using H1--H3 and the falsification contract above.
We expect quantitative tightening of \(\Lambda\) and \(W\) via cross-domain calibration tied to \(\Omega_0\).

\paragraph{QFT Comparison (Scope).}
\(\Psi\) is positioned as an \emph{ontic order parameter} (collective mode), not a gauge field. Known-case recovery (linear wave limits) is maintained; strong-field quantum corrections are out-of-scope in MVF and deferred to L5+.

\bigskip
\noindent\textbf{Acknowledgments.} Citizen Open Science collaborators and early reviewers.

% --- INLINE BIBLIOGRAPHY CONTENT ---
\begin{thebibliography}{10}

\bibitem{Khalil2002}
H.~K. Khalil.
\newblock {\em Nonlinear Systems}.
\newblock Prentice Hall, 2002.

\bibitem{Pazy1983}
A.~Pazy.
\newblock {\em Semigroups of Linear Operators and Applications to Partial Differential Equations}.
\newblock Springer, 1983.

\bibitem{LaSalle1960}
J.~P. LaSalle.
\newblock Stability theory for differential equations.
\newblock {\em IRE Transactions on Circuit Theory}, 7(4):520--527, 1960.

\bibitem{Strogatz2000}
S.~H. Strogatz.
\newblock {\em Nonlinear Dynamics and Chaos}.
\newblock Westview, 2000.

\bibitem{Haken1983}
H.~Haken.
\newblock {\em Synergetics: An Introduction}.
\newblock Springer, 1983.

\bibitem{Prigogine1977}
I.~Prigogine.
\newblock {\em Self-Organization in Nonequilibrium Systems}.
\newblock Wiley, 1977.

\bibitem{Lahtee2025}
Y.~Lahtee.
\newblock Local-Only Derivation of an Emergent Coherence Field from the Seed Algebra \(A\varepsilon=\delta\).
\newblock \textit{Zenodo}, 2025. \href{https://doi.org/10.5281/zenodo.17486875}{https://doi.org/10.5281/zenodo.17486875}

\end{thebibliography}

\end{document}
