% ===================== Chapter 1 =====================

\section{Notation \& Preliminaries}

\subsection{Core Symbols}
Vectors are column vectors. For a matrix $M$, $M^\top$ is transpose.
All quantities in this monograph are made \emph{dimensionless} first.

\begin{center}
\begin{tabular}{ll}
$\omega,\kappa,b,\rho$ & primitive variables (frequency-like, curvature/diffusion-like, coupling, density-like)\\
$\Omega_0,B_0,C_0,\rho_0$ & reference anchors (units) for $(\omega,\kappa,b,\rho)$\\
$\hat{\omega},\hat{\kappa},\hat b,\hat{\rho}$ & near-unity variables (dimensionless)\\
$\Pi_i$ & mother ratios (Buckingham-$\Pi$ style)\\
$\varepsilon$ & residual state (internal)\\
$\delta$ & measured deviation (external)\\
$A$ & Jacobian of $\log\Pi$ at the unity state (local linear map)\\
$W\succeq0$ & data-weight (often inverse covariance) \\
$\Lambda\succeq0$ & step/metric in residual space \\
\end{tabular}
\end{center}

\subsection{Dimensionless Separation (All-Hat First)}
Fix a single frequency anchor $\Omega_0>0$ and define
\[
B_0:=\Omega_0^2,\qquad C_0:=\Omega_0,\qquad \rho_0:=\Omega_0^{-1},\qquad 
\tau_0:=\Omega_0^{-1}.
\]
Declare the dimensionless \emph{near-unity} variables
\[
\hat\omega=\frac{\omega}{\Omega_0}=1+\varepsilon_\omega,\quad
\hat\kappa=\frac{\kappa}{B_0}=1+\varepsilon_\kappa,\quad
\hat b=\frac{b}{C_0}=1+\varepsilon_b,\quad
\hat\rho=\frac{\rho}{\rho_0}=1+\varepsilon_\rho,\quad \|\varepsilon\|\ll1.
\]
All subsequent algebra is performed in hats; results are unit/scale-invariant.

\subsection{Mother Ratios and Log-Linearization}
Define the mother ratios around unity,
\[
\Pi_1=\frac{\kappa}{\omega^2}=1+\delta_1,\quad
\Pi_2=\frac{b}{\omega}=1+\delta_2,\quad
\Pi_3=\omega\rho=1+\delta_3,\quad
\Pi_4=\frac{\kappa b}{C\,\Omega_0}=1+\delta_4,
\]
and use $\log(1+x)=x+O(x^2)$ to obtain the local log-linear system
\[
A\varepsilon=\delta,\qquad
A=
\begin{pmatrix}
-2&1&0&0\\
-1&0&1&0\\
1&0&0&1\\
0&1&1&0
\end{pmatrix}.
\]
This is the \textbf{Seed} relation of the entire framework.

\subsection{Invertibility and Closed-Form Recovery}
Since $\det A=-3\neq0$, the map is locally invertible and
\[
A^{-1}=\frac{1}{3}
\begin{pmatrix}
-1&-1&0&1\\
1&-2&0&2\\
-1&2&0&1\\
1&1&3&-1
\end{pmatrix},\qquad \varepsilon=A^{-1}\delta.
\]
Componentwise,
\[
\varepsilon_\omega=\tfrac{-\delta_1-\delta_2+\delta_4}{3},\quad
\varepsilon_\kappa=\tfrac{\delta_1-2\delta_2+2\delta_4}{3},\quad
\varepsilon_b=\tfrac{-\delta_1+2\delta_2+\delta_4}{3},\quad
\varepsilon_\rho=\tfrac{\delta_1+\delta_2+3\delta_3-\delta_4}{3}.
\]

\subsection{Norms, Inner Products, and Weights}
We use the standard Euclidean inner product on $\mathbb{R}^n$,
$\langle x,y\rangle=x^\top y$, with weighted seminorm induced by $W\succeq0$,
\[
\|r\|_{W}^2 := r^\top W r.
\]
The residual-space Lyapunov candidate is
\[
V(\varepsilon;\delta):=\frac12\,(A\varepsilon-\delta)^\top W (A\varepsilon-\delta).
\]

\subsection{Residual Dynamics (Preview)}
The canonical gradient flow on the residual manifold reads
\[
\dot\varepsilon=-\Lambda A^\top W\,(A\varepsilon-\delta),\qquad 
\dot V=- (A\varepsilon-\delta)^\top W A \Lambda A^\top W (A\varepsilon-\delta)\le 0.
\]
All quantities are already dimensionless; one may rescale time by $\hat t=t/\tau_0$ and absorb $\tau_0$ into $\Lambda$ if desired.

\subsection{Operator Substitution (Preview)}
When embedding $\varepsilon(\mathbf x,t)$ into a field $\Psi$ and applying the symbolic map
\[
\omega\mapsto i\partial_t,\qquad \kappa\mapsto -\nabla^2,
\]
we will obtain a dissipative/telegraph-like core for $\Psi$ in later chapters.
\qedhere
% ===================== End Chapter 1 =====================

\begin{tcolorbox}[title=Compatibility Note (Units and Symbols)]
Throughout the monograph we adopt the rad/s convention:
\[
\Omega_0=\omega_0,\qquad
B_0=\Omega_0^2,\quad C_0=\Omega_0,\quad \rho_0=\Omega_0^{-1},\quad \tau_0=\Omega_0^{-1}.
\]
When reporting frequencies in Hz for figures or tables, define
\(f_0:=\Omega_0/(2\pi)\). To avoid symbol collision with Hz-based narratives,
the anchor that appears in the auxiliary ratio is denoted \(\mathcal C\), so that
\[
\Pi_4=\frac{\kappa\, b}{\mathcal C\,\Omega_0}.
\]
This keeps all algebra unit-free and invariant while allowing practical
Hz-based reporting without altering any core formulas.
\end{tcolorbox}


% ===================== Chapter 2 =====================

\section{Dimensionless Separation (All-Hat First)}

\subsection{H0 Anchors (Nondimensionalization)}
Let $[\omega]=T^{-1}$, $[\kappa]=T^{-2}$, $[b]=T^{-1}$, $[\rho]=T^{+1}$. 
Fix a characteristic frequency $\Omega_0>0$ and define
\[
B_0=\Omega_0^2,\quad C_0=\Omega_0,\quad \rho_0=\Omega_0^{-1},\quad C=\Omega_0^2,\quad \tau_0=\Omega_0^{-1}.
\]
All algebra henceforth uses these anchors. 
\hfill{\small (cf. H0 anchor set)} 
``{\footnotesize :contentReference[oaicite:0]{index=0}}''

\subsection{H1 Near-One Dimensionless Variables}
\[
\hat\omega=\omega/\Omega_0=1+\varepsilon_\omega,\quad
\hat\kappa=\kappa/B_0=1+\varepsilon_\kappa,\quad
\hat b=b/C_0=1+\varepsilon_b,\quad
\hat\rho=\rho/\rho_0=1+\varepsilon_\rho,\quad \|\varepsilon\|\ll1.
\]
\textbf{Unit-change invariance.} If $t\mapsto\tilde t= s t$ with $s>0$, then $\tilde\omega=\omega/s$, $\tilde\Omega_0=\Omega_0/s$, etc., hence all hats are invariant $\Rightarrow$ algebra in hats is scale/unit invariant.
\hfill{\small (near-one hats \& invariance)} 
``{\footnotesize :contentReference[oaicite:1]{index=1}}'' 
``{\footnotesize :contentReference[oaicite:2]{index=2}}''

\subsection{H2 Mother Ratios and the Scale Generator}
Scale generator $S_\lambda:(\kappa,\omega,b,\rho)\mapsto(\lambda^2\kappa,\lambda\omega,\lambda b,\lambda^{-1}\rho)$ yields the invariant basis 
\(
\Pi_1=\kappa/\omega^2,\ \Pi_2=b/\omega,\ \Pi_3=\omega\rho
\),
and an anchor-assisted near-one ratio 
\(
\Pi_4=\kappa b/(C\Omega_0)
\).
We write the near-unity forms
\[
\Pi_1=1+\delta_1,\quad \Pi_2=1+\delta_2,\quad \Pi_3=1+\delta_3,\quad \Pi_4=1+\delta_4.
\]
\hfill{\small (scale group \& invariant basis)} 
``{\footnotesize :contentReference[oaicite:3]{index=3}}'' 
``{\footnotesize :contentReference[oaicite:4]{index=4}}''

\subsection{H3 Log-Linearization to the Seed System}
Using $\log(1+x)=x+O(x^2)$ componentwise around the unity state gives the local linear system
\[
A\varepsilon=\delta,\qquad
A=
\begin{pmatrix}
-2&1&0&0\\
-1&0&1&0\\
1&0&0&1\\
0&1&1&0
\end{pmatrix},
\qquad \det A=-3\neq 0.
\]
\hfill{\small (matrix form and $\det A$)} 
``{\footnotesize :contentReference[oaicite:5]{index=5}}''

\subsection{H4 Exact Inverse and Closed-Form Recovery}
Since $\det A\neq0$, the inverse exists and
\[
A^{-1}=\tfrac{1}{3}
\begin{pmatrix}
-1&-1&0&1\\
1&-2&0&2\\
-1&2&0&1\\
1&1&3&-1
\end{pmatrix},\qquad \varepsilon=A^{-1}\delta,
\]
so that
\[
\varepsilon_\omega=\tfrac{-\delta_1-\delta_2+\delta_4}{3},\quad
\varepsilon_\kappa=\tfrac{\delta_1-2\delta_2+2\delta_4}{3},\quad
\varepsilon_b=\tfrac{-\delta_1+2\delta_2+\delta_4}{3},\quad
\varepsilon_\rho=\tfrac{\delta_1+\delta_2+3\delta_3-\delta_4}{3}.
\]
\hfill{\small (closed forms)} 
``{\footnotesize :contentReference[oaicite:6]{index=6}}''

\subsection{H5 Recovery to Physical (for Examples)}
\[
\omega=\Omega_0(1+\varepsilon_\omega),\quad
\kappa=B_0(1+\varepsilon_\kappa),\quad
b=C_0(1+\varepsilon_b),\quad
\rho=\rho_0(1+\varepsilon_\rho).
\]
Indicators (symmetry-preserving / weak symmetry-breaking) may be defined from these recoveries.
\hfill{\small (recovery \& $\beta$-indicators)} 
``{\footnotesize :contentReference[oaicite:7]{index=7}}'' 
``{\footnotesize :contentReference[oaicite:8]{index=8}}''

\subsection{Quadratic Remainder \& Validity Radius}
Local log-linearization admits a quadratic remainder: 
\(
\log F(1+\varepsilon)=J\varepsilon+R(\varepsilon),\ \|R(\varepsilon)\|_2\le \tfrac{1}{2}M\|\varepsilon\|_2^2
\).
Hence the linear closure holds inside a radius governed by $M$; sensitivity obeys 
\(
\|\Delta\varepsilon\|\le\|A^{-1}\|\,\|\Delta\delta\|
\).
\hfill{\small (quadratic bound \& sensitivity)} 
``{\footnotesize :contentReference[oaicite:9]{index=9}}'' 
``{\footnotesize :contentReference[oaicite:10]{index=10}}''

\subsection{Overdetermined Closures (Optional)}
With $m\ge4$ near-one relations $G\varepsilon\simeq\delta$ and a weight $W\succeq0$,
\[
\hat\varepsilon=\arg\min_\varepsilon\|W^{1/2}(G\varepsilon-\delta)\|_2^2=(G^TWG)^{-1}G^TW\delta,\quad
\hat\varepsilon_\lambda=(G^TWG+\lambda I)^{-1}G^TW\delta.
\]
\hfill{\small (WLS/Tikhonov in the hat-frame)} 
``{\footnotesize :contentReference[oaicite:11]{index=11}}''

\subsection{(Box) Domain-Agnostic Normalization}
A global hat-frame may also be declared by spatial/temporal/field units $(\ell_0,t_0,\Psi_0)$ and then suppressing hats in the sequel; one may absorb a dimensionless Planck factor $\hat\hbar$ into prefactors as needed.
\hfill{\small (global normalization)} 
``{\footnotesize :contentReference[oaicite:12]{index=12}}''

\subsection{Recipe (One-Pass)}
(i) nondimensionalize (H0–H1); (ii) form ratios \& linearize (H2–H3);
(iii) solve $\varepsilon=A^{-1}\delta$ (H4–H5); (iv) evaluate indicators; (v) bound errors/sensitivity; (vi) add WLS/Tikhonov gates if needed.
\hfill{\small (H15 step recipe)} 
``{\footnotesize :contentReference[oaicite:13]{index=13}}''

\paragraph{Summary.} บทนี้สถาปนา “กรอบหมวก” ที่ไร้มิติและอินแวเรียนซ์อย่างเคร่งครัด—ตั้งแต่ anchor → hats → mother ratios → ลอค–ลิเนียร์ไรซ์ → คำตอบย้อนกลับ—พร้อมกรอบความถูกต้องอันดับสองและสถาปัตยกรรมปิดงาน (H15) ที่จะใช้ซ้ำทั้งเล่ม
\qed
% ===================== End Chapter 2 =====================

% ===================== Chapter 3 =====================
\section{Lie-Scale Group and Invariants}

\subsection{The Lie Scale Action}
Let the primitive variables be $(\kappa,\omega,b,\rho)$, already nondimensional
(see Chapter~2). Consider the one-parameter scale action
\begin{equation}
S_\lambda:\ (\kappa,\omega,b,\rho)\ \mapsto\ (\lambda^{2}\kappa,\ \lambda\,\omega,\ \lambda\,b,\ \lambda^{-1}\rho),
\qquad \lambda>0.
\label{eq:S_lambda}
\end{equation}
This action is smooth, multiplicative in $\lambda$, and preserves the hat-frame
constructed in Chapter~2.

\paragraph{Infinitesimal generator.}
Write $x=(\kappa,\omega,b,\rho)^\top$ and define the generator $L$ by
\(
\frac{d}{d(\log\lambda)}\,x(\lambda)\big|_{\lambda=1}=L\,x.
\)
From \eqref{eq:S_lambda},
\begin{equation}
L=\mathrm{diag}(2,\,1,\,1,\,-1).
\label{eq:generator_L}
\end{equation}

\subsection{Scale Invariants and the Nullspace Principle}
Define the mother ratios (scale invariants)
\begin{equation}
\Pi_1=\frac{\kappa}{\omega^2},\qquad
\Pi_2=\frac{b}{\omega},\qquad
\Pi_3=\omega\,\rho,
\label{eq:Pi_basic}
\end{equation}
and, when an anchor $C\Omega_0$ is employed (cf. Chapter~2), an auxiliary
near-one ratio
\begin{equation}
\Pi_4=\frac{\kappa\,b}{C\,\Omega_0}.
\label{eq:Pi_aux}
\end{equation}
Each $\Pi_j$ is invariant under \eqref{eq:S_lambda}. In logarithmic coordinates
$y=\log x=(\log\kappa,\log\omega,\log b,\log\rho)^\top$, the generator
\eqref{eq:generator_L} induces the linear form
\[
\frac{d}{d(\log\lambda)}\,y(\lambda)\Big|_{\lambda=1}
=\begin{pmatrix}2\\1\\1\\-1\end{pmatrix} =: \ell,
\]
and the invariance of $\Pi$ is equivalent to the orthogonality constraint
\begin{equation}
\nabla_y \big(\log \Pi_j\big)\cdot \ell=0\quad\text{for each }j.
\label{eq:null_constraint}
\end{equation}
Hence, the gradients $\{\nabla_y\log\Pi_j\}$ lie in the nullspace of the
rank-one operator $y\mapsto \ell^\top y$, and the space of independent invariants
has dimension $4-1=3$.

\subsection{Jacobian of \texorpdfstring{$\log\Pi$}{logPi} and the Local Linear Map}
Let $\Pi=(\Pi_1,\Pi_2,\Pi_3,\Pi_4)^\top$ with the near-unity parameterization
\[
\Pi_i=1+\delta_i,\qquad \|\delta\|\ll 1,
\]
and let $\varepsilon=(\varepsilon_\omega,\varepsilon_\kappa,\varepsilon_b,\varepsilon_\rho)^\top$
collect the near-one residuals of the hat-variables
(Chapter~2). The Jacobian of $\log \Pi$ at the unity state (\emph{i.e.},
$\varepsilon=0$) yields the local linear map
\begin{equation}
A\,\varepsilon=\delta,\qquad
A=
\begin{pmatrix}
-2&1&0&0\\
-1&0&1&0\\
1&0&0&1\\
0&1&1&0
\end{pmatrix},\qquad \det A=-3\neq 0,
\label{eq:seed_system}
\end{equation}
which is precisely the seed system used in Chapter~2.

\subsection{Local Log-Linearization Theorem}
\begin{theorem}[Local log-linearization with quadratic remainder]
Let $\Pi(\varepsilon)$ be defined from the hat-variables as in Chapter~2 and
\eqref{eq:Pi_basic}--\eqref{eq:Pi_aux}. Then, in a neighborhood of the unity
state $\varepsilon=0$,
\begin{equation}
\log \Pi(1+\varepsilon)\ =\ A\,\varepsilon\ +\ R(\varepsilon),
\qquad \|R(\varepsilon)\|_2\ \le\ \tfrac12\,M\,\|\varepsilon\|_2^{2},
\label{eq:loglin_remainder}
\end{equation}
for some constant $M$ depending on second derivatives of $\log\Pi$ on the
neighborhood. Consequently, the inverse map satisfies
\(
\|\Delta\varepsilon\|\le \|A^{-1}\|\,\|\Delta\delta\|
\)
whenever $\|\varepsilon\|$ is small enough for \eqref{eq:loglin_remainder}
to hold.
\end{theorem}
\begin{proof}[Sketch]
Taylor-expand $\log\Pi$ about $\varepsilon=0$; the first-order term is the Jacobian
$A$ in \eqref{eq:seed_system}. The Lagrange form of the remainder yields the
quadratic bound with $M$ controlled by the supremum of the operator norm of the
Hessian of $\log\Pi$ on the neighborhood. The sensitivity estimate follows by
applying $A^{-1}$ to the first-order relation and bounding the remainder.
\end{proof}

\subsection{Rank Considerations and Independence}
Because \eqref{eq:S_lambda} is generated by a single parameter, the linear
constraint \eqref{eq:null_constraint} removes exactly one dimension from the
four logarithmic coordinates. Therefore, at most three independent scale
invariants exist locally, aligning with the triple $(\Pi_1,\Pi_2,\Pi_3)$ in
\eqref{eq:Pi_basic}. The auxiliary ratio $\Pi_4$ in \eqref{eq:Pi_aux} is near-one
by construction (anchor-assisted) and is convenient for conditioning but not
independent of the first three at the exact unity state.

\subsection{Consequences for the Seed System}
Equation \eqref{eq:seed_system} is canonical for three reasons:
(i) it is derived from the Lie scale symmetry via $\log$ coordinates;
(ii) it enjoys unit/scale invariance inherited from the hat-frame
(Chapter~2); and (iii) its invertibility ($\det A\neq 0$) provides the closed-form
recovery $\varepsilon=A^{-1}\delta$ used throughout the monograph.

\paragraph{Takeaway.} The Lie-scale viewpoint explains \emph{why} the mother
ratios close linearly near unity and \emph{how} the matrix $A$ is fixed by
symmetry, not by ad-hoc modeling.
% ===================== End Chapter 3 =====================

% ===================== Chapter 4 =====================
\section{Mother Ratios and Local Log-Linearization (Seed System)}

\subsection{Near-Unity Variables (Dimensionless Hats)}
We work entirely in the dimensionless hat-frame: 
\[
\hat\omega=\omega/\Omega_0=1+\varepsilon_\omega,\quad
\hat\kappa=\kappa/B_0=1+\varepsilon_\kappa,\quad
\hat b=b/C_0=1+\varepsilon_b,\quad
\hat\rho=\rho/\rho_0=1+\varepsilon_\rho,\quad \|\varepsilon\|\ll1.
\]
Unit changes $t\mapsto s\,t$ co-scale the anchors so all hat-variables remain invariant (algebra in hats is unit/scale invariant). 
{\small \textit{(H1 hats \& unit invariance)} \,}:contentReference[oaicite:0]{index=0}\,:contentReference[oaicite:1]{index=1}

\subsection{Mother Ratios from the Scale Generator}
Under the scale action $S_\lambda:(\kappa,\omega,b,\rho)\mapsto(\lambda^2\kappa,\lambda\omega,\lambda b,\lambda^{-1}\rho)$, the following three are invariant:
\[
\Pi_1=\frac{\kappa}{\omega^2},\qquad
\Pi_2=\frac{b}{\omega},\qquad
\Pi_3=\omega\rho,
\]
and we employ an anchor-assisted near-one ratio
\[
\Pi_4=\frac{\kappa b}{C\,\Omega_0}.
\]
We adopt the near-unity parameterization $\Pi_i=1+\delta_i$ with $\|\delta\|\ll1$. 
{\small \textit{(Scale group \& invariant basis; anchor-assisted \(\Pi_4\))} }:contentReference[oaicite:2]{index=2}\,:contentReference[oaicite:3]{index=3}

\subsection{Log-Linearization to the Seed System}
Using $\log(1+x)=x+O(x^2)$ componentwise around the unity state yields
\[
-2\varepsilon_\omega+\varepsilon_\kappa=\delta_1,\quad
-\varepsilon_\omega+\varepsilon_b=\delta_2,\quad
\varepsilon_\omega+\varepsilon_\rho=\delta_3,\quad
\varepsilon_\kappa+\varepsilon_b=\delta_4,
\]
which we write compactly as
\[
A\varepsilon=\delta,\qquad
A=
\begin{pmatrix}
-2&1&0&0\\
-1&0&1&0\\
1&0&0&1\\
0&1&1&0
\end{pmatrix},\qquad \det A=-3\neq0.
\]
{\small \textit{(Matrix form and determinant)}} :contentReference[oaicite:4]{index=4}

\subsection{Closed-Form Inverse and Component Formulas}
Since $\det A\neq0$, the inverse exists and
\[
A^{-1}=\tfrac{1}{3}
\begin{pmatrix}
-1&-1&0&1\\
1&-2&0&2\\
-1&2&0&1\\
1&1&3&-1
\end{pmatrix},\qquad \varepsilon=A^{-1}\delta,
\]
so that
\[
\varepsilon_\omega=\tfrac{-\delta_1-\delta_2+\delta_4}{3},\quad
\varepsilon_\kappa=\tfrac{\delta_1-2\delta_2+2\delta_4}{3},\quad
\varepsilon_b=\tfrac{-\delta_1+2\delta_2+\delta_4}{3},\quad
\varepsilon_\rho=\tfrac{\delta_1+\delta_2+3\delta_3-\delta_4}{3}.
\]
{\small \textit{(Exact inverse \& closed forms)}} :contentReference[oaicite:5]{index=5}

\subsection{Quadratic Remainder and Sensitivity}
Let $\delta=\log\Pi(1+\varepsilon)$ be the componentwise log-residual. Then
\[
\log\Pi(1+\varepsilon)=A\varepsilon+R(\varepsilon),\qquad 
\|R(\varepsilon)\|_2\le\tfrac12\,M\,\|\varepsilon\|_2^2,
\]
and perturbations satisfy the standard bound 
$\|\Delta\varepsilon\|\le\|A^{-1}\|\,\|\Delta\delta\|$.
{\small \textit{(Quadratic remainder \& sensitivity)}} :contentReference[oaicite:6]{index=6}\,:contentReference[oaicite:7]{index=7}

\subsection{Overdetermined Closures (Optional)}
For $m\ge4$ relations $G\varepsilon\simeq\delta$ with a weight $W\succeq 0$,
\[
\hat\varepsilon=\arg\min_\varepsilon\|W^{1/2}(G\varepsilon-\delta)\|_2^2=(G^TWG)^{-1}G^TW\delta,\qquad
\hat\varepsilon_\lambda=(G^TWG+\lambda I)^{-1}G^TW\delta,
\]
providing WLS and Tikhonov-regularized estimates in the same hat-frame.
{\small \textit{(WLS/Tikhonov)}} :contentReference[oaicite:8]{index=8}

\subsection{Recovery to Physical Variables (for Examples)}
When needed for exemplification,
\[
\omega=\Omega_0(1+\varepsilon_\omega),\quad
\kappa=B_0(1+\varepsilon_\kappa),\quad
b=C_0(1+\varepsilon_b),\quad
\rho=\rho_0(1+\varepsilon_\rho),
\]
matching the anchors declared in Chapter~2.
{\small \textit{(Recovery with anchors)}} :contentReference[oaicite:9]{index=9}

\paragraph{Takeaway.}
The seed system \(A\varepsilon=\delta\) is a symmetry-fixed, scale-invariant closure of the mother ratios near unity. Its explicit inverse, remainder control, and WLS/Tikhonov extensions make it a robust bridge from measurements \(\delta\) to internal state \(\varepsilon\) in data-rich, multi-domain settings.
% ===================== End Chapter 4 =====================

% ===================== Chapter 5 =====================
\section{Residual Dynamics and Lyapunov Dissipation}

\subsection{Setup: Residual and Weighted Energy}
Let $r:=A\varepsilon-\delta$ be the componentwise log-residual from Chapter~4.
Choose $W\succeq 0$ and $\Lambda\succeq 0$ (often positive definite in practice)
and define the residual-space Lyapunov candidate
\begin{equation}
V(\varepsilon) := \tfrac12\, r^\top W r .
\label{eq:V_def}
\end{equation}
This is the same energy used in the notes and the v5.0 pack.{\small \,:contentReference[oaicite:0]{index=0}\,:contentReference[oaicite:1]{index=1}}

\subsection{Gradient-Flow Residual Dynamics}
Assume $W\succ0$ and $\Lambda\succ0$. The canonical residual-space gradient flow is
\begin{equation}
\dot\varepsilon \;=\; -\,\Lambda\,A^\top W\, (A\varepsilon-\delta)\,,
\label{eq:gf_flow}
\end{equation}
with energy
\begin{equation}
V(\varepsilon)=\tfrac12(A\varepsilon-\delta)^\top W (A\varepsilon-\delta)\,.
\label{eq:V_again}
\end{equation}
Along \eqref{eq:gf_flow},
\begin{equation}
\dot V \;=\; -\,(A\varepsilon-\delta)^\top W\, A\,\Lambda\,A^\top W\, (A\varepsilon-\delta) \;\le\; 0.
\label{eq:Vdot_le0}
\end{equation}
Equations \eqref{eq:gf_flow}--\eqref{eq:Vdot_le0} match the recorded formulation.{\small \,:contentReference[oaicite:2]{index=2}\,:contentReference[oaicite:3]{index=3}}

\subsection{Time Rescaling and Dimensionless Form}
Let $\hat t := t/\tau_0$ (see Chapter~2 for anchors). Then
\[
\frac{d\varepsilon}{d\hat t} \;=\; -\,(\tau_0\Lambda)\,A^\top W\, (A\varepsilon-\delta)\,.
\]
Absorbing $\tau_0$ into $\hat\Lambda:=\tau_0\Lambda$ keeps the form \eqref{eq:gf_flow} unchanged and entirely dimensionless.{\small \,:contentReference[oaicite:4]{index=4}}

\subsection{Well-Posedness: Choices of $W$ and $\Lambda$}
The matrix $W$ encodes data fidelity (e.g.\ inverse covariance); $\Lambda$ sets the metric/step in residual space. With $W\succ0$ and $\Lambda\succ0$, \eqref{eq:Vdot_le0} shows strict dissipation unless $r=0$. In practice, one may:
\begin{itemize}
\item use diagonal $W$ for heteroscedastic residuals or $W=\Sigma^{-1}$ for correlated noise;
\item precondition via $\Lambda$ (e.g.\ diagonal or quasi-Newton) to improve conditioning of $A^\top W A$.
\end{itemize}
Identifiability under overdetermined closures requires $\mathrm{rank}(G)=4$ (or $\mathrm{rank}(G^\top W G)=4$).{\small \,:contentReference[oaicite:5]{index=5}\,:contentReference[oaicite:6]{index=6}\,:contentReference[oaicite:7]{index=7}}

\subsection{Rates and Conditioning}
Linearizing \eqref{eq:gf_flow} about a fixed $\delta$ yields
\[
\dot\varepsilon = -\,\Lambda A^\top W A\,\varepsilon + \Lambda A^\top W\,\delta .
\]
The homogeneous part decays with a rate governed by the spectrum of 
$H:=\Lambda^{1/2} (A^\top W A) \Lambda^{1/2}\succeq 0$.
If $\lambda_{\min}(H)>0$, then
\[
V(t)\le V(0)\,e^{-2\,\lambda_{\min}(H)\, t}\quad\text{and}\quad \|r(t)\|_W \le \|r(0)\|_W\,e^{-\lambda_{\min}(H)\, t}.
\]
(See also the sensitivity and remainder bounds compiled earlier.){\small \,:contentReference[oaicite:8]{index=8}}

\subsection{Discrete-Time Realizations}
For a stepsize $\eta>0$, the explicit scheme
\begin{equation}
\varepsilon^{k+1} \;=\; \varepsilon^k \;-\; \eta\,\Lambda\,A^\top W\,(A\varepsilon^k-\delta)
\label{eq:gd_scheme}
\end{equation}
is stable if $\eta<2/\lambda_{\max}(H)$ (same $H$ as above). One may add line search or use a Gauss--Newton update when $A$ is replaced by a Jacobian $J$ in a nonlinear extension.

\subsection{Robust Extensions (Data-Rich)}
When $m\ge 4$ relations $G\varepsilon\simeq\delta$ are available, a WLS or Tikhonov gate may be used to initialize $\varepsilon$ near the minimizer before running \eqref{eq:gd_scheme}:
\[
\hat\varepsilon=(G^\top W G)^{-1}G^\top W\delta,\qquad
\hat\varepsilon_\lambda=(G^\top W G+\lambda I)^{-1}G^\top W\delta.
\]
These closures are stated in the source notes and dovetail with the gradient-flow layer.{\small \,:contentReference[oaicite:9]{index=9}}

\subsection{Interface to the Field Layer (Preview)}
The residual-space dissipation \eqref{eq:Vdot_le0} is consistent with the field-level, causal telegraph dynamics where the dispersion satisfies $\Re s<0$, ensuring finite-rate decay. This consistency is recorded in both the working notes and the consolidated v5.0 pack.{\small \,:contentReference[oaicite:10]{index=10}\,:contentReference[oaicite:11]{index=11}}

\paragraph{Takeaway.}
Equation \eqref{eq:gf_flow} provides a symmetry-consistent, dimensionless, and provably dissipative evolution from measurements $\delta$ to state $\varepsilon$. Its conditioning and rates are fully determined by $(A,W,\Lambda)$, and its structure aligns with the causal telegraph layer used later in the monograph.
% ===================== End Chapter 5 =====================

% ===================== Chapter 6 =====================
\section{Unified Telegraph Dynamics (Stable, Causal)}

\subsection{Dimensionless setting and notation}
We adopt the global, dimensionless normalization introduced earlier:
choose $(\ell_0,t_0,\Psi_0)$ and report all equations in hats thereafter
(hats omitted for readability). A dimensionless Planck factor $\hat\hbar$
may be absorbed into energy prefactors.%
{\small \,:contentReference[oaicite:0]{index=0}}

\paragraph{Fourier--Laplace convention.}
Time derivatives map to $s\in\mathbb C$ and Laplacian to $-|k|^2$
under $(\partial_t,\nabla^2)\mapsto(s,-|k|^2)$; $\tau>0$ and $c>0$
denote the (dimensionless) relaxation time and finite propagation speed.%
{\small \,:contentReference[oaicite:1]{index=1}}

\subsection{Telegraph-stable core and dispersion}
The unified (dimensionless) telegraph core for the coherence field $\Psi$ is
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi - c^2\nabla^2\Psi
\;=\; \mathcal{F}(\Psi;\,a,\rho,J_{\!fb}) \;+\; \Gamma(\Psi_q,\varphi),
\label{eq:telegraph_core}
\end{equation}
where the right-hand side collects divergence of fluxes and feedback currents
(and may include weakly nonlinear source terms), together with causal stochastic
forcing $\Gamma$.%
{\small \,:contentReference[oaicite:2]{index=2}}

\paragraph{Dispersion and stability window.}
In the homogeneous limit of \eqref{eq:telegraph_core}, the dispersion relation is
\begin{equation}
\tau s^2+s+c^2|k|^2=0
\quad\Longrightarrow\quad
s(k)=\frac{-1\pm\sqrt{1-4\tau c^2|k|^2}}{2\tau}.
\label{eq:dispersion}
\end{equation}
For $4\tau c^2|k|^2>1$ (underdamped), $\Re s=-\tfrac{1}{2\tau}$; otherwise
$\,s\in\mathbb R_{<0}$ (over/critical damping). In all cases $\Re s<0$, ensuring
finite-rate decay and causality.%
{\small \,:contentReference[oaicite:3]{index=3}\,:contentReference[oaicite:4]{index=4}}

\subsection{Stochastic assumptions (causality)}
All stochastic sources are zero-mean and causal, with white-noise correlations
of the form
\[
\langle \Gamma(\cdot)\rangle=0,\qquad
\langle \xi(t)\rangle=0,\qquad
\langle \xi(t)\,\xi(t')\rangle=2D\,\delta(t-t'),\quad D\ge0.
\]
These conventions keep the PDE layer compatible with the residual-space
Lyapunov dissipation.%
{\small \,:contentReference[oaicite:5]{index=5}\,:contentReference[oaicite:6]{index=6}}

\subsection{Boundary conditions and well-posedness}
To guarantee existence/uniqueness for the telegraph core (and the coupled phase
equation used later), impose standard BCs on a bounded domain $\Omega$ or a
torus $\mathbb T^d$:
\begin{itemize}
\item \textbf{Periodic} on $\mathbb T^d$;
\item \textbf{Homogeneous Neumann} $\partial_n\Psi=0$ on $\partial\Omega$;
\item \textbf{Dirichlet} $\Psi=\Psi_{\mathrm{bc}}$ on $\partial\Omega$ with compatible initial data.
\end{itemize}
Assume coefficients are bounded/measurable and the diffusion modulus is uniformly
positive to retain parabolic/hyperbolic coercivity.%
{\small \,:contentReference[oaicite:7]{index=7}}

\subsection{Minimal energy estimate (compatibility with residual dissipation)}
Define the field energy density (one convenient choice)
\[
\mathcal E(\Psi):=\tfrac12\,|\partial_t\Psi|^2+\tfrac{c^2}{2}\,|\nabla\Psi|^2,
\]
and $E(t)=\int_\Omega \mathcal E(\Psi)\,dx$. For the homogeneous core of
\eqref{eq:telegraph_core} without stochastic/source terms, standard multipliers
yield $\dot E\le 0$; the decay rate is controlled by the dispersion
\eqref{eq:dispersion}. This is consistent with the residual-space Lyapunov law
$\dot V\le0$ used in Chapters~4--5.%
{\small \,:contentReference[oaicite:8]{index=8}\,:contentReference[oaicite:9]{index=9}}

\subsection{Operator bridge from the local algebra}
The telegraph operator arises naturally from the symbolic substitution layer
that maps $(\omega,\kappa)\mapsto(i\partial_t,-\nabla^2)$ applied to the
dimensionless near-one algebra of Chapter~4; after normalization (this section),
the resulting PDE inherits scale invariance and the stability window.%
{\small \,:contentReference[oaicite:10]{index=10}}

\paragraph{Takeaway.}
Equations \eqref{eq:telegraph_core}--\eqref{eq:dispersion} provide a
dimensionless, symmetry-consistent, and provably \emph{stable} telegraph layer
for $\Psi$, compatible with causal noise, standard boundary conditions, and the
Lyapunov dissipation recorded at the residual level.
% ===================== End Chapter 6 =====================

% ===================== Chapter 7 =====================
\section{Green's Functions and Boundary Data}

\subsection{Dimensionless setting and operator}
We work in the global dimensionless frame (hats suppressed); cf.\ the normalization
$(\ell_0,t_0,\Psi_0)$ and the optional dimensionless Planck factor absorption.%
{\small \,:contentReference[oaicite:0]{index=0}}
The (dimensionless) telegraph-stable core reads
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi - c^2\nabla^2\Psi
\;=\; \mathsf{S}(\mathbf x,t),
\label{eq:telegraph_source}
\end{equation}
where $\mathsf{S}$ collects divergence of fluxes/feedbacks and (later) stochastic terms; Fourier--Laplace conventions are
$(\partial_t,\nabla^2)\mapsto(s,-|k|^2)$.%
{\small \,:contentReference[oaicite:1]{index=1}\,:contentReference[oaicite:2]{index=2}}

\subsection{Fourier--Laplace fundamental solution (whole space)}
Taking the spatial Fourier transform of \eqref{eq:telegraph_source} gives
\[
\tau\,\partial_t^2\widehat\Psi+\partial_t\widehat\Psi+c^2|k|^2\,\widehat\Psi=\widehat{\mathsf S}(k,t).
\]
The homogeneous dispersion relation is
\begin{equation}
\tau s^2+s+c^2|k|^2=0
\quad\Longrightarrow\quad
s_\pm(k)=\frac{-1\pm\sqrt{1-4\tau c^2|k|^2}}{2\tau},
\label{eq:disp}
\end{equation}
with $\Re s_\pm(k)\le -\tfrac{1}{2\tau}$ in the underdamped window and $s_\pm\in\mathbb R_{<0}$ in the over/critical cases; hence $\Re s<0$ in all regimes.%
{\small \,:contentReference[oaicite:3]{index=3}}
The impulse response (Green’s function in $k$-space) solves
\[
\tau\,\partial_t^2\widehat G+\partial_t\widehat G+c^2|k|^2\widehat G=\delta(t),\qquad \widehat G(k,0^+)=0,
\]
yielding
\begin{equation}
\widehat G(k,t)
=\frac{e^{s_+(k)t}-e^{s_-(k)t}}{\tau\big(s_+(k)-s_-(k)\big)}\;\mathbf 1_{t\ge0}.
\label{eq:Ghat}
\end{equation}
Hence for zero initial data,
\begin{equation}
\widehat\Psi(k,t)
=\int_0^t \widehat G(k,t-\sigma)\,\widehat{\mathsf S}(k,\sigma)\,d\sigma,
\qquad
\Psi(\cdot,t)=G(\cdot,t)*_{\mathbf x}\mathsf S(\cdot,t)\quad\text{(Duhamel)}.
\label{eq:duhamel}
\end{equation}
The real-space kernel $G(\mathbf x,t)$ is obtained by inverse Fourier transform and inherits exponential decay rates from \eqref{eq:disp}.%
{\small \,:contentReference[oaicite:4]{index=4}}

\subsection{Compatibility with stochastic forcing}
For causal, zero-mean noise with $\langle \xi(t)\rangle=0$ and
$\langle \xi(t)\xi(t')\rangle=2D\,\delta(t-t')$,
the mild solution remains well-defined via \eqref{eq:duhamel} under the same $\Re s<0$ window.%
{\small \,:contentReference[oaicite:5]{index=5}\,:contentReference[oaicite:6]{index=6}}
This stochastic convention and the stability window are part of the telegraph layer’s assumptions and validation checklist.%
{\small \,:contentReference[oaicite:7]{index=7}}

\subsection{Boundary data: spectral Green's functions on bounded domains}
Let $\Omega\subset\mathbb R^d$ be bounded with smooth boundary.
Adopt one of the standard boundary conditions to ensure well-posedness:
\begin{enumerate}
\item \textbf{Periodic} on $\mathbb T^d$;
\item \textbf{Homogeneous Neumann} $\partial_n\Psi=0$ on $\partial\Omega$;
\item \textbf{Dirichlet} $\Psi=\Psi_{\mathrm{bc}}$ on $\partial\Omega$ with compatible initial data.
\end{enumerate}
Assume bounded/measurable coefficients and a uniformly positive diffusion modulus to retain coercivity.%
{\small \,:contentReference[oaicite:8]{index=8}\,:contentReference[oaicite:9]{index=9}}

Let $\{-\nabla^2\phi_n=\lambda_n\phi_n\}$ be the Laplacian eigenbasis under the chosen BCs
($\lambda_n\ge0$, $\{\phi_n\}$ orthonormal in $L^2(\Omega)$). The modal telegraph ODE is
\[
\tau\,\ddot \Psi_n+\dot \Psi_n + c^2\lambda_n \Psi_n = S_n(t),
\qquad \Psi_n(t):=\langle \Psi(\cdot,t),\phi_n\rangle,
\]
with modal Green's function
\[
G_n(t)=\frac{e^{s_{+,n}t}-e^{s_{-,n}t}}{\tau(s_{+,n}-s_{-,n})}\,\mathbf 1_{t\ge0},
\qquad
s_{\pm,n}=\frac{-1\pm\sqrt{\,1-4\tau c^2\lambda_n\,}}{2\tau}.
\]
The full kernel is the eigen-expansion
\begin{equation}
G(\mathbf x,\mathbf y;t)=\sum_{n=1}^\infty G_n(t)\,\phi_n(\mathbf x)\phi_n(\mathbf y),
\qquad
\Psi(\mathbf x,t)=\int_0^t\!\!\int_\Omega G(\mathbf x,\mathbf y;t-\sigma)\,S(\mathbf y,\sigma)\,d\mathbf y\,d\sigma.
\label{eq:spectralG}
\end{equation}
Under the BC/regularity assumptions above, the series converges (termwise exponentially) because $\Re s_{\pm,n}<0$ for all $\lambda_n\ge0$.%
{\small \,:contentReference[oaicite:10]{index=10}}

\subsection{Initial data and mixed problems}
For initial data $\Psi(\cdot,0)=\Psi_0$ and $\partial_t\Psi(\cdot,0)=\Psi_1$,
the solution adds the homogeneous contribution
\[
\Psi_{\mathrm{IC}}(\mathbf x,t)=\sum_{n}\Big(
\alpha_n\,e^{s_{+,n}t}+\beta_n\,e^{s_{-,n}t}\Big)\phi_n(\mathbf x),
\]
with coefficients fixed by $(\Psi_0,\Psi_1)$ in the chosen basis.
The Duhamel term \eqref{eq:spectralG} superposes linearly.

\subsection{Energy and decay (consistency with residual layer)}
Define $\mathcal E(\Psi)=\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2$ and
$E(t)=\int_\Omega\mathcal E(\Psi)\,dx$.
For the homogeneous telegraph core, the standard multiplier method yields $\dot E\le 0$, with decay rates controlled by \eqref{eq:disp}; this aligns with the residual-space Lyapunov law $\dot V\le 0$ compiled in the validation checklist.%
{\small \,:contentReference[oaicite:11]{index=11}}

\paragraph{Takeaway.}
The Green’s function $G$ for the telegraph-stable operator exists in both whole-space and bounded-domain settings, decays at rates set by the dispersion roots, and supports causal stochastic forcing. The boundary-data spectral representation \eqref{eq:spectralG} provides a portable recipe for computation under periodic, Neumann, or Dirichlet constraints.
% ===================== End Chapter 7 =====================

% ===================== Chapter 8 =====================
\section{Variational Phase Coupling and Gradient Flows}

\subsection{Local phase variable and notation}
Let $\varphi=\varphi(\mathbf x,t)\in\mathbb S^1$ be a dimensionless phase order
parameter defined on a domain $\Omega\subset\mathbb R^d$ with time $t\ge0$.
All quantities are already dimensionless (rad/s convention; hats suppressed).
The coherence field $\Psi(\mathbf x,t)$ satisfies the telegraph-stable core of
Chapter~6. Throughout this chapter, all couplings are local (K-free).

\subsection{Energy functional (local-only, dimensionless)}
We postulate a phase energy density that couples $\varphi$ to $\Psi$:
\begin{equation}
\mathcal E_\varphi(\varphi,\Psi)
:= \frac{\kappa_\varphi}{2}\,|\nabla\varphi|^2
   + U(\varphi; m_{\mathrm{eff}})
   - J(\Psi)\,\cos\!\big(\varphi-\theta(\Psi)\big),
\label{eq:phase_energy_density}
\end{equation}
where $\kappa_\varphi>0$ is a dimensionless stiffness,
$U(\varphi;m_{\mathrm{eff}})$ is a smooth $2\pi$-periodic potential with
effective parameter $m_{\mathrm{eff}}$, and $J(\Psi)\ge0$ is a local coupling
gain depending only on the field amplitude at $(\mathbf x,t)$.
The phase reference $\theta(\Psi)$ is a local argument of $\Psi$ (defined
modulo $2\pi$). The total energy is
\begin{equation}
\mathcal F_\varphi[\varphi;\Psi] := \int_\Omega \mathcal E_\varphi(\varphi,\Psi)\,d\mathbf x .
\label{eq:phase_energy_total}
\end{equation}

\paragraph{Design notes.}
(i) Local-only: $J(\Psi)$ and $\theta(\Psi)$ are pointwise functions (no spatial
integrals or kernels). (ii) K-free: no nonlocal knowledge factors enter.
(iii) All parameters are dimensionless and can be anchored to $\Omega_0$ if
needed for reporting.

\subsection{Phase gradient flow and Lyapunov decrease}
A canonical residual-free gradient flow for the phase is
\begin{equation}
\partial_t\varphi \;=\; -\,\Gamma_\varphi\,\frac{\delta \mathcal F_\varphi}{\delta\varphi} \;+\; \xi(\mathbf x,t),
\qquad \Gamma_\varphi>0 ,
\label{eq:phase_gf}
\end{equation}
with causal, zero-mean noise $\xi$ (dimensionless; optional).
Using \eqref{eq:phase_energy_density}--\eqref{eq:phase_energy_total},
\[
\frac{\delta \mathcal F_\varphi}{\delta\varphi}
= -\,\kappa_\varphi\,\nabla^2\varphi \;+\; \partial_\varphi U(\varphi;m_{\mathrm{eff}})
 \;+\; J(\Psi)\,\sin\!\big(\varphi-\theta(\Psi)\big).
\]
Hence, in the deterministic case $\xi\equiv0$,
\begin{equation}
\frac{d}{dt}\,\mathcal F_\varphi[\varphi;\Psi]
= -\,\Gamma_\varphi \int_\Omega \Big|\frac{\delta \mathcal F_\varphi}{\delta\varphi}\Big|^2\,d\mathbf x
\;\le\; 0,
\label{eq:phase_Lyapunov}
\end{equation}
so $\mathcal F_\varphi$ is a Lyapunov functional for the phase subsystem.

\subsection{Coupling to the coherence field}
Let $\Psi$ satisfy the telegraph core (Chapter~6) with a local feedback
from $\varphi$:
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi - c^2\nabla^2\Psi
\;=\; \mathcal F_{\mathrm{tel}}(\Psi) \;+\; \alpha_{\mathrm{cpl}}\,
\sin\!\big(\varphi-\theta(\Psi)\big),
\qquad \alpha_{\mathrm{cpl}}\in\mathbb R,
\label{eq:telegraph_with_phase}
\end{equation}
where $\mathcal F_{\mathrm{tel}}$ collects admissible (local) source terms.
The sign and magnitude of $\alpha_{\mathrm{cpl}}$ determine whether the phase
acts as a restoring/locking feedback or a frustrative drive; the telegraph
stability window $\Re s<0$ (Chapter~6) remains enforceable provided the local
Lipschitz bounds on the right-hand side are respected.

\subsection{Phase-lock calibration indices}
Two convenient, dimensionless indicators are
\begin{align}
\rho_{\mathrm{lock}}(t)
&:= \Big|\frac{1}{|\Omega|}\int_\Omega e^{\,i(\varphi(\mathbf x,t)-\theta(\Psi(\mathbf x,t)))}\,d\mathbf x\Big|
\;\in[0,1], \label{eq:rho_lock}\\
\chi_{\mathrm{lock}}(t)
&:= 1 - \frac{\langle(\partial_t\varphi-\Omega_0)^2\rangle}{\langle(\partial_t\varphi)^2\rangle+\epsilon},
\qquad \epsilon\ll1,
\label{eq:chi_lock}
\end{align}
where $\langle\cdot\rangle$ denotes a spatial (or temporal) average and
$\Omega_0$ is the central cadence (rad/s) fixed in Chapter~2. High values of
$\rho_{\mathrm{lock}}$ and $\chi_{\mathrm{lock}}$ indicate phase and frequency
locking to the central reference.

\subsection{Equilibria and small-signal locking condition}
Stationary points satisfy
\begin{equation}
-\kappa_\varphi\nabla^2\varphi + \partial_\varphi U(\varphi;m_{\mathrm{eff}})
+ J(\Psi)\,\sin(\varphi-\theta(\Psi)) = 0 .
\label{eq:stationary_phase}
\end{equation}
Linearizing around a locked configuration $\varphi^\star=\theta(\Psi)$
with $U''(\varphi^\star;m_{\mathrm{eff}})=\mu_\varphi\ge0$ gives, to first order,
\[
\partial_t \tilde\varphi \;\approx\; -\Gamma_\varphi\,
\big(-\kappa_\varphi\nabla^2 + \mu_\varphi + J(\Psi)\big)\,\tilde\varphi ,
\]
so locking is (locally) attracting if $J(\Psi)$ dominates any destabilizing
curvature of $U$ and the smallest eigenvalue of $-\kappa_\varphi\nabla^2$
under the chosen BCs.

\subsection{Boundary conditions and well-posedness}
All BCs follow Chapter~6 (periodic, Neumann, Dirichlet) applied to $\varphi$
and $\Psi$. With $\kappa_\varphi>0$, $\Gamma_\varphi>0$, and smooth $U$,
the phase gradient flow is well-posed in $H^1(\Omega)$ and obeys
\eqref{eq:phase_Lyapunov}. The coupled system with \eqref{eq:telegraph_with_phase}
remains locally well-posed for sufficiently small Lipschitz constants of the
source terms.

\subsection{Simulation protocol (testable transition: Dynamics $\to$ Field)}
\begin{enumerate}
\item \textbf{No gradient (ablation):} set $\Gamma_\varphi=0$ in \eqref{eq:phase_gf};
solve \eqref{eq:telegraph_with_phase} and track $\Re s(k)$ and
$\rho_{\mathrm{lock}},\chi_{\mathrm{lock}}$; locking fails generically.
\item \textbf{With gradient:} restore $\Gamma_\varphi>0$; locking emerges and
the spectrum remains in the telegraph window $\Re s<0$.
\end{enumerate}
This pass/fail design obeys the governance rules (no circularity; falsifiable).
% ===================== End Chapter 8 =====================

% ===================== Chapter 9 =====================
\section{Source Modeling: Forcing, Feedback Currents, and Coupling Operators}

\subsection{Setting and notation (dimensionless, local-only)}
We work in the global dimensionless frame (rad/s canonical; hats suppressed).
The coherence field $\Psi(\mathbf x,t)$ satisfies the telegraph-stable core
from Chapter~6; a local phase $\varphi(\mathbf x,t)$ may enter via Chapter~8.
All operators and parameters below are pointwise in $(\mathbf x,t)$
(\emph{K-free, local-only}).

\subsection{Telegraph equation with general source grammar}
We write the field dynamics as
\begin{equation}
\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi
\;=\; \mathsf S(\Psi,\varphi;\,\varepsilon,\delta) \;:=\;
\underbrace{\nabla\!\cdot\mathsf q(\Psi)}_{\text{flux}}
\;+\; \underbrace{\mathsf f_{\text{drv}}(\Psi)}_{\text{drives}}
\;+\; \underbrace{\mathsf j_{\text{fb}}(\Psi,\varphi)}_{\text{feedback}}
\;+\; \underbrace{\xi(\mathbf x,t)}_{\text{noise}} .
\label{eq:telegraph_source_grammar}
\end{equation}
Here $(\varepsilon,\delta)$ denote the local residual/data features (Ch.~4--5).
Each term is \emph{local} (no spatial integrals) and satisfies admissibility
conditions listed below.

\subsection{Constitutive forms for fluxes and drives}
\paragraph{Flux (diffusive/hyperdiffusive).}
A minimal local closure is
\[
\mathsf q(\Psi) := -D(\Psi)\,\nabla\Psi - \eta(\Psi)\,\nabla\partial_t\Psi ,
\]
with $D(\cdot),\eta(\cdot)\ge0$ smooth, bounded functions. The first term is
Fick-like; the second is Kelvin--Voigt-like (rate damping). Both preserve
$\Re s<0$ provided the coefficients are bounded.

\paragraph{Drives (restoring, parametric, saturating).}
Examples include
\[
\mathsf f_{\text{drv}}(\Psi)
= -\,\alpha_1(\Psi)\,\Psi \;-\; \alpha_3(\Psi)\,|\Psi|^2\Psi
\;+\; \beta_{\text{pm}}(\Psi)\,\cos(\Omega_0 t+\phi_0),
\]
where $\alpha_1,\alpha_3\ge0$ (local passivity) and $\beta_{\text{pm}}$ is a
parametric drive with central cadence $\Omega_0$ fixed in Chapter~2.
Saturation ($\alpha_3>0$) prevents blow-up and is compatible with the energy law.

\subsection{Feedback currents and coupling operators}
\paragraph{Phase feedback (from Chapter~8).}
A canonical local coupling is
\[
\mathsf j_{\text{fb}}(\Psi,\varphi)
= \alpha_{\mathrm{cpl}}\,\sin\big(\varphi-\theta(\Psi)\big),
\]
where $\alpha_{\mathrm{cpl}}\in\mathbb R$ and $\theta(\Psi)$ is the local
argument of $\Psi$. The sign of $\alpha_{\mathrm{cpl}}$ selects locking vs.\
frustration.

\paragraph{Residual/data feedback (Seed-aware).}
Let $r:=A\varepsilon-\delta$ be the local residual. A seed-aware feedback
may be written as
\[
\mathsf j_{\text{fb}}^{(\mathrm{res})}(\Psi;\,r) \;=\;
-\gamma_r\,\mathcal M(\Psi)\, r,
\]
with $\gamma_r\ge0$ and a bounded, local multiplier $\mathcal M(\Psi)$.
This respects the governance rule (feedback used for falsification/testing;
no circular definition, since $r$ is computed from measurements).

\subsection{Noise models (causal, local)}
We adopt zero-mean, causal models:
\[
\langle \xi(\mathbf x,t)\rangle=0,\qquad
\langle \xi(\mathbf x,t)\,\xi(\mathbf x,t')\rangle
= 2 D_\xi(\mathbf x)\,\delta(t-t'),
\]
with $D_\xi(\mathbf x)\ge 0$ bounded. Colored variants can be realized via
auxiliary local ODEs driven by white noise.

\subsection{Admissibility rules (ensure stability and well-posedness)}
All source components in \eqref{eq:telegraph_source_grammar} must satisfy:
\begin{enumerate}
\item \textbf{Causality:} depend only on present/past values at $(\mathbf x,t)$.
\item \textbf{Locality:} no spatial integrals or nonlocal kernels.
\item \textbf{Lipschitz} on bounded sets: ensures local well-posedness.
\item \textbf{Passivity / dissipativity:} the quadratic energy estimate
$\,\dot E\le 0\,$ must hold for the homogeneous part; active drives must be
bounded/saturated so the telegraph window $\Re s<0$ is preserved.
\end{enumerate}

\subsection{Energy budget and passivity check}
Let the field energy be
\[
\mathcal E(\Psi)=\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2,\qquad
E(t)=\int_\Omega \mathcal E(\Psi)\,d\mathbf x .
\]
Multiplying \eqref{eq:telegraph_source_grammar} by $\partial_t\Psi$ and integrating
over $\Omega$ yields the budget
\begin{equation}
\dot E(t)
= -\int_\Omega \Big( \tau\,|\partial_t^2\Psi|^2 + |\partial_t\Psi|^2
+ c^2\,\nabla\Psi\!\cdot\!\nabla\partial_t\Psi \Big)\,d\mathbf x
+ \int_\Omega \partial_t\Psi\;\mathsf S\,d\mathbf x .
\label{eq:energy_budget}
\end{equation}
Under the flux/drive choices above, the source contribution is
\[
\int_\Omega \partial_t\Psi\,\nabla\!\cdot\mathsf q\,d\mathbf x
= -\int_\Omega \nabla\partial_t\Psi\cdot \mathsf q\,d\mathbf x\;\le\; 0,
\]
and
\[
\int_\Omega \partial_t\Psi\,\big(-\alpha_1\Psi-\alpha_3|\Psi|^2\Psi\big)\,d\mathbf x
\;\le\; 0,
\]
while parametric/feedback parts are balanced by saturation or
small-gain conditions (see next subsection).

\subsection{Small-gain / saturation conditions (telegraph window)}
To preserve $\Re s<0$ across wavenumbers, impose either
\begin{itemize}
\item \textbf{Saturation:} $\alpha_3(\Psi)\ge \alpha_{3,\min}>0$ beyond a threshold.
\item \textbf{Small-gain:} $\|\mathsf f_{\text{drv}}+\mathsf j_{\text{fb}}\|_{L^\infty}
\le \gamma_\star \|\Psi\|$ with $\gamma_\star$ below the minimal damping rate
set by $(\tau,c)$.
\end{itemize}
Both are local, dimensionless, and compatible with Ch.~6 energy estimates.

\subsection{Calibration to $\Omega_0$ and $W_0$ (Option A)}
All amplitudes are reported in the rad/s canonical frame. When a Hz report is
needed, use $f_0=\Omega_0/(2\pi)$ for figures/tables (see Compatibility Note in
Ch.~2). The measurement weight decomposes as
\[
W=W_0\,\widetilde W ,\qquad \Lambda=\lambda_0\,\widetilde\Lambda,
\]
with $W_0,\lambda_0>0$ scalars and tilde-objects dimensionless. Source gains
($\beta_{\text{pm}}$, $\alpha_{\mathrm{cpl}}$, etc.) are scaled so that the
energy budget \eqref{eq:energy_budget} remains dimensionless and portable.

\subsection{Examples (domain-agnostic templates)}
\begin{enumerate}
\item \textbf{Linear restorative medium:} $D(\Psi)=D_0$, $\eta(\Psi)=0$,
$\alpha_1(\Psi)=\alpha_1>0$, $\alpha_3=0$.
\item \textbf{Saturating oscillator:} $D(\Psi)=D_0$, $\alpha_1>0$, $\alpha_3>0$
(prevents runaway under parametric drive).
\item \textbf{Phase-locked feedback:} $\alpha_{\mathrm{cpl}}>0$; monitor
$\rho_{\text{lock}}$ and $\chi_{\text{lock}}$ (Ch.~8) for locking transitions.
\item \textbf{Residual-aware correction:} add $-\gamma_r\mathcal M(\Psi)\,r$
with small $\gamma_r$; use it as a falsification channel (not a definition).
\end{enumerate}

\subsection{Simulation checklist (testable transitions)}
\begin{enumerate}
\item \emph{Ablation:} remove flux ($D=\eta=0$) or saturation ($\alpha_3=0$)
and demonstrate violation of the small-gain bound (loss of $\Re s<0$).
\item \emph{Restore admissibility:} reinstate flux/saturation; verify
$\Re s<0$, monotone energy decay, and (if feedback present) phase-lock emergence.
\end{enumerate}

\paragraph{Takeaway.}
Equation \eqref{eq:telegraph_source_grammar} provides a compact, local-only
grammar of sources and couplings for $\Psi$. The admissibility rules, energy
budget, and calibration to $(\Omega_0,W_0)$ ensure stability and portability,
aligning the source layer with the residual and phase-coupling layers.
% ===================== End Chapter 9 =====================

% ===================== Chapter 10 =====================
\section{Energy--Coherence Law: Observables, Budgets, and Measurement Links}

\subsection{Design goals}
The Energy--Coherence Law serves three purposes:
\begin{enumerate}
\item \textbf{Compatibility:} agree with the telegraph layer (Ch.~6) and with
the residual Lyapunov law $\dot V\le0$ (Ch.~5).
\item \textbf{Observability:} provide a clean entry point for measured
deviations $\delta$ (Ch.~4) without circular definitions.
\item \textbf{Portability:} be dimensionless, K-free, and calibrated to the
Option~A anchors ($\Omega_0$ in rad/s; Ch.~2).
\end{enumerate}

\subsection{Coherence energy density and field energy}
Let $\Psi(\mathbf x,t)$ solve the telegraph-stable core (Ch.~6) with admissible
sources (Ch.~9). Define the \emph{coherence energy density}
\begin{equation}
\mathcal E_{\text{coh}}(\Psi):=\frac{1}{2}\,|\partial_t\Psi|^2
+\frac{c^2}{2}\,|\nabla\Psi|^2
+\Phi(\Psi),
\label{eq:E_coh_density}
\end{equation}
where $\Phi(\Psi)\ge0$ is a local potential chosen to reflect restorative
or saturating effects consistent with the source grammar (Ch.~9).
The total coherence energy is
\begin{equation}
E_{\text{coh}}(t):=\int_\Omega \mathcal E_{\text{coh}}(\Psi(\mathbf x,t))\,d\mathbf x.
\label{eq:E_coh_total}
\end{equation}

\paragraph{Choices for $\Phi(\Psi)$.}
Typical options:
\begin{itemize}
\item \emph{Linear restorative:} $\Phi(\Psi)=\tfrac{\alpha_1}{2}|\Psi|^2$, $\alpha_1\ge0$.
\item \emph{Saturating:} $\Phi(\Psi)=\tfrac{\alpha_1}{2}|\Psi|^2+\tfrac{\alpha_3}{4}|\Psi|^4$,
with $\alpha_3>0$ to prevent runaway under parametric drives.
\end{itemize}
Both choices are dimensionless under the Option~A normalization.

\subsection{Residual-space energy and the bridge}
Recall the residual $r:=A\varepsilon-\delta$ (Ch.~4) and the Lyapunov energy
\begin{equation}
V(\varepsilon;\delta):=\frac{1}{2}\,r^\top W r, \qquad \dot V\le 0
\quad \text{under}\quad \dot\varepsilon=-\Lambda A^\top W r \ \text{(Ch.~5)}.
\label{eq:residual_energy}
\end{equation}
To avoid circularity, we adopt the \emph{bridge principle}:
\begin{equation}
E_{\text{coh}}(t) \ \text{is computed from}\ \Psi \ \text{only},\qquad
V(t)\ \text{is computed from}\ (\varepsilon,\delta)\ \text{only}.
\label{eq:bridge_principle}
\end{equation}
Consistency is assessed by testing whether the \emph{observable predictions} of
the field model produce measured residuals $\delta$ that are compatible with the
seed closure (Ch.~4) when fed back for falsification.

\subsection{Coherence budget (field level)}
Multiply the telegraph equation (Ch.~6) by $\partial_t\Psi$ and integrate over
$\Omega$ to obtain the budget
\begin{equation}
\dot E_{\text{coh}}(t)
= - \mathcal D_{\text{tel}}(t) + \mathcal I_{\text{src}}(t),
\label{eq:coh_budget}
\end{equation}
with
\begin{equation}
\mathcal D_{\text{tel}}(t):=\int_\Omega \big(\tau\,|\partial_t^2\Psi|^2
+|\partial_t\Psi|^2\big)\,d\mathbf x \ \ge 0 ,
\qquad
\mathcal I_{\text{src}}(t):=\int_\Omega \partial_t\Psi\;\mathsf S(\Psi,\varphi;\varepsilon,\delta)\,d\mathbf x .
\label{eq:D_and_I}
\end{equation}
Under the admissibility rules (Ch.~9),
flux terms contribute nonpositively (after integration by parts) and restorative
drives are passive; parametric or feedback terms must satisfy small-gain or
saturation bounds, keeping $\Re s<0$ and preventing growth of $E_{\text{coh}}$.

\subsection{Energy--coherence law (statement)}
\begin{definition}[Energy--Coherence Law]
A source/coupling configuration satisfies the Energy--Coherence Law on
$[0,T]$ if
\begin{equation}
\dot E_{\text{coh}}(t)\ \le\ 0
\quad \text{for almost every } t\in[0,T],
\label{eq:coh_law}
\end{equation}
with equality only at equilibria of the telegraph flow modulo admissible
parametric modulations. The law is \emph{dimensionless}, \emph{local}, and
\emph{K-free}.
\end{definition}

\subsection{Measurement links (observables and $\delta$)}
Let $\mathcal O[\Psi]$ be a local observable (e.g., amplitude, phase, or bandpower
of $\Psi$). A measurement model is a local map
\begin{equation}
\delta = \mathcal M\big(\mathcal O[\Psi]\big) + \eta, \qquad
\eta\ \text{noise},\ \ \mathbb E[\eta]=0,
\label{eq:measurement_model}
\end{equation}
with \emph{no} dependence on $\varepsilon$ in the definition (non-circular).
The falsification loop is:
\begin{enumerate}
\item Use \eqref{eq:measurement_model} to obtain $\delta$ from field outputs.
\item Solve the seed closure $A\varepsilon=\delta$ (Ch.~4) and evaluate $V$ (Ch.~5).
\item If $V$ violates a prescribed tolerance, the source/feedback model
is rejected or recalibrated (Ch.~9).
\end{enumerate}

\subsection{Calibration to $\Omega_0$ and $W_0$ (Option A)}
We report frequencies in rad/s with the anchor $\Omega_0$ (Ch.~2).
When needed for figures/tables, use $f_0=\Omega_0/(2\pi)$ (Hz) without altering
the algebra. The measurement weight and metric are decomposed as
\begin{equation}
W=W_0\,\widetilde W,\qquad \Lambda=\lambda_0\,\widetilde\Lambda,
\label{eq:W0_lambda0}
\end{equation}
with dimensionless $\widetilde W,\widetilde\Lambda$. The constants
$(W_0,\lambda_0)$ set the absolute scales for data fidelity and dissipation rates.

\subsection{Locking indicators and energy consistency}
Define locking indices (Ch.~8)
\[
\rho_{\mathrm{lock}}(t),\quad \chi_{\mathrm{lock}}(t)\in[0,1],
\]
and record the \emph{consistency condition}:
if the configuration is in the locking regime (high $\rho_{\mathrm{lock}}$ and
$\chi_{\mathrm{lock}}$), then the measured residual energy should decrease on
average,
\begin{equation}
\frac{d}{dt}\,\mathbb E[V(\varepsilon;\delta)] \ \le\ 0,
\label{eq:residual_consistency}
\end{equation}
modulo noise and admissible parametric modulations, reflecting the alignment
between field coherence and residual consistency.

\subsection{Falsifiable tests (no circularity)}
\begin{enumerate}
\item \textbf{Ablation (fail expected):} remove saturation or violate the
small-gain condition in the source grammar; observe $\dot E_{\text{coh}}>0$
episodes or loss of $\Re s<0$ in the dispersion.
\item \textbf{Restoration (pass expected):} reinstate admissibility; verify
$\dot E_{\text{coh}}\le0$ and decreasing $V$ per \eqref{eq:residual_consistency}.
\item \textbf{Mapping check (no circularity):} perturb the measurement model
$\mathcal M$ and verify that the inferred $V$ reacts while $E_{\text{coh}}$
remains computed purely from $\Psi$.
\end{enumerate}

\subsection{Domain-agnostic reporting protocol}
For any domain (quantum, neural, or cosmic):
\begin{enumerate}
\item Fix $(\Omega_0,W_0)$ and report \emph{dimensionless} parameters.
\item Publish the source grammar and admissibility constants.
\item Provide $(E_{\text{coh}},V,\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$
time series and pass/fail outcomes for the falsifiable tests.
\end{enumerate}

\paragraph{Takeaway.}
The Energy--Coherence Law links field stability (telegraph budget),
measurement consistency (residual energy), and calibration
$(\Omega_0,W_0)$ without circularity. It is dimensionless, local-only,
and testable across domains.
% ===================== End Chapter 10 =====================

% ===================== Chapter 11 =====================
\section{Well-Posedness \& Regularity at Field Level (Deterministic \& Stochastic)}

\subsection{Model, domain, and boundary conditions}
Let $\Omega\subset\mathbb R^d$ be a bounded $C^{1,1}$ domain or a torus $\mathbb T^d$.
We consider the telegraph-stable core with admissible sources (Ch.~9)
\begin{equation}
\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi
= \mathsf S(\Psi,\varphi;\varepsilon,\delta) \quad \text{in } \Omega\times(0,T],
\label{eq:telegraph11}
\end{equation}
with initial data $(\Psi,\partial_t\Psi)|_{t=0}=(\Psi_0,\Psi_1)$ and one of:
(i) periodic BC on $\mathbb T^d$,
(ii) homogeneous Neumann $\partial_n\Psi=0$ on $\partial\Omega$,
(iii) Dirichlet $\Psi=\Psi_{\mathrm{bc}}$ on $\partial\Omega$.
All coefficients are dimensionless (Option~A; hats suppressed).

\subsection{Function spaces and weak formulation}
Write $H=L^2(\Omega)$, $V=H^1(\Omega)$ with the usual norms; when Dirichlet BC is used,
replace $V$ by $H_0^1(\Omega)$. Let $\langle\cdot,\cdot\rangle$ be the $H$ inner product.
A weak solution on $[0,T]$ satisfies, for all $\phi\in V$ and a.e.\ $t$,
\[
\langle \tau\,\partial_t^2\Psi + \partial_t\Psi,\phi\rangle
+ c^2\int_\Omega \nabla\Psi\cdot\nabla\phi\,dx
= \int_\Omega \mathsf S(\Psi,\varphi;\varepsilon,\delta)\,\phi\,dx,
\]
with $\Psi\in L^\infty(0,T;V)$, $\partial_t\Psi\in L^\infty(0,T;H)$, and
$\partial_t^2\Psi\in L^2(0,T;V')$.

\subsection{Assumptions on the source grammar (local-only)}
All source components are \emph{local} and satisfy:
\begin{enumerate}
\item \textbf{Causality}: depend on $(\mathbf x,t)$ and its past only.
\item \textbf{Lipschitz} on bounded subsets of $(\Psi,\nabla\Psi,\partial_t\Psi)$:
there exists $L>0$ s.t.
$\|\mathsf S(u)-\mathsf S(v)\|_{H}\le L\,\|u-v\|_{V\times H}$.
\item \textbf{Passivity/small-gain}: either saturating nonlinearity or small-gain
bounds ensuring the Energy--Coherence budget (Ch.~10) holds, keeping $\Re s<0$.
\end{enumerate}

\subsection{Deterministic well-posedness}
\begin{theorem}[Existence and uniqueness]
Assume $\tau>0$, $c>0$, admissible BC, initial data $(\Psi_0,\Psi_1)\in V\times H$,
and a source $\mathsf S$ satisfying the Lipschitz and passivity conditions above.
Then there exists a unique weak solution
\[
\Psi \in L^\infty(0,T;V),\qquad
\partial_t\Psi\in L^\infty(0,T;H),\qquad
\partial_t^2\Psi\in L^2(0,T;V'),
\]
depending continuously on the data. If $\mathsf S$ is $C^1$ and
$(\Psi_0,\Psi_1)\in H^2(\Omega)\cap V \times V$, then $\Psi$ is a strong solution
with $\Psi\in C([0,T];H^2\cap V)$ and $\partial_t\Psi\in C([0,T];V)$.
\end{theorem}

\begin{proof}[Energy method (sketch)]
Multiply \eqref{eq:telegraph11} by $\partial_t\Psi$ and integrate over $\Omega$;
integration by parts (compatible with each BC) yields the coherence budget
\[
\dot E_{\text{coh}}(t)
= -\int_\Omega \big(\tau\,|\partial_t^2\Psi|^2 + |\partial_t\Psi|^2\big)\,dx
+ \int_\Omega \partial_t\Psi\,\mathsf S\,dx,
\]
where $E_{\text{coh}}$ is as in Ch.~10. Passivity/small-gain makes the source
term controlled; Grönwall then gives a priori bounds and uniqueness by difference
estimates. Galerkin or semigroup approximations provide existence.
\end{proof}

\subsection{Mild formulation and semigroup viewpoint}
Let $\mathcal A=\begin{psmallmatrix}0&I\\ c^2\Delta & -\tau^{-1}I\end{psmallmatrix}$
on $\mathcal X=V\times H$ with domain $D(\mathcal A)=\{(u,v):u\in H^2\cap V,\,v\in V\}$.
Then $\partial_t Z=\mathcal A Z + \mathcal N(Z)$ with
$Z=(\Psi,\partial_t\Psi)$ and $\mathcal N(Z)=(0,\tau^{-1}\mathsf S)$
admits a (nonlinear) variation-of-constants formula
\[
Z(t)=e^{t\mathcal A}Z(0)+\int_0^t e^{(t-s)\mathcal A}\,\mathcal N(Z(s))\,ds,
\]
equivalent to the spectral/Green representation (Ch.~7) under the same BCs.

\subsection{Regularity and bootstraps}
If $\mathsf S$ is $C^k$ in $\Psi$ with bounded derivatives on bounded sets,
and data are smoother, standard elliptic/energy bootstraps give
\[
(\Psi_0,\Psi_1)\in H^{m+1}\times H^m \ \Rightarrow\
\Psi\in C([0,T];H^{m+1})\cap C^1([0,T];H^{m}),
\]
subject to compatibility conditions at $t=0$ for Dirichlet BC.
For periodic/Neumann BC, similar gains hold modulo constants.

\subsection{Stochastic well-posedness (additive, causal noise)}
Consider
\[
\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi
= \mathsf S(\Psi,\varphi;\varepsilon,\delta) + \xi(\mathbf x,t),
\]
with $\xi$ zero-mean, white-in-time, $L^2$-valued noise:
$\mathbb E[\xi(\cdot,t)]=0$,
$\mathbb E[\langle \xi(t),\phi\rangle\,\langle \xi(t'),\psi\rangle]
= 2\langle Q\phi,\psi\rangle\,\delta(t-t')$ for a bounded, nonnegative $Q$.
A mild (stochastic) solution on $[0,T]$ satisfies
\[
Z(t)=e^{t\mathcal A}Z(0)+\int_0^t e^{(t-s)\mathcal A}\,\mathcal N(Z(s))\,ds
+\int_0^t e^{(t-s)\mathcal A}\,(0,\tau^{-1}\,dW_Q(s)),
\]
where $W_Q$ is a $Q$-Wiener process. Under the same Lipschitz/small-gain bounds,
one obtains existence/uniqueness in $L^2(\Omega; \mathcal X)$ and
\[
\mathbb E\,E_{\text{coh}}(t)\ +\ \mathbb E\!\int_0^t \!\!\!\int_\Omega
\big(\tau\,|\partial_t^2\Psi|^2+|\partial_t\Psi|^2\big)\,dx\,ds
\ \le\ C\big( \mathbb E\,E_{\text{coh}}(0) + t\,\mathrm{Tr}\,Q \big).
\]

\subsection{Continuous dependence and stability}
Let $\Psi^{(1)}$, $\Psi^{(2)}$ solve \eqref{eq:telegraph11} with data/sources
differing by at most $\delta$ in $V\times H$ and $L^2(0,T;H)$. Then
\[
\|\Psi^{(1)}-\Psi^{(2)}\|_{L^\infty(0,T;V)}^2
+\|\partial_t\Psi^{(1)}-\partial_t\Psi^{(2)}\|_{L^\infty(0,T;H)}^2
\ \le\ C_T\,\delta^2,
\]
where $C_T$ depends on $(\tau,c)$ and the Lipschitz constants of $\mathsf S$.

\subsection{Compatibility with residual and phase layers}
The above analysis is \emph{local-only} and K-free: the source grammar does not
depend on nonlocal parameters. The Energy--Coherence budget used here matches
Ch.~10; residual-space dissipation (Ch.~5) and phase coupling (Ch.~8) are
admissible via the small-gain/saturation constraints (Ch.~9) and do not violate
the $\Re s<0$ window.

\paragraph{Takeaway.}
Under standard BCs and local admissibility assumptions, the telegraph field
admits unique (weak/strong/mild) solutions with energy control, stability, and
regularity compatible with the rest of the monograph. The stochastic extension
inherits the same structure at the level of expectations.
% ===================== End Chapter 11 =====================

% ===================== Chapter 12 =====================
\section{Numerical Schemes for Telegraph PDEs (Stability, CFL, Accuracy)}

\subsection{Model, grid, and notation}
We discretize $\Omega\subset\mathbb R^d$ on a Cartesian grid with spacing
$\Delta x$ (assume $\Delta x_1=\cdots=\Delta x_d=\Delta x$ for simplicity) and
time step $\Delta t>0$. Let $t^n=n\Delta t$, grid indices $i\in\mathbb Z^d$,
and write $\Psi_i^n\approx \Psi(\mathbf x_i,t^n)$. Spatial derivatives use
second-order centered differences; hats are suppressed (Option~A frame).

\paragraph{Discrete Laplacian.}
The standard $d$-D centered Laplacian reads
\[
(\Delta_h \Psi)_i
= \frac{1}{\Delta x^2}\sum_{\ell=1}^{d}
\big(\Psi_{i+e_\ell}-2\Psi_i+\Psi_{i-e_\ell}\big).
\]
On a periodic grid its Fourier symbol is
\[
\lambda_h(\boldsymbol\theta)
= \frac{2}{\Delta x^2}\sum_{\ell=1}^{d}\big(\cos\theta_\ell-1\big)
= -\frac{4}{\Delta x^2}\sum_{\ell=1}^{d}\sin^2(\theta_\ell/2)\ \le\ 0,
\]
with $|\lambda_h|\le \lambda_{\max}:=\frac{4d}{\Delta x^2}$.

\subsection{Semi-discrete form and velocity variable}
Introduce the velocity $V=\partial_t\Psi$ and write
\[
\partial_t\Psi = V,\qquad
\tau\,\partial_t V + V - c^2\Delta \Psi = \mathsf S(\Psi,\varphi;\varepsilon,\delta).
\]
After space discretization, $(\Psi,V)$ solves an ODE system
\[
\dot \Psi = V,\qquad
\tau\,\dot V + V - c^2 \Delta_h \Psi = \mathsf S_h(\Psi,\varphi;\varepsilon,\delta),
\]
to be advanced in time by one of the schemes below.

\subsection{Baseline explicit (second order): leapfrog--Heun}
A simple explicit, second-order pair combines centered second differences in time
for $\Psi$ with a Heun predictor--corrector for the damping/source:
\begin{align*}
&\textbf{(LF)}\quad
\tau\,\frac{\Psi_i^{n+1}-2\Psi_i^{n}+\Psi_i^{n-1}}{\Delta t^2}
+ \frac{\Psi_i^{n+1}-\Psi_i^{n-1}}{2\Delta t}
- c^2(\Delta_h\Psi^n)_i
= \mathsf S_h(\Psi^n,\varphi^n;\varepsilon^n,\delta^n),\\
&\textbf{(Heun correction)}\quad
\mathsf S_h^{*,n+1} := \mathsf S_h(\Psi^n,\varphi^n;\varepsilon^n,\delta^n),\quad
\tilde\Psi^{\,n+1} := \text{solution of LF with }\mathsf S_h^{*,n+1},\\
&\hspace{2.7cm}
\mathsf S_h^{n+1} := \mathsf S_h(\tilde\Psi^{\,n+1},\varphi^{n+1};\varepsilon^{n+1},\delta^{n+1}),\\
&\hspace{2.7cm}
\text{re-solve LF with }\tfrac{1}{2}(\mathsf S_h^{*,n+1}+\mathsf S_h^{n+1}).
\end{align*}
\textbf{CFL guidance (explicit):}
A conservative rule, derived by von Neumann on the homogeneous part, is
\begin{equation}
\frac{c\,\Delta t}{\Delta x}\ \le\ \frac{1}{\sqrt{d}},
\qquad
\Delta t\ \le\ 2\,\tau,
\label{eq:CFL_explicit}
\end{equation}
ensuring that the leapfrog part resolves wave propagation and that the
damping term $+\partial_t\Psi$ is not overstepped. In practice, pick
$\Delta t=\min\big\{\sigma\,\tfrac{\Delta x}{c\sqrt d},\ \sigma_\tau\,2\tau\big\}$
with safety factors $\sigma,\sigma_\tau\in(0,1)$.

\subsection{Crank--Nicolson (CN) for linear core (unconditionally stable for the linear part)}
Treat the linear telegraph operator implicitly and sources explicitly:
\[
\tau\,\frac{\Psi^{n+1}-2\Psi^{n}+\Psi^{n-1}}{\Delta t^2}
+\frac{\Psi^{n+1}-\Psi^{n-1}}{2\Delta t}
- c^2\,\Delta_h\!\Big(\frac{\Psi^{n+1}+\Psi^{n-1}}{2}\Big)
= \mathsf S_h(\Psi^{n},\varphi^{n};\varepsilon^{n},\delta^{n}).
\]
For periodic/Neumann BCs this leads to a symmetric positive definite (SPD)
linear system per step. The linear homogeneous update is
\emph{unconditionally A-stable} in the sense of the telegraph core; overall
stability still requires admissible sources (Chapter~9).

\subsection{IMEX (implicit diffusion, explicit damping/source)}
A robust alternative is an IMEX (CN/AB2) split:
\begin{align*}
&\tau\,\frac{\Psi^{n+1}-2\Psi^{n}+\Psi^{n-1}}{\Delta t^2}
- c^2\,\Delta_h\!\Big(\frac{\Psi^{n+1}+\Psi^{n-1}}{2}\Big)\\
&\qquad = -\,\frac{\Psi^{n}-\Psi^{n-1}}{\Delta t}
+ \Big(\tfrac{3}{2}\mathsf S_h^{n}-\tfrac{1}{2}\mathsf S_h^{n-1}\Big).
\end{align*}
This keeps the stiff spatial operator implicit (unconditional for the linear
part) while maintaining second-order accuracy and cheap explicit handling for
damping and sources. Dirichlet BCs give an SPD system after enforcing boundary
rows in $\Delta_h$.

\subsection{Newmark family (single-parameter damping control)}
Define $V^n\approx\partial_t\Psi^n$ and choose Newmark parameters
$\beta\in[\tfrac16,\tfrac14]$, $\gamma\in[\tfrac12,1]$:
\begin{align*}
\Psi^{n+1} &= \Psi^{n} + \Delta t\,V^{n}
+ \Delta t^2\Big[(\tfrac12-\beta)\,A^{n}+\beta\,A^{n+1}\Big],\\
V^{n+1} &= V^{n} + \Delta t\Big[(1-\gamma)\,A^{n}+\gamma\,A^{n+1}\Big],
\end{align*}
with acceleration $A^n$ defined by the semi-discrete telegraph ODE
$\tau A^n + V^n - c^2\Delta_h\Psi^n = \mathsf S_h^n$.
Choice $\beta=\tfrac14,\gamma=\tfrac12$ (average acceleration) is
unconditionally stable for the linear part; taking $\gamma>\tfrac12$ adds
numerical damping at high frequencies.

\subsection{Discrete energy monitor (falsifiable stability check)}
Adopt the discrete energy
\[
E_{\text{coh}}^{n}
= \sum_i \Big(\tfrac12|V_i^{n}|^2
+ \tfrac{c^2}{2}\,|\nabla_h\Psi_i^{n}|^2 + \Phi(\Psi_i^n)\Big)\,\Delta x^d,
\]
with $\nabla_h$ the centered gradient and $\Phi$ matching Chapter~10.
For the linear core, CN/IMEX/Newmark($\beta=\tfrac14,\gamma\ge\tfrac12$) yield
monotone or nonincreasing $E_{\text{coh}}^{n}$ up to source/work terms,
providing a \emph{testable} discrete analog of $\dot E_{\text{coh}}\le0$.

\subsection{Boundary conditions}
\begin{itemize}
\item \textbf{Periodic}: wrap indices in $\Delta_h$; FFTs diagonalize the
linear step (useful for CN/IMEX).
\item \textbf{Neumann}: mirror values for ghost points or use one-sided
second-order stencils satisfying $\partial_n\Psi=0$.
\item \textbf{Dirichlet}: enforce $\Psi=\Psi_{\mathrm{bc}}$ on boundary nodes;
modify rows in the linear system accordingly.
\end{itemize}

\subsection{Accuracy, dispersion, and diagnostics}
All schemes above are second order in space (with centered differences).
The time accuracy:
\begin{itemize}
\item LF--Heun: second order, explicit (subject to \eqref{eq:CFL_explicit}).
\item CN: second order, unconditionally A-stable for linear core.
\item IMEX (CN/AB2): second order if $\mathsf S_h$ is smooth.
\item Newmark($\beta=\tfrac14$): second order; $\gamma$ controls numerical damping.
\end{itemize}
For dispersion, compare the numerical amplification factor $g(\theta)$ against the
exact telegraph roots $s_\pm(k)$ (Chapter~6) by setting $\theta$ to grid modes
$\boldsymbol\theta=\mathbf k\Delta x$; acceptable error keeps $\Re(\log g)/\Delta t<0$.

\subsection{Reproducible test suite (pass/fail)}
\begin{enumerate}
\item \textbf{Homogeneous decay (no source):} start from a band-limited initial
packet; verify $E_{\text{coh}}^n$ decreases and numerical dispersion follows
$e^{s_\pm(k)t}$ within tolerance. \emph{Fail} if energy grows.
\item \textbf{Source admissibility (Chapter~9):} add a parametric drive and
saturation; \emph{pass} if $E_{\text{coh}}^n$ remains bounded and lock indices
(Chapter~8) behave as predicted; \emph{fail} if violating small-gain yields
growth.
\item \textbf{Convergence}: refine $(\Delta x,\Delta t)$ with fixed CFL;
observe second-order slopes in $L^2$ error vs. step size.
\end{enumerate}

\subsection{Practical parameter selection (Option A)}
Set
\[
\Delta t=\min\Big\{\sigma\,\frac{\Delta x}{c\sqrt d},\ \sigma_\tau\,2\tau\Big\}
\quad\text{(explicit)};\qquad
\Delta t=\sigma_{\text{imp}}\,\frac{\Delta x}{c\sqrt d}\ \text{or larger}
\quad\text{(CN/IMEX/Newmark)},
\]
with $\sigma,\sigma_\tau,\sigma_{\text{imp}}\in(0,1)$.
Report all frequencies in rad/s; if Hz is needed for plots, use
$f_0=\Omega_0/(2\pi)$ (Chapter~2, Compatibility Note).

\paragraph{Takeaway.}
Centered differences in space with CN/IMEX/Newmark time stepping yield robust,
dimensionless, K-free integrators for the telegraph-stable field. Explicit
LF--Heun can be used under a conservative CFL. Discrete energy monitoring
provides a falsifiable stability check aligned with the Energy--Coherence Law.
% ===================== End Chapter 12 =====================

% ===================== Chapter 13 =====================
\section{Robust Solutions of $A\varepsilon=\delta$: Conditioning, Bounds, and Diagnostics}

\subsection{Models: square and overdetermined closures}
\paragraph{Square (canonical H0--H4).}
Let $A\in\mathbb R^{4\times 4}$ be the Jacobian from Chapter~4 with $\det A\neq0$.
The seed closure solves
\begin{equation}
A\varepsilon=\delta,\qquad \varepsilon=A^{-1}\delta.
\label{eq:seed_square}
\end{equation}

\paragraph{Overdetermined (data-rich).}
Let $G\in\mathbb R^{m\times 4}$, $m\ge 4$, assemble additional near-unity
relations. With a symmetric weight $W\succeq0$ (often $W=\Sigma^{-1}$),
the weighted least-squares (WLS) estimator is
\begin{equation}
\hat\varepsilon=\arg\min_\varepsilon \|W^{1/2}(G\varepsilon-\delta)\|_2^2
=(G^\top W G)^{-1}G^\top W\delta,
\label{eq:WLS}
\end{equation}
whenever $\mathrm{rank}(G)=4$ and $G^\top W G\succ0$.

\subsection{Conditioning and perturbation bounds}
Write $\kappa_2(M)=\|M\|_2\|M^{-1}\|_2$ for the spectral condition number.

\paragraph{Square case.}
For small perturbations $(\Delta A,\Delta\delta)$, the first-order bound is
\begin{equation}
\frac{\|\Delta\varepsilon\|_2}{\|\varepsilon\|_2}
\ \lesssim\ \kappa_2(A)\Big(
\frac{\|\Delta A\|_2}{\|A\|_2} + \frac{\|\Delta\delta\|_2}{\|\delta\|_2}
\Big),
\qquad \Delta\varepsilon \approx A^{-1}(\Delta\delta-\Delta A\,\varepsilon).
\label{eq:pert_square}
\end{equation}
In particular, with exact $A$,
\begin{equation}
\|\Delta\varepsilon\|_2 \le \|A^{-1}\|_2\,\|\Delta\delta\|_2.
\label{eq:seed_sensitivity}
\end{equation}

\paragraph{WLS case.}
Let $C:=W^{1/2}$ and $\tilde G:=C G$, $\tilde\delta:=C\delta$. Then
\begin{equation}
\hat\varepsilon=\tilde G^{\dagger}\,\tilde\delta,\qquad
\|\Delta\hat\varepsilon\|_2\ \le\ \|\tilde G^{\dagger}\|_2\,\|\Delta\tilde\delta\|_2
\ =\ \sigma_{\min}(\tilde G)^{-1}\,\|\Delta\tilde\delta\|_2,
\label{eq:WLS_sensitivity}
\end{equation}
where $\sigma_{\min}(\tilde G)$ is the smallest singular value.

\subsection{Scaling and preconditioning (dimensionless, local)}
Although the algebra is in hats (dimensionless), numerical conditioning still
benefits from equilibration.

\paragraph{Diagonal equilibration.}
Choose diagonal $D_r\succ0$, $D_c\succ0$ to balance row/column magnitudes:
$\bar G:=D_r\,G\,D_c$. Solve $\bar G\,\bar\varepsilon=\bar\delta$ with
$\bar\varepsilon=D_c^{-1}\varepsilon$, $\bar\delta=D_r\,\delta$.
This reduces $\kappa_2$ while preserving invariance.

\paragraph{QR with column pivoting (QRCP).}
For $G$, compute $G\Pi = Q R$ with permutation $\Pi$. Then
\[
\hat\varepsilon=\Pi\,R_1^{-1}\,Q_1^\top\,\delta,
\]
where $R_1$ is the leading $4\times4$ block if $\mathrm{rank}(G)=4$.
QRCP avoids squaring the condition number (unlike normal equations).

\paragraph{Left preconditioning (WLS).}
With $C=W^{1/2}$, solve $\min_\varepsilon\|\tilde G\varepsilon-\tilde\delta\|_2$
by QR/SVD on $(\tilde G,\tilde\delta)$.

\subsection{Regularization (Tikhonov) and filter factors}
When $G$ is ill-conditioned or noisy, use zero-order Tikhonov:
\begin{equation}
\varepsilon_\lambda=(G^\top W G+\lambda I)^{-1}G^\top W \delta,\qquad \lambda>0.
\label{eq:tikhonov}
\end{equation}
With $\tilde G=C G$ ($C=W^{1/2}$) and SVD $\tilde G=U\Sigma V^\top$,
\begin{equation}
\varepsilon_\lambda=V\,\underbrace{\mathrm{diag}\!\Big(
\frac{\sigma_i}{\sigma_i^2+\lambda}\Big)}_{=:F_\lambda(\Sigma)}\,U^\top\,\tilde\delta,
\label{eq:filter_factors}
\end{equation}
showing the \emph{filter factors} $f_i(\lambda)=\sigma_i/(\sigma_i^2+\lambda)$
that suppress small singular directions.

\paragraph{Choosing $\lambda$ (falsifiable rules).}
\begin{itemize}
\item \textbf{Discrepancy principle:} pick $\lambda$ s.t.
$\|W^{1/2}(G\varepsilon_\lambda-\delta)\|_2 \approx \sqrt{m}\,\sigma_\eta$
for noise level $\sigma_\eta$.
\item \textbf{L-curve:} choose the corner of $\big(\log\|G\varepsilon_\lambda-\delta\|,
\ \log\|\varepsilon_\lambda\|\big)$.
\item \textbf{GCV (if cross-validation feasible):} minimize the generalized
cross-validation score for WLS.
\end{itemize}

\subsection{Backward/forward errors and residual diagnostics}
Let $r:=G\hat\varepsilon-\delta$ and define the weighted residual norm
$\|r\|_W=\|W^{1/2}r\|_2$.

\paragraph{Forward error proxy.}
If a ground truth is unknown, estimate sensitivity by
\[
\mathrm{Amplification} := \|\tilde G^{\dagger}\|_2 \quad\text{with}\quad
\tilde G=W^{1/2}G.
\]

\paragraph{Backward error (square case).}
The minimal perturbation $(\Delta A,\Delta\delta)$ such that
$(A+\Delta A)\hat\varepsilon=\delta+\Delta\delta$ obeys
\[
\frac{\|\Delta\delta\|_2}{\|\delta\|_2}
\ \le\ \frac{\|r\|_2}{\|A\|_2\,\|\hat\varepsilon\|_2+\|\delta\|_2}.
\]

\paragraph{Leverage and influence (overdetermined).}
Define the \emph{hat matrix}
\[
H:=G(G^\top W G)^{-1}G^\top W=\tilde G\,\tilde G^\dagger,
\]
then $h_{ii}$ (diagonal of $H$) measures the leverage of equation $i$.
Standardized residuals $s_i:=r_i/\sqrt{(1-h_{ii})\,\hat\sigma^2}$ flag outliers
(with $\hat\sigma^2$ estimated from $\|r\|_W^2/(m-4)$).

\subsection{Uncertainty propagation (local, dimensionless)}
If $\delta$ carries noise with covariance $\mathrm{Cov}(\delta)=\Sigma$ and
$W=\Sigma^{-1}$, then the WLS covariance is
\begin{equation}
\mathrm{Cov}(\hat\varepsilon)\approx (G^\top W G)^{-1}.
\label{eq:cov_WLS}
\end{equation}
If $\Sigma=\sigma_\eta^2\,W^{-1}$ (homoscedastic in $W$-metric),
\(
\mathrm{Cov}(\hat\varepsilon)\approx \sigma_\eta^2\,(G^\top W G)^{-1}.
\)

\subsection{Numerical solvers and recommendations}
\begin{enumerate}
\item \textbf{Square:} use QR or LU with equilibration; verify residual and
backward error. Avoid forming $A^{-1}$ explicitly in floating point unless needed.
\item \textbf{WLS:} form $\tilde G=W^{1/2}G$; solve by QR/SVD; avoid normal
equations if conditioning is marginal.
\item \textbf{Tikhonov:} prefer SVD route when $\kappa_2(\tilde G)$ is large;
use filter factors \eqref{eq:filter_factors}.
\item \textbf{Diagnostics:} always report $\|r\|_W$, $\sigma_{\min}(\tilde G)$,
$\kappa_2(\tilde G)$, and, when applicable, leverage $h_{ii}$.
\end{enumerate}

\subsection{Falsifiable test plan (local-only)}
\begin{enumerate}
\item \emph{Noise scaling:} inject Gaussian noise with known $\sigma_\eta$;
verify $\|W^{1/2}(G\hat\varepsilon-\delta)\|_2\approx \sqrt{m}\,\sigma_\eta$
(discrepancy) and empirical coverage from \eqref{eq:cov_WLS}.
\item \emph{Ill-conditioning:} synthesize $G$ with a controlled singular value
drop; show that Tikhonov improves stability and that the error follows
filter factors.
\item \emph{Outlier resistance:} flip a small subset of equations; leverage
and standardized residuals flag them; refit after downweighting.
\end{enumerate}

\subsection{Compact algorithms (pseudo-code)}
\paragraph{WLS via QR (local, dimensionless).}
\begin{enumerate}
\item Build $\tilde G=W^{1/2}G$, $\tilde\delta=W^{1/2}\delta$.
\item Compute $\tilde G=Q R$ (thin QR); solve $R\hat\varepsilon=Q^\top\tilde\delta$.
\item Diagnostics: residual $\|Q^\top(\tilde G\hat\varepsilon-\tilde\delta)\|_2$,
$\sigma_{\min}$ via $\mathrm{chol}(R^\top R)$ or SVD if needed.
\end{enumerate}

\paragraph{Tikhonov via SVD.}
\begin{enumerate}
\item SVD: $\tilde G=U\Sigma V^\top$.
\item Choose $\lambda$ (discrepancy/L-curve/GCV).
\item Compute $\varepsilon_\lambda=V F_\lambda(\Sigma)U^\top\tilde\delta$.
\end{enumerate}

\paragraph{Square solve with backward error.}
\begin{enumerate}
\item Solve $A\hat\varepsilon=\delta$ (QR/LU).
\item Compute $r=A\hat\varepsilon-\delta$ and backward error proxy
$\eta=\|r\|_2 / (\|A\|_2\|\hat\varepsilon\|_2+\|\delta\|_2)$.
\item If $\eta$ exceeds tolerance, apply equilibration or switch to WLS/Tikhonov.
\end{enumerate}

\paragraph{Takeaway.}
The seed closure admits robust, dimensionless solvers with quantifiable
sensitivity, diagnostics, and uncertainty. QR/SVD with WLS/Tikhonov gives
portable routines that respect the governance rules and remain K-free/local-only.
% ===================== End Chapter 13 =====================

% ===================== Chapter 14 =====================
\section{Algorithms in Residual Space: Gradient, (Gauss-)Newton, and Line Search}

\subsection{Problem statements (local-only)}
Given a residual map $r(\varepsilon)$ and a weight $W\succeq0$, define
\[
V(\varepsilon)\;:=\;\tfrac12\,r(\varepsilon)^\top W\,r(\varepsilon).
\]
Two cases appear throughout the monograph:
\begin{enumerate}
\item \textbf{Linear seed/data-rich:} $r(\varepsilon)=G\varepsilon-\delta$ with $G\in\mathbb R^{m\times 4}$; $A$ is the $4\times4$ seed case (Ch.~4).
\item \textbf{Mildly nonlinear:} $r(\varepsilon)=F(\varepsilon)-\delta$, with $F:\mathbb R^{4}\to\mathbb R^{m}$ smooth and $J(\varepsilon):=\partial F/\partial\varepsilon$.
\end{enumerate}
All quantities are in the hat-frame (dimensionless; Option~A), evaluated locally at each space–time location.

\subsection{Gradient flow and gradient descent}
The weighted gradient and a continuous-time flow are
\[
\nabla V(\varepsilon)=J(\varepsilon)^\top W\,r(\varepsilon),\qquad
\dot\varepsilon=-\Lambda\,\nabla V(\varepsilon),
\]
with $\Lambda\succeq0$ (Ch.~5). A basic discrete update is
\begin{equation}
\varepsilon^{k+1}=\varepsilon^k-\eta_k\,\Lambda_k\,\nabla V(\varepsilon^k),
\label{eq:gd}
\end{equation}
where $\eta_k>0$ is the stepsize and $\Lambda_k$ a (possibly changing) preconditioner.
For the linear case $r=G\varepsilon-\delta$, $\nabla V=G^\top W(G\varepsilon-\delta)$.

\subsection{Gauss--Newton and Levenberg--Marquardt}
For $r(\varepsilon)=F(\varepsilon)-\delta$, the Gauss--Newton step solves
\begin{equation}
\big(J^\top W J\big)\,p_k=-J^\top W\,r,\qquad \varepsilon^{k+1}=\varepsilon^k+p_k.
\label{eq:GN}
\end{equation}
When $J^\top W J$ is ill-conditioned or rank-deficient, use the LM (trust-region) variant
\begin{equation}
\big(J^\top W J+\lambda_k I\big)\,p_k=-J^\top W\,r,\qquad \lambda_k>0,
\label{eq:LM}
\end{equation}
with $\lambda_k$ adjusted by a ratio test (see below). For linear $F(\varepsilon)=G\varepsilon$, \eqref{eq:GN} reduces to the WLS normal equations (Ch.~13).

\subsection{Line searches (Armijo/Wolfe) and step selection}
Given a descent direction $d_k$ (from \eqref{eq:gd}, \eqref{eq:GN}, or \eqref{eq:LM}), choose $\eta_k$ by backtracking to satisfy Armijo (sufficient decrease):
\[
V(\varepsilon^k+\eta_k d_k)\ \le\ V(\varepsilon^k)+c_1\,\eta_k\,\nabla V(\varepsilon^k)^\top d_k,\quad c_1\in(0,1).
\]
When curvature control is desired, enforce Wolfe conditions. For cheap heuristics, Barzilai--Borwein steps can be used in the linear case; ensure safeguards to keep $V$ decreasing.

\subsection{Trust-region update (ratio test)}
For LM, define the predicted reduction $\mathrm{pred}_k:= -\tfrac12\,p_k^\top(J^\top W J+\lambda_k I)\,p_k$ and the actual reduction $\mathrm{ared}_k:=V(\varepsilon^k)-V(\varepsilon^k+p_k)$. Let $\rho_k:=\mathrm{ared}_k/\mathrm{pred}_k$:
\[
\lambda_{k+1}=
\begin{cases}
\tfrac{1}{\nu}\lambda_k, & \rho_k>\rho_{\text{hi}} \ (\text{step very successful}),\\[2pt]
\lambda_k, & \rho_{\text{lo}}\le \rho_k\le \rho_{\text{hi}},\\[2pt]
\nu\lambda_k, & \rho_k<\rho_{\text{lo}} \ (\text{reject or shrink step}),
\end{cases}
\]
with $0<\rho_{\text{lo}}<\rho_{\text{hi}}<1$ and $\nu>1$. Accept the step if $\rho_k\ge\rho_{\text{lo}}$.

\subsection{Preconditioning and scaling (dimensionless but numerically vital)}
Even in the hat-frame, numerical conditioning benefits from preconditioning.
\begin{itemize}
\item \textbf{Diagonal equilibration:} scale rows/cols of $G$ (or $J$) to unit magnitude; apply the same transform to $\varepsilon$ and $\delta$.
\item \textbf{Quasi-Newton metric:} choose $\Lambda_k\approx (G^\top W G)^{-1}$ (linear) or $\Lambda_k\approx (J^\top W J)^{-1}$ (nonlinear) when cheap approximations exist.
\item \textbf{Projected metric:} if constraints are present (next section), restrict the metric to the feasible tangent space.
\end{itemize}

\subsection{Constraints (box, $\ell_1$, and feasibility)}
Local constraints appear naturally (e.g., bounds for near-unity deviations).
\begin{enumerate}
\item \textbf{Box constraints:} $\ell\le \varepsilon\le u$ via projected gradient:
$\varepsilon^{k+1}=\Pi_{[\ell,u]}(\varepsilon^k-\eta_k\Lambda_k\nabla V)$.
\item \textbf{$\ell_1$-regularization (sparse deviations):} minimize 
$V+\lambda\|\varepsilon\|_1$ by soft-thresholding after a gradient or proximal step.
\item \textbf{Feasibility first:} if the initial $\varepsilon^0$ violates constraints, project once before iterations.
\end{enumerate}

\subsection{Stopping criteria and diagnostics}
Terminate when any of the following holds:
\[
\|\nabla V(\varepsilon^k)\|_\infty \le \epsilon_g,\quad
\|d_k\|_2 \le \epsilon_x(1+\|\varepsilon^k\|_2),\quad
\frac{|V(\varepsilon^{k})-V(\varepsilon^{k-1})|}{\max\{1,V(\varepsilon^{k-1})\}}\le \epsilon_V,
\]
or upon reaching a maximum iteration budget. Always report
$\|r\|_{W}$, $\sigma_{\min}$ of $(G^\top W G)$ or $(J^\top W J)$,
and (if LM) the accepted $\rho_k$ sequence.

\subsection{Convergence remarks (local)}
Under standard assumptions (Lipschitz $J$, positive-definite $J^\top W J$ on a neighborhood, and bounded line-search parameters), gradient descent with backtracking converges to a first-order stationary point. Gauss--Newton is locally (super)linear when second-order terms are small; LM inherits global robustness and local fast convergence as $\lambda_k\to0$.

\subsection{Pseudocode (local-only, dimensionless)}
\paragraph{Projected Gauss--Newton / LM (with backtracking).}
\begin{algorithmic}[1]
\State \textbf{Input:} initial $\varepsilon^0$, weight $W$, tolerance $(\epsilon_g,\epsilon_x,\epsilon_V)$, bounds $[\ell,u]$ (optional).
\For{$k=0,1,2,\dots$}
  \State Form $r=F(\varepsilon^k)-\delta$, $J=\partial F/\partial\varepsilon$.
  \State Compute search direction $p_k$ from \eqref{eq:GN} or \eqref{eq:LM}.
  \State If bounds: $p_k \leftarrow \Pi_{T_{\varepsilon^k}}(p_k)$ (project to feasible tangent).
  \State Backtracking: choose $\eta_k\in(0,1]$ (Armijo); set $\tilde\varepsilon=\Pi_{[\ell,u]}(\varepsilon^k+\eta_k p_k)$.
  \State If LM: evaluate $\rho_k$ and update $\lambda_{k+1}$ by the ratio test.
  \State Update: $\varepsilon^{k+1}\leftarrow \tilde\varepsilon$.
  \State Check stopping criteria; break if satisfied.
\EndFor
\end{algorithmic}

\paragraph{Projected preconditioned gradient (seed/data-rich).}
\begin{algorithmic}[1]
\State $\varepsilon^{k+1}=\Pi_{[\ell,u]}\big(\varepsilon^k-\eta_k\,\Lambda_k\,G^\top W(G\varepsilon^k-\delta)\big)$.
\State Choose $\eta_k$ by backtracking; update $\Lambda_k$ by diagonal equilibration.
\end{algorithmic}

\subsection{Falsifiable tests (governance-compliant)}
\begin{enumerate}
\item \textbf{Monotone decrease:} verify $V(\varepsilon^{k+1})\le V(\varepsilon^k)$ for accepted steps.
\item \textbf{Ill-conditioning ablation:} degrade conditioning of $G$/$J$; show LM stabilizes while plain GN fails.
\item \textbf{Constraint check:} inject infeasible $\varepsilon^0$; verify projection keeps $V$ decreasing and feasibility preserved.
\item \textbf{Noise robustness:} add noise with known covariance; assess coverage using $(G^\top W G)^{-1}$ as in Ch.~13.
\end{enumerate}

\subsection{Interfaces to other layers}
The residual-space algorithms are \emph{local-only} and feed:
(i) the field layer via source calibration and falsification (Ch.~9--10),
(ii) the phase layer via indices $\rho_{\mathrm{lock}},\chi_{\mathrm{lock}}$ (Ch.~8),
and (iii) the numerical PDE layer via data assimilation hooks (Ch.~12),
without introducing nonlocal dependencies.
\paragraph{Takeaway.} Residual-space optimization admits a principled toolbox---gradient, Gauss--Newton/LM, and line search/trust-region---that is dimensionless, K-free, and falsifiable, providing stable bridges between data $\delta$ and internal state $\varepsilon$.
% ===================== End Chapter 14 =====================

% ===================== Chapter 15 =====================
\section{Data Models \& Noise: Weights $W$, Covariances, and Calibration}

\subsection{Measurement model (local-only, non-circular)}
Let $\mathcal O[\Psi]$ be a local observable (amplitude, phase, bandpower, etc.)
built from the coherence field $\Psi$ (Ch.~10). Measurements are modeled as
\begin{equation}
\delta \;=\; \mathcal M\!\big(\mathcal O[\Psi]\big) \;+\; \eta,
\qquad \mathbb E[\eta]=0,
\label{eq:meas15}
\end{equation}
with $\eta$ a local noise term. The definition of $\delta$ contains no $\varepsilon$
(Seed variables), avoiding circularity. Residual inference proceeds via
$A\varepsilon=\delta$ (Ch.~4) or $G\varepsilon\simeq\delta$ (Ch.~13).

\subsection{Noise families and local statistics}
We consider dimensionless, local noise models:
\begin{align*}
\text{Gaussian: }& \eta\sim\mathcal N(0,\Sigma),\\
\text{Laplace: }& p(\eta)\propto \exp\!\big(-\|\eta\|_{1,\Lambda}\big),\\
\text{Student-\emph{t}}:&\ p(\eta)\propto \prod_i \big(1+\tfrac{1}{\nu}\eta_i^2/s_i^2\big)^{-(\nu+1)/2},
\end{align*}
where $\Sigma\succeq0$ is a local covariance, $\Lambda\succ0$ is a diagonal
scale for Laplace (componentwise), and $(\nu,s_i)$ are degrees-of-freedom and
scales for heavy tails. Families are chosen per-sensor or per-feature, locally
in space--time.

\subsection{Weights, whitening, and the Mahalanobis metric}
For Gaussian noise, the negative log-likelihood gives the \emph{Mahalanobis} loss
\begin{equation}
V(\varepsilon)=\tfrac12\,(G\varepsilon-\delta)^\top W\,(G\varepsilon-\delta),
\qquad W=\Sigma^{-1}\succeq 0,
\label{eq:mahal}
\end{equation}
(Ch.~13). Whitening uses $C=W^{1/2}$:
\[
\tilde G:=CG,\quad \tilde\delta:=C\delta,\quad
V(\varepsilon)=\tfrac12\|\tilde G\varepsilon-\tilde\delta\|_2^2.
\]
If $\Sigma$ is unknown, estimate \emph{locally} via windowed statistics
(rolling mean/variance or robust MAD) and form $W$ accordingly.

\subsection{Heteroscedasticity and correlation}
\paragraph{Diagonal (heteroscedastic).} If components have different local variances
$\sigma_i^2$, set $W=\mathrm{diag}(\sigma_1^{-2},\dots,\sigma_m^{-2})$.
\paragraph{Correlated.} For a short window $\mathcal W$ around $(\mathbf x,t)$
collect centered samples $\{\hat\delta^{(s)}\}$ and use
\[
\hat\Sigma=\frac{1}{|\mathcal W|-1}\sum_{s\in\mathcal W}\hat\delta^{(s)}\hat\delta^{(s)\top},
\qquad W=\hat\Sigma^{-1} \ \ \text{(regularized if needed)}.
\]
Regularize by $\hat\Sigma_\lambda=(1-\lambda)\hat\Sigma+\lambda\,\mathrm{diag}(\hat\Sigma)$
with $\lambda\in[0,1)$ to maintain positive definiteness.

\subsection{Robust losses via IRLS (local-only)}
To resist outliers or heavy tails, minimize a robust loss
\[
V_\rho(\varepsilon)=\sum_{i=1}^m \rho\big((r_i/\sigma_i)\big),
\quad r:=G\varepsilon-\delta,
\]
where $\rho$ is Huber or Tukey’s biweight and $\sigma_i$ are local scales.
\paragraph{IRLS (Iteratively Reweighted Least Squares).}
At iteration $k$, set weights $w_i^{(k)}:=\psi(z_i^{(k)})/z_i^{(k)}$ with
$z_i^{(k)}=r_i^{(k)}/\sigma_i$ and $\psi=\rho'$; solve
\[
\varepsilon^{(k+1)}=\arg\min_\varepsilon
\sum_i w_i^{(k)} \Big(\frac{r_i}{\sigma_i}\Big)^2
\quad\Longleftrightarrow\quad
\hat W^{(k)}=\mathrm{diag}\!\Big(\frac{w_i^{(k)}}{\sigma_i^2}\Big),\ 
(G^\top \hat W^{(k)} G)\varepsilon=G^\top \hat W^{(k)}\delta.
\]
Stop by residual/gradient tolerances (Ch.~14). All operations remain local.

\subsection{Bias, variance, and sandwich corrections}
When the noise model is misspecified, use the \emph{sandwich} covariance
\[
\widehat{\mathrm{Cov}}(\hat\varepsilon)
\approx (G^\top W G)^{-1}\Big(G^\top W\,\widehat S\,W G\Big)(G^\top W G)^{-1},
\quad \widehat S := \sum_i r_i^2\,e_i e_i^\top,
\]
where $e_i$ are standard basis vectors. For IRLS, plug in the effective weights.

\subsection{Local calibration to $\Omega_0$ and $W_0$ (Option A)}
All frequencies are reported in rad/s with anchor $\Omega_0$ (Ch.~2; Compatibility Note).
When Hz is needed, use $f_0=\Omega_0/(2\pi)$ for reporting only. Decompose
\begin{equation}
W=W_0\,\widetilde W,\qquad \Lambda=\lambda_0\,\widetilde\Lambda,
\label{eq:W_decomp}
\end{equation}
with $\widetilde W,\widetilde\Lambda$ dimensionless and local. $W_0>0$ sets the
absolute data-fidelity scale; tune $W_0$ by a discrepancy target
$\mathbb E[\|W^{1/2}(G\hat\varepsilon-\delta)\|_2^2]\approx m$ (per window).

\subsection{Windowing, stationarity, and short-time spectra}
To maintain locality, estimate statistics over short windows (e.g., $K$ samples or a
small spatial patch). For phase/frequency observables, use short-time Fourier
windows aligned with $\Omega_0$ so that variance estimates reflect the local cadence.
Avoid windows so long that nonstationarity contaminates $\Sigma$.

\subsection{Model selection and hyperparameters (falsifiable)}
Choose among Gaussian/Laplace/$t$ and set hyperparameters by local cross-validation
or information criteria computed per window. A configuration is accepted only if:
(i) Energy--Coherence budget holds (Ch.~10), and
(ii) residual diagnostics (Ch.~13) pass thresholds.

\subsection{Diagnostics and acceptance rules}
Always report (per window):
\[
\|r\|_{W},\quad \sigma_{\min}(G^\top W G),\quad
\kappa_2(G^\top W G),\quad
\text{standardized residuals},\quad
\text{(if robust) IRLS weights}.
\]
Accept a run if $\|r\|_{W}^2\approx m$ within tolerance and discrete energy checks
(Ch.~12) show no growth beyond source work.

\subsection{Falsifiable test plan}
\begin{enumerate}
\item \textbf{Noise scaling:} inject synthetic Gaussian noise with known variances;
verify that the estimated $W$ makes $\|W^{1/2}r\|_2^2\approx m$.
\item \textbf{Outliers:} corrupt a small fraction of entries; robust IRLS should
recover with bounded influence and improved coverage.
\item \textbf{Correlation:} introduce known local covariance; whitening should
restore isotropic residuals (no spurious growth in $E_{\text{coh}}$).
\item \textbf{Nonstationarity:} shorten/lengthen windows; detect bias/variance trade-offs
and pick the smallest window that stabilizes diagnostics.
\end{enumerate}

\paragraph{Takeaway.}
Local, dimensionless weighting via $W=\Sigma^{-1}$ (with robust/whitened variants)
turns raw measurements into principled residuals for $A\varepsilon=\delta$.
Calibrating $(\Omega_0,W_0)$ and enforcing falsifiable diagnostics keeps the data
layer portable, K-free, and consistent with the Energy--Coherence Law.
% ===================== End Chapter 15 =====================

% ===================== Chapter 16 =====================
\section{Numerical Experiments \& Benchmarks (Dimensionless Protocols)}

\subsection{Principles and reporting}
All experiments must be: (i) dimensionless (Option A, rad/s), (ii) K-free and
local-only, (iii) falsifiable with pass/fail criteria, and (iv) reproducible.
Each run reports
\[
E_{\text{coh}}(t),\quad V(t),\quad \rho_{\mathrm{lock}}(t),\quad \chi_{\mathrm{lock}}(t),
\]
plus solver diagnostics (Chapter~12) and data diagnostics (Chapters~13--15).

\subsection{Common setup (domain, grids, time)}
Let $\Omega=[0,L]^d$ with $d\in\{1,2\}$, periodic BC unless stated otherwise.
Use $N$ grid points per direction ($\Delta x=L/N$) and total time $T$ with
step $\Delta t$. For Option A:
\[
\text{report }\Omega_0\ (\text{rad/s}),\quad f_0=\Omega_0/(2\pi)\ \text{(Hz, if needed)},
\]
and $(\tau,c)$ dimensionless. Unless otherwise noted, initialize
$\Psi(\mathbf x,0)$ as a band-limited packet and set $\partial_t\Psi(\mathbf x,0)=0$.

\subsection{Canonical test problems}
\paragraph{TP1: Homogeneous decay (no source).}
Telegraph core without source: validate $\dot E_{\text{coh}}\le 0$ and numerical
dispersion vs.\ $s_\pm(k)$ (Chapter~6). \emph{Pass} if energy decreases and
mode decay matches the analytic rate within tolerance.

\paragraph{TP2: Parametric drive with saturation.}
Add $\mathsf f_{\text{drv}}=-\alpha_1\Psi-\alpha_3|\Psi|^2\Psi
+\beta_{\text{pm}}\cos(\Omega_0 t)$ and verify bounded $E_{\text{coh}}$
with lock indices trending upward. \emph{Fail} if energy grows without bound.

\paragraph{TP3: Phase-locked feedback.}
Couple to a phase $\varphi$ (Chapter~8) with $\alpha_{\mathrm{cpl}}>0$ and
measure $(\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$. \emph{Pass} if
locking emerges without violating $\Re s<0$.

\paragraph{TP4: Residual-aware correction.}
Inject a small seed-aware feedback $-\gamma_r\mathcal M(\Psi)\,r$, $r:=A\varepsilon-\delta$;
\emph{pass} if $V$ decreases on average while $E_{\text{coh}}$ remains nonincreasing.

\paragraph{TP5: Data-rich inference.}
Synthesize additional near-unity relations $G\varepsilon\simeq\delta$; solve by
WLS/Tikhonov (Ch.~13) and verify coverage and discrepancy (Ch.~15).

\subsection{Metrics and acceptance (pass/fail)}
\paragraph{Energy--Coherence.}
Compute
\[
E_{\text{coh}}(t)=\int_\Omega\!\Big(\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2+\Phi(\Psi)\Big)\,d\mathbf x,
\]
with $\Phi$ chosen per Chapter~10. Criterion: $E_{\text{coh}}(t_{n+1})\le E_{\text{coh}}(t_n)$
up to source work under admissibility (Ch.~9).

\paragraph{Residual consistency.}
With $r:=G\varepsilon-\delta$ and $W$, track $V=\tfrac12\,r^\top W r$.
Criterion: when in the locking regime (high $\rho_{\mathrm{lock}},\chi_{\mathrm{lock}}$),
$\Delta V\le0$ on average over windows.

\paragraph{Locking indices.}
Compute (Chapter~8)
\[
\rho_{\mathrm{lock}}=\Big|\frac{1}{|\Omega|}\int_\Omega e^{i(\varphi-\theta(\Psi))}\,dx\Big|,\quad
\chi_{\mathrm{lock}}=1-\frac{\langle(\partial_t\varphi-\Omega_0)^2\rangle}{\langle(\partial_t\varphi)^2\rangle+\epsilon}.
\]
Acceptance: monotone approach to a plateau in TP3/TP2 with admissible sources.

\subsection{Grid/time convergence and dispersion checks}
Run TP1 on grids $N=2^k N_0$, $k=0,1,2$, and time steps with fixed CFL or
implicit stability. Measure $L^2$ error against either the spectral solution
(Chapter~7) or a highly resolved reference; expect second-order slopes for
CN/IMEX/Newmark($\beta=\tfrac14$). For dispersion, compare numerical amplification
$g(\theta)$ with $e^{s_\pm(k)\Delta t}$; require
$\Re(\log g)/\Delta t<0$ for all grid modes.

\subsection{Ablations (governance-compliant)}
\paragraph{A1: Remove saturation.}
Set $\alpha_3=0$ in TP2; \emph{fail} expected via energy growth or loss of
$\Re s<0$. Restore $\alpha_3>0$ and re-test (\emph{pass}).

\paragraph{A2: Remove gradient in phase flow.}
Set $\Gamma_\varphi=0$ in Chapter~8 coupling; locking fails (\emph{fail}).
Restore $\Gamma_\varphi>0$ (\emph{pass}).

\paragraph{A3: Remove flux.}
Set $D=\eta=0$ in flux; observe degraded stability. Restore admissible flux
and re-test.

\subsection{Noise and robustness (data layer)}
Generate $\eta$ with known local covariance $\Sigma$ and set $W=\Sigma^{-1}$.
\emph{Pass} if $\|W^{1/2}r\|_2^2\approx m$ (windowed discrepancy) and sandwich
covariance (Ch.~15) predicts empirical coverage. For outliers, run IRLS and
report leverage and standardized residuals (Ch.~13).

\subsection{Calibration to $\Omega_0$ and $W_0$}
Report $\Omega_0$ (rad/s) and, if needed, $f_0=\Omega_0/(2\pi)$ for figures.
Use the decomposition $W=W_0\widetilde W$, $\Lambda=\lambda_0\widetilde\Lambda$
with dimensionless tilded objects; tune $W_0$ by discrepancy targets, and
$\lambda_0$ by desired residual decay rate.

\subsection{Benchmark suites (portable templates)}
\paragraph{Suite S1 (linear restorative).}
Parameters: $(\tau,c,\alpha_1)$; sources: restorative only; goals:
energy decay, dispersion match, second-order convergence.

\paragraph{Suite S2 (saturating with parametric drive).}
Parameters: $(\tau,c,\alpha_1,\alpha_3,\beta_{\text{pm}})$; goals:
bounded energy and controlled lock indices.

\paragraph{Suite S3 (phase feedback).}
Parameters: $(\tau,c,\kappa_\varphi,\Gamma_\varphi,\alpha_{\mathrm{cpl}})$;
goals: emergence of locking and no violation of telegraph window.

\paragraph{Suite S4 (data-rich inference).}
Parameters: $(W_0,\widetilde W)$ and noise specs; goals: coverage,
discrepancy, and conditioning diagnostics.

\subsection{Reproducibility kit (deliverables)}
Every benchmark release must include:
\begin{enumerate}
\item Versioned code and parameter files (text-readable).
\item Seeds for noise and initial conditions.
\item Grid/time information, BCs, and solver options (CN/IMEX/Newmark etc.).
\item CSV/NPZ of $(E_{\text{coh}},V,\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$ time series.
\item A short \emph{protocol sheet}: what constitutes pass/fail and why.
\end{enumerate}

\paragraph{Takeaway.}
These dimensionless, K-free protocols make experiments portable across domains,
falsifiable via explicit pass/fail checks, and reproducible by construction.
They connect the telegraph PDE, residual inference, phase coupling, and data
layers under a single reporting grammar.
% ===================== End Chapter 16 =====================

% ===================== Chapter 17 =====================
\section{Case Study — Neural Dynamics (Mesoscopic Coherence)}

\subsection{Anchors, scales, and interpretation}
We adopt the Option~A normalization (rad/s; hats suppressed). The central cadence
$\Omega_0$ is chosen from a narrowband neural rhythm (e.g., $\beta$ or $\gamma$).
Report $f_0=\Omega_0/(2\pi)$ when Hz units are needed for figures. The coherence
field $\Psi(\mathbf x,t)$ represents the local envelope (or analytic amplitude)
of the narrowband population activity; the phase $\varphi(\mathbf x,t)\in\mathbb S^1$
represents a local phase order parameter (Chapter~8).

\subsection{Neural telegraph core with admissible sources}
We use the telegraph-stable core (Chapter~6) with local source grammar (Chapter~9):
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi-c^2\nabla^2\Psi
= -\alpha_1\Psi-\alpha_3|\Psi|^2\Psi
+ \beta_{\mathrm{pm}}\,\cos(\Omega_0 t+\phi_0)
+ \alpha_{\mathrm{cpl}}\sin\!\big(\varphi-\theta(\Psi)\big)
+ \xi(\mathbf x,t),
\label{eq:neuro_telegraph}
\end{equation}
with $\tau,c,\alpha_1,\alpha_3,\beta_{\mathrm{pm}},\alpha_{\mathrm{cpl}}\ge 0$
dimensionless and local-only (space--time dependent if desired). The stochastic
term $\xi$ is zero-mean, white-in-time, locally parameterized (Chapter~11).

\subsection{Phase coupling and gradient flow}
The local phase subsystem (Chapter~8) reads
\begin{align}
\partial_t\varphi &=
-\,\Gamma_\varphi\Big(-\kappa_\varphi\nabla^2\varphi
+\partial_\varphi U(\varphi;m_{\mathrm{eff}})
+ J(\Psi)\,\sin\!\big(\varphi-\theta(\Psi)\big)\Big)
+ \zeta(\mathbf x,t),
\label{eq:neuro_phase}
\end{align}
with $\Gamma_\varphi,\kappa_\varphi>0$, periodic potential $U$, and local gain
$J(\Psi)\ge0$. Noise $\zeta$ is zero-mean, causal, and local. The pair
\eqref{eq:neuro_telegraph}–\eqref{eq:neuro_phase} is well-posed under the rules
of Chapter~11.

\subsection{Observables and measurement map}
Let $\mathcal O[\Psi]$ denote an observable computed locally from the simulated
$\Psi$ that corresponds to an experimental measurement (e.g., instantaneous
power or Hilbert-envelope magnitude). Define the local measurement model
\begin{equation}
\delta=\mathcal M\!\big(\mathcal O[\Psi]\big)+\eta,\qquad \mathbb E[\eta]=0,
\label{eq:neuro_measure}
\end{equation}
with $\eta$ local noise. The residual inference follows $A\varepsilon=\delta$
(Chapter~4) or $G\varepsilon\simeq\delta$ (Chapter~13), and $W$ is built from
local noise estimates (Chapter~15). No circularity: $\delta$ depends on $\Psi$
only; $\varepsilon$ is inferred from $\delta$ only.

\subsection{Energy--coherence law and locking indices}
Define the coherence energy (Chapter~10)
\[
\mathcal E_{\text{coh}}(\Psi)=\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2
+\tfrac{\alpha_1}{2}|\Psi|^2+\tfrac{\alpha_3}{4}|\Psi|^4,\quad
E_{\text{coh}}(t)=\int_\Omega \mathcal E_{\text{coh}}\,dx.
\]
Locking indices (Chapter~8) quantify neural phase/frequency alignment to $\Omega_0$:
\[
\rho_{\mathrm{lock}}(t)=\Big|\tfrac{1}{|\Omega|}\int_\Omega e^{i(\varphi-\theta(\Psi))}\,dx\Big|,
\qquad
\chi_{\mathrm{lock}}(t)=1-\frac{\langle(\partial_t\varphi-\Omega_0)^2\rangle}{\langle(\partial_t\varphi)^2\rangle+\epsilon}.
\]
Admissible sources guarantee $\dot E_{\text{coh}}\le 0$ (modulo source work),
and in the locking regime we expect a decreasing residual energy
$V=\tfrac12(A\varepsilon-\delta)^\top W (A\varepsilon-\delta)$ on average.

\subsection{Parameter regimes (dimensionless, local)}
Select regimes representative of mesoscopic rhythms:
\begin{itemize}
\item \textbf{Restorative/saturating:} $\alpha_1\in[0.05,0.5]$, $\alpha_3\in[0.1,1]$.
\item \textbf{Propagation:} $c\in[0.3,1.2]$ (group velocity scale), $\tau\in[0.05,0.5]$.
\item \textbf{Parametric drive:} $\beta_{\mathrm{pm}}\in[0,0.5]$; phase offset $\phi_0\in[0,2\pi)$.
\item \textbf{Phase coupling:} $\Gamma_\varphi\in[0.1,1]$, $\kappa_\varphi\in[0.01,0.2]$, $J(\Psi)=J_0\,\min\{1,|\Psi|\}$ with $J_0\in[0,1]$, $\alpha_{\mathrm{cpl}}\in[0,1]$.
\end{itemize}
All can vary slowly in space to encode cortical heterogeneity while remaining local-only.

\subsection{Neural benchmarks (falsifiable)}
We adopt the dimensionless protocols of Chapter~16 with neural specializations.
\paragraph{NB1: Homogeneous decay.}
Set $\beta_{\mathrm{pm}}=\alpha_{\mathrm{cpl}}=0$, random band-limited initial
$\Psi$, periodic BC. \emph{Pass:} $E_{\text{coh}}$ decreases; modal decay matches
telegraph dispersion.

\paragraph{NB2: Locked response under admissible drive.}
Enable a small parametric drive ($\beta_{\mathrm{pm}}>0$) with saturation
($\alpha_3>0$). \emph{Pass:} bounded $E_{\text{coh}}$ and increasing
$(\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$; \emph{fail:} growth due to violating
small-gain bounds.

\paragraph{NB3: Phase feedback.}
Activate \eqref{eq:neuro_phase} and $\alpha_{\mathrm{cpl}}>0$; \emph{pass:}
emergent locking without violating $\Re s<0$; \emph{ablation fail:} set
$\Gamma_\varphi=0$ and observe loss of locking.

\paragraph{NB4: Residual consistency.}
Construct $\delta$ via \eqref{eq:neuro_measure}; solve $A\varepsilon=\delta$ and
track $V$. \emph{Pass:} during high locking, $V$ decreases on average; \emph{fail:}
miscalibrated sources produce sustained $V$ growth.

\subsection{Measurement, weights, and diagnostics}
Choose $\mathcal O[\Psi]$ as a local envelope or bandpower; define
$\delta=\mathcal M(\mathcal O[\Psi])+\eta$ with locally estimated covariance
$\Sigma$ and $W=\Sigma^{-1}$ (Chapter~15). Diagnostics include
$\|W^{1/2}(G\varepsilon-\delta)\|_2^2$, $\sigma_{\min}(G^\top W G)$, and
discrete energy monitors (Chapter~12).

\subsection{Boundary conditions and heterogeneity}
Use periodic or Neumann BCs to emulate contiguous cortex patches. Spatially
varying $\alpha_1,\alpha_3,J_0$ encode heterogeneous excitability/connectivity.
All variations remain local; no nonlocal kernels are introduced.

\subsection{Representative outcomes (qualitative)}
\begin{itemize}
\item \textbf{Stable coherence:} With admissible parameters, $\Psi$ exhibits
bounded envelopes and traveling/standing waves with $\Re s<0$.
\item \textbf{Locking:} Under moderate drive or feedback, $\rho_{\mathrm{lock}}$
and $\chi_{\mathrm{lock}}$ rise towards a plateau near the cadence $\Omega_0$.
\item \textbf{Consistency:} In locked regimes, the residual energy $V$ declines
on average, aligning field-level coherence with seed-level consistency.
\end{itemize}

\paragraph{Takeaway.}
The neural case realizes a dimensionless, local-only, falsifiable link from
mesoscopic field dynamics to measurements and residual inference. Telegraph
stability, energy--coherence compatibility, and phase-lock emergence to $\Omega_0$
manifest without invoking any nonlocal knowledge.
% ===================== End Chapter 17 =====================

% ===================== Chapter 18 =====================
\section{Case Study — Diffusion--Reaction / Materials Systems}

\subsection{Anchors, interpretation, and scope}
We adopt the Option~A normalization (rad/s; hats suppressed). The coherence field
$\Psi(\mathbf x,t)$ represents a smoothed order parameter (e.g., phase fraction,
normalized composition, or an effective microstructural envelope). The central
cadence $\Omega_0$ plays the role of a reference timescale associated with
microscopic rearrangements; report $f_0=\Omega_0/(2\pi)$ in figures when Hz is
useful for comparison.

\subsection{Telegraph-stable core with local reaction/transport sources}
We use the telegraph equation (Chapter~6) with a dimensionless, local source
grammar (Chapter~9):
\begin{equation}
\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi
= -\alpha_1\Psi - \alpha_3|\Psi|^2\Psi
+ \underbrace{\mathcal R(\Psi;\theta)}_{\text{local reaction}}
+ \underbrace{\nabla\!\cdot\!\big(D(\Psi)\nabla\Psi + \eta(\Psi)\nabla\partial_t\Psi\big)}_{\text{flux}}
+ \xi(\mathbf x,t).
\label{eq:dr_telegraph}
\end{equation}
Here $\alpha_1,\alpha_3\ge0$ encode restorative/saturating effects; $D(\Psi)\ge0$
and $\eta(\Psi)\ge0$ are local transport moduli; $\mathcal R(\Psi;\theta)$ is a
local reaction term parameterized by $\theta$ (e.g., driving force, temperature).
The noise $\xi$ is zero-mean, causal, and local (Chapter~11).

\paragraph{Reaction templates (local-only).}
\begin{align}
\text{Logistic/precipitation: }\quad
&\mathcal R(\Psi;\theta)=k(\theta)\,\Psi(1-\Psi),\label{eq:logistic}\\
\text{Allen--Cahn-like (relaxation to wells): }\quad
&\mathcal R(\Psi;\theta)=-\beta(\theta)\,\Phi'(\Psi),
\quad \Phi(\Psi)=\tfrac{\alpha_1}{2}\Psi^2+\tfrac{\alpha_3}{4}\Psi^4.
\label{eq:allen_cahn_like}
\end{align}
Both are pointwise, Lipschitz on bounded sets, and compatible with the
Energy--Coherence budget (Chapter~10).

\subsection{Energy--coherence and admissibility}
Choose
\[
\mathcal E_{\text{coh}}(\Psi)=\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2+\Phi(\Psi),
\quad
\Phi(\Psi)=\tfrac{\alpha_1}{2}\Psi^2+\tfrac{\alpha_3}{4}\Psi^4.
\]
Under the flux form in \eqref{eq:dr_telegraph} with $D,\eta\ge0$ and reaction
templates \eqref{eq:logistic}--\eqref{eq:allen_cahn_like} (with saturating
$\Phi$), the Energy--Coherence budget satisfies $\dot E_{\text{coh}}\le0$ modulo
controlled source work (Chapter~10). The telegraph window $\Re s<0$ is preserved
provided small-gain/saturation bounds hold (Chapter~9).

\subsection{Boundary conditions and materials settings}
Take $\Omega\subset\mathbb R^d$ ($d\in\{1,2,3\}$). Typical BCs:
\begin{itemize}
\item \textbf{Periodic} for bulk periodic microstructures.
\item \textbf{Neumann} $\partial_n\Psi=0$ for symmetry or no-flux boundaries.
\item \textbf{Dirichlet} $\Psi=\Psi_{\mathrm{bc}}$ for imposed phase fraction on surfaces.
\end{itemize}
Well-posedness and regularity follow Chapter~11 under the local Lipschitz and
passivity assumptions.

\subsection{Observables, measurement map, and weights}
Define a local observable $\mathcal O[\Psi]$ matching experiment/simulation
(e.g., normalized XRD peak envelope, optical/IR transmittance proxy, impedance
magnitude). The measurement model is
\[
\delta=\mathcal M\big(\mathcal O[\Psi]\big)+\eta,\qquad \mathbb E[\eta]=0,
\]
with local noise $\eta$. Build $W=\Sigma^{-1}$ from local covariance estimates
(Chapter~15). Inference uses $A\varepsilon=\delta$ (Chapter~4) or its WLS/Tikhonov
extension (Chapters~13--14). No circularity: $\delta$ depends on $\Psi$ only.

\subsection{Parameter regimes (dimensionless, local)}
Illustrative dimensionless ranges:
\[
\tau\in[0.02,0.5],\quad c\in[0.2,1.0],\quad
\alpha_1\in[0,0.5],\quad \alpha_3\in[0.1,1.5],\quad
k(\theta),\beta(\theta)\in[0,1],
\]
and transport moduli $D(\Psi)\in[D_{\min},D_{\max}]$, $\eta(\Psi)\in[0,\eta_{\max}]$ with
$0\le D_{\min}\le D_{\max}=O(1)$, $0\le\eta_{\max}=O(1)$. All quantities may vary
slowly in space or time to encode heterogeneity, remaining local-only.

\subsection{Benchmarks (falsifiable, materials-focused)}
We reuse Chapter~16 protocols with the following specializations.

\paragraph{MB1: Homogeneous relaxation (no reaction/drive).}
Set $\mathcal R\equiv0$, $D,\eta$ constant. \emph{Pass:} monotone
$E_{\text{coh}}$ decrease and dispersion consistent with telegraph roots.

\paragraph{MB2: Reaction-controlled front.}
Activate $\mathcal R$ (e.g., logistic) with small $D$; observe sigmoidal growth
of the mean $\langle\Psi\rangle$ with bounded energy. \emph{Fail:} loss of
$\Re s<0$ if saturation is removed.

\paragraph{MB3: Diffusion-controlled smoothing.}
Set $\mathcal R\equiv0$, moderate $D(\Psi)$; verify front smoothing and energy
decay without overshoot; boundary conditions influence the front in expected ways.

\paragraph{MB4: Residual consistency.}
Construct $\delta$ from $\mathcal O[\Psi]$; solve $A\varepsilon=\delta$ and track
$V=\tfrac12(A\varepsilon-\delta)^\top W(A\varepsilon-\delta)$. During steady
front propagation (near a locked regime), $V$ should decrease on average.

\subsection{Numerical notes (discretization and stability)}
Spatially discretize with centered differences or finite elements; time integration
via CN/IMEX/Newmark (Chapter~12). Prefer implicit treatment for the diffusion
operator to avoid restrictive CFL, keeping sources explicit (or IMEX). Monitor
the discrete energy to enforce the Energy--Coherence law numerically.

\subsection{Representative outcomes (qualitative)}
\begin{itemize}
\item \textbf{Stable front propagation:} bounded $E_{\text{coh}}$, fronts with
finite width set by $c,\tau$ and reaction/transport parameters.
\item \textbf{Microstructure smoothing:} diffusion-dominated runs show curvature
driven smoothing with monotone energy decrease.
\item \textbf{Consistency with measurements:} when observables lock to $\Omega_0$
(e.g., cyclic loading or periodic driving), residual energy $V$ decreases on average.
\end{itemize}

\paragraph{Takeaway.}
The diffusion--reaction case demonstrates a K-free, local-only bridge from
telegraph-stable order-parameter dynamics to measurement-linked residual
inference, with explicit, falsifiable tests that preserve $\Re s<0$ and
the Energy--Coherence budget across materials scenarios.
% ===================== End Chapter 18 =====================

% ===================== Chapter 19 =====================
\section{Case Study — Fluids/Plasma-like Toy Model}

\subsection{Anchors, interpretation, and scope}
We adopt Option~A (rad/s; hats suppressed). Interpret $\Psi(\mathbf x,t)$ as a
dimensionless scalar amplitude tracing compressive energy packets (e.g., an
acoustic or magnetoacoustic envelope). The central cadence is $\Omega_0$; report
$f_0=\Omega_0/(2\pi)$ only for figures. All operators and sources remain local
and K-free.

\subsection{Telegraph--advection core with admissible sources}
Extend the telegraph-stable core (Chapter~6) by a local advection drift
$\mathbf u(\mathbf x,t)$:
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi - c^2\nabla^2\Psi
\;+\; \mathbf u(\mathbf x,t)\!\cdot\!\nabla\Psi
\;=\; -\alpha_1\Psi-\alpha_3|\Psi|^2\Psi
\;+\; \nabla\!\cdot\!\big(D(\Psi)\nabla\Psi\big)
\;+\; \xi(\mathbf x,t).
\label{eq:plasma_tele_adv}
\end{equation}
Here $c,\tau>0$; $\alpha_1,\alpha_3\ge0$ are restorative/saturating gains;
$D(\Psi)\ge0$ is a local diffusion modulus; and $\xi$ is causal, zero-mean,
local noise (Chapter~11). The drift $\mathbf u$ is prescribed or obtained from a
separate local rule (e.g., a filtered background flow) subject to boundedness.

\paragraph{Local admissibility.}
Assume $\mathbf u\in L^\infty$ and Lipschitz on bounded sets, with
$\|\mathbf u\|_\infty\le u_{\max}=O(1)$; $D(\Psi)\in[0,D_{\max}]$, smooth;
$\alpha_3>0$ when parametric drives are present.

\subsection{Energy--coherence and advection balance}
Define the coherence density with a drift contribution handled by
skew-symmetric work:
\[
\mathcal E_{\text{coh}}(\Psi)=\tfrac12|\partial_t\Psi|^2+\tfrac{c^2}{2}|\nabla\Psi|^2
+\tfrac{\alpha_1}{2}|\Psi|^2+\tfrac{\alpha_3}{4}|\Psi|^4,
\quad E_{\text{coh}}(t)=\int_\Omega \mathcal E_{\text{coh}}\,d\mathbf x .
\]
Multiplying \eqref{eq:plasma_tele_adv} by $\partial_t\Psi$ and integrating over
$\Omega$ with periodic or appropriate boundary conditions yields
\[
\dot E_{\text{coh}}(t)
= -\int_\Omega \big(\tau\,|\partial_t^2\Psi|^2+|\partial_t\Psi|^2\big)\,dx
-\int_\Omega \nabla\partial_t\Psi\cdot D(\Psi)\nabla\Psi\,dx
+ \underbrace{\int_\Omega \partial_t\Psi\,(\mathbf u\!\cdot\!\nabla\Psi)\,dx}_{\text{skew (transport)}} 
+ \int_\Omega \partial_t\Psi\,\xi\,dx .
\]
Under periodic/Neumann boundaries, the advection term is skew and integrates to
a boundary flux; thus it does not create energy. With $D(\Psi)\ge0$ the flux is
dissipative. Hence the Energy--Coherence budget (Chapter~10) persists.

\subsection{Measurement map and residual link}
Choose a local observable $\mathcal O[\Psi]$ relevant to fluids/plasma
(e.g., band-limited pressure envelope, integrated emission proxy, or synthetic
line-of-sight signal). The measurement is
\begin{equation}
\delta=\mathcal M\big(\mathcal O[\Psi]\big)+\eta,\qquad \mathbb E[\eta]=0.
\label{eq:plasma_meas}
\end{equation}
Build the weight $W$ from local noise statistics (Chapter~15); infer $\varepsilon$
via $A\varepsilon=\delta$ or $G\varepsilon\simeq\delta$ (Chapters~4 and 13).
Non-circularity is preserved: $\delta$ depends on $\Psi$ only.

\subsection{Parameter regimes (dimensionless, local)}
Illustrative ranges:
\[
\tau\in[0.02,0.5],\quad c\in[0.3,1.5],\quad
\alpha_1\in[0,0.5],\ \alpha_3\in[0.1,1.5],\quad
D(\Psi)\in[0,D_{\max}],\ D_{\max}=O(1),\quad
\|\mathbf u\|_\infty\le u_{\max}=O(1).
\]
All quantities may vary slowly in space–time while remaining local-only.

\subsection{Boundary conditions}
\begin{itemize}
\item \textbf{Periodic}: energy-conserving for the drift term; convenient for FFT
diagnostics.
\item \textbf{Neumann}: reflect advection; no normal flux for diffusion.
\item \textbf{Dirichlet}: inflow/outflow walls modeled by prescribed $\Psi$ on
the boundary; ensure compatibility at $t=0$.
\end{itemize}
Well-posedness follows Chapter~11 under Lipschitz and boundedness hypotheses.

\subsection{Benchmarks (falsifiable, fluids/plasma-focused)}
\paragraph{FB1: Advected decay (no source).}
Set $\alpha_1=\alpha_3=0$, $D>0$, and a steady $\mathbf u$. \emph{Pass:}
$E_{\text{coh}}$ decreases and the packet translates without growth.

\paragraph{FB2: Saturated wave packet.}
Turn on $\alpha_1>0$, $\alpha_3>0$ with moderate drift; verify bounded envelopes,
telegraph dispersion, and no energy injection by advection.

\paragraph{FB3: Sheared drift (heterogeneous $\mathbf u$).}
Specify $\mathbf u(\mathbf x)= (s\,y,0)$ (2D) with small shear $s$; \emph{pass:}
no violation of $\Re s<0$; discrete energy remains nonincreasing modulo work.

\paragraph{FB4: Residual consistency under advection.}
Construct $\delta$ via \eqref{eq:plasma_meas}; solve $A\varepsilon=\delta$; \emph{pass:}
during advected steady regimes, $V=\tfrac12(A\varepsilon-\delta)^\top W (A\varepsilon-\delta)$
decreases on average.

\subsection{Numerical notes (CN/IMEX with drift)}
Discretize advection by centered differencing with a skew-symmetric form or by
upwind (for robustness). Use CN/IMEX/Newmark (Chapter~12) with the diffusion
term implicit and advection/source explicit (or extrapolated). Monitor the
discrete energy; enforce small-gain/saturation to prevent growth.

\subsection{Representative outcomes (qualitative)}
\begin{itemize}
\item \textbf{Transported coherence:} packets advect with $\mathbf u$ while
damping via telegraph and diffusion; energy decays monotonically.
\item \textbf{Saturated stability:} with $\alpha_3>0$, parametric or restorative
drives do not cause runaway; spectra remain in the $\Re s<0$ window.
\item \textbf{Residual alignment:} measurement-consistent runs show decreasing $V$
when the field is in a drift-locked regime with stable envelopes.
\end{itemize}

\paragraph{Takeaway.}
Adding a local advection drift preserves the dimensionless, K-free, telegraph
stability of the field while enabling transport-like behaviors relevant to
fluids/plasma surrogates. Energy–Coherence, admissibility, and residual
consistency remain falsifiable and portable.
% ===================== End Chapter 19 =====================

% ===================== Chapter 20 =====================
\section{Cross-Domain Synthesis: Parameter Transfer and Scaling Rules}

\subsection{Similarity classes (Option A, rad/s)}
Two configurations $\mathcal P$ and $\mathcal P'$ (possibly in different
domains) are in the same similarity class if, after a local rescaling by a factor
$\lambda>0$,
\begin{equation}
(\kappa,\omega,b,\rho)\ \mapsto\ (\lambda^{2}\kappa,\ \lambda\,\omega,\
\lambda\,b,\ \lambda^{-1}\rho),
\qquad
(\tau,c,\alpha_1,\alpha_3)\ \mapsto\ (\lambda^{-1}\tau,\ \lambda\,c,\
\alpha_1,\ \alpha_3),
\label{eq:group_xfer}
\end{equation}
their \emph{invariants} match:
\begin{equation}
\Pi_1=\frac{\kappa}{\omega^2},\quad
\Pi_2=\frac{b}{\omega},\quad
\Pi_3=\omega\rho,
\qquad
\Theta_1:=c^2\tau,\quad
\Theta_2:=\frac{\alpha_1}{c^2},\quad
\Theta_3:=\frac{\alpha_3}{c^2}.
\label{eq:invariants}
\end{equation}
The \(\Pi\)'s arise from the seed algebra (Ch.~3--4); the \(\Theta\)'s from the
telegraph core (Ch.~6). Matching \((\Pi,\Theta)\) yields identical local
dispersion and near-unity algebra after rescaling time/space units.

\subsection{Transfer of the telegraph core}
Let $\Psi$ solve (Ch.~6) with parameters $(\tau,c,\alpha_1,\alpha_3)$ and valid
sources. Under \eqref{eq:group_xfer}, the rescaled field
\[
\Psi'(\mathbf x,t):=\Psi(\lambda^{-1}\mathbf x,\lambda\,t)
\]
satisfies a telegraph equation with $(\tau',c',\alpha_1',\alpha_3')$
given by \eqref{eq:group_xfer}. Its dispersion roots obey
\[
s'_\pm(k)=\lambda\,s_\pm(\lambda^{-1}k),
\]
so $\Re s'<0$ iff $\Re s<0$. Hence the \emph{telegraph stability window}
and the Energy--Coherence budget are invariant under transfer.

\subsection{Source grammar portability (local-only)}
Each source block transfers with a scale weight determined by its differential
order:
\begin{equation}
\nabla\cdot \big(D(\Psi)\nabla\Psi\big)\ \leadsto\ \lambda^{2}\,(\cdot),\qquad
\partial_t\Psi\ \leadsto\ \lambda\,(\cdot),\qquad
\partial_t^2\Psi,\ \nabla^2\Psi\ \leadsto\ \lambda^{2}\,(\cdot).
\label{eq:weights}
\end{equation}
To preserve admissibility (Ch.~9) after transfer, scale any \emph{active} drive
gain \(\beta_{\mathrm{pm}}\) so that
\[
\frac{\beta_{\mathrm{pm}}}{c^2}\ \text{is invariant},\qquad
\alpha_3>0\ \text{(kept)}.
\]
Thus small-gain/saturation bounds and $\Re s<0$ remain intact.

\subsection{Residual layer transfer}
The seed closure is algebraic: $A\varepsilon=\delta$ (Ch.~4). Under
\eqref{eq:group_xfer}, the near-unity algebra in hats is unchanged, hence
\begin{equation}
A' = A,\qquad \varepsilon'=\varepsilon,\qquad
\delta'=\delta,\qquad
V'=\tfrac12\,\delta^\top W'\delta,
\label{eq:residual_xfer}
\end{equation}
with $W'=W_0'\,\widetilde W$ and the same \(\widetilde W\) (dimensionless). Only
the scalar anchors $(W_0,\lambda_0)$ change,
\[
W'=W_0'\,\widetilde W,\qquad \Lambda'=\lambda_0'\,\widetilde\Lambda,
\]
to keep the same discrepancy targets and decay rates (Ch.~10).

\subsection{Phase coupling transfer}
For the phase energy (Ch.~8), set
\[
\kappa_\varphi'=\lambda^{0}\kappa_\varphi,\quad
\Gamma_\varphi'=\lambda^{1}\Gamma_\varphi,\quad
J'(\Psi')=J(\Psi),\quad
\theta(\Psi')=\theta(\Psi),
\]
so that the phase Lyapunov decrease is preserved under time rescaling $t\mapsto \lambda t$.

\subsection{Domain crosswalk (neural, materials, fluids-like)}
A compact mapping among the three case studies:

\begin{center}
\begin{tabular}{lccc}
\toprule
Quantity & Neural & Materials (DR) & Fluids-like \\
\midrule
$\Psi$ & band-limited envelope & phase-fraction envelope & compressive amplitude \\
$\mathsf f_{\text{drv}}$ & parametric drive & local reaction $\mathcal R(\Psi)$ & restorative/saturation \\
Flux & $D\nabla\Psi$ (diffusive) & $D\nabla\Psi+\eta\nabla\partial_t\Psi$ & $D\nabla\Psi$ \\
Advection & -- & -- & $\mathbf u\cdot\nabla\Psi$ \\
Locking metric & $(\rho_{\text{lock}},\chi_{\text{lock}})$ & residual consistency during cycles & drift-locked residual consistency \\
\bottomrule
\end{tabular}
\end{center}

Under \eqref{eq:group_xfer}--\eqref{eq:weights}, these ingredients match
one-to-one after unit rescaling and gain renormalization.

\subsection{Renormalization-style tuning (local recipe)}
Given a reference configuration $\mathcal P=(\tau,c,\alpha_1,\alpha_3,\beta_{\mathrm{pm}})$
and a target configuration $\mathcal P'$ with the same invariants \eqref{eq:invariants}:
\begin{enumerate}
\item Pick $\lambda=\dfrac{c'}{c}$ (space/time speed ratio).
\item Set $\tau'=\lambda^{-1}\tau$ and keep $\alpha_1'=\alpha_1$, $\alpha_3'=\alpha_3$.
\item Scale active drive $\beta_{\mathrm{pm}}'=\big(\tfrac{c'}{c}\big)^{2}\beta_{\mathrm{pm}}$
to keep $\beta_{\mathrm{pm}}/c^2$ invariant.
\item Keep $A$, $\widetilde W$, $\widetilde\Lambda$ unchanged; tune $(W_0',\lambda_0')$
by discrepancy and decay-rate targets (Ch.~10).
\end{enumerate}
This preserves dispersion, Lyapunov monotonicity, and residual consistency.

\subsection{Falsifiable transfer tests}
\begin{enumerate}
\item \textbf{Dispersion preservation:} compare $s_\pm(k)$ and $s'_\pm(k)$ via
the scaling rule $s'_\pm(k)=\lambda\,s_\pm(\lambda^{-1}k)$; require matching
real parts (stability) and scaled imaginary parts (oscillation).
\item \textbf{Energy budget:} verify $E_{\text{coh}}'(t)$ is nonincreasing modulo
source work, mirroring $E_{\text{coh}}(t)$ after rescaling time by $\lambda$.
\item \textbf{Residual statistics:} show that the whitened residual norm
$\|W^{1/2}(G\varepsilon-\delta)\|_2$ has the same distribution before/after transfer,
up to the scalar anchors $(W_0,\lambda_0)$.
\item \textbf{Locking behavior:} when applicable, demonstrate that peak values
and settling times of $(\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$ scale by
the predicted factor $\lambda$.
\end{enumerate}

\subsection{What does \emph{not} transfer (guardrails)}
The cross-domain map is \emph{local-only}. Nonlocal couplings, global constraints,
or history-dependent kernels (ruled out by the governance) are \emph{not}
transported by \eqref{eq:group_xfer}. Any empirical improvement from such terms
must be treated as a separate, explicitly declared model layer.

\paragraph{Takeaway.}
Similarity via \((\Pi,\Theta)\), combined with the rescaling
\eqref{eq:group_xfer} and the local source grammar, yields a portable and
falsifiable map between domains. It preserves the telegraph stability window,
Energy--Coherence budget, and residual-layer statistics while keeping the entire
construction dimensionless, K-free, and local-only.
% ===================== End Chapter 20 =====================

% ===================== Chapter 21 =====================
\section{Beyond Log-Linear: Second-Order Corrections and Validity Windows}

\subsection{Motivation and setting}
The hat-frame mother ratios (Ch.~4) are written as near-one quantities
$\Pi_i(\varepsilon)=1+\delta_i(\varepsilon)$ with $\|\varepsilon\|\ll1$.
Log-linearization yields the seed system $A\varepsilon=\delta$.
To quantify departures from strict linearity and extend the workable range of
$\|\varepsilon\|$, we keep quadratic terms from $\log(1+x)=x-\tfrac12x^2+O(x^3)$.

\subsection{Quadratic expansion of the mother map}
Let $\delta(\varepsilon)$ denote the vector whose components are the log-residuals:
$\delta_i(\varepsilon)=\log \Pi_i(\varepsilon)$.
A second-order Taylor expansion at $\varepsilon=0$ gives
\begin{equation}
\delta(\varepsilon)
= A\,\varepsilon \;+\; \tfrac12 \,\mathcal B[\varepsilon,\varepsilon] \;+\; R_3(\varepsilon),
\qquad \|R_3(\varepsilon)\|=O(\|\varepsilon\|^3),
\label{eq:quad_expansion}
\end{equation}
where $\mathcal B:\mathbb R^4\times\mathbb R^4\to\mathbb R^m$ is a symmetric
bilinear form that collects second derivatives at the unity state.
Equivalently, in components
\begin{equation}
\delta_i(\varepsilon)=\sum_{a} A_{ia}\,\varepsilon_a
+ \tfrac12\sum_{a,b} H^{(i)}_{ab}\,\varepsilon_a\varepsilon_b
+ O(\|\varepsilon\|^3),\qquad H^{(i)}_{ab}=H^{(i)}_{ba},
\label{eq:comp_form}
\end{equation}
and $\mathcal B[\varepsilon,\varepsilon]$ has entries
$[\mathcal B[\varepsilon,\varepsilon]]_i=\sum_{a,b}H^{(i)}_{ab}\varepsilon_a\varepsilon_b$.

\subsection{Quadratic residual model and inversion}
The second-order residual can be modeled as
\begin{equation}
\delta \;\approx\; A\,\varepsilon + \tfrac12\,\mathcal B[\varepsilon,\varepsilon],
\label{eq:quad_model}
\end{equation}
with two local inversion strategies:

\paragraph{(i) Corrected one-shot (small curvature).}
Solve the linear seed $\varepsilon^{(0)}=A^{-1}\delta$, then apply a Newton-like correction:
\begin{align}
\delta_{\text{curv}}^{(0)} &:= \tfrac12\,\mathcal B[\varepsilon^{(0)},\varepsilon^{(0)}],\nonumber\\
\varepsilon^{(1)} &:= A^{-1}\!\left(\delta-\delta_{\text{curv}}^{(0)}\right),
\label{eq:oneshot_correction}
\end{align}
repeat once if needed. This is accurate when $\|\mathcal B\|\,\|\varepsilon^{(0)}\|\ll\|A\|$.

\paragraph{(ii) Gauss--Newton on the quadratic map.}
Define the nonlinear residual
\[
r(\varepsilon):=A\varepsilon+\tfrac12\,\mathcal B[\varepsilon,\varepsilon]-\delta.
\]
Linearize at $\varepsilon^k$ using the Jacobian
\[
J(\varepsilon^k)=A+\mathcal B[\varepsilon^k,\cdot],
\]
and compute a step $p^k$ from $J(\varepsilon^k)^\top W J(\varepsilon^k)\,p^k
=-J(\varepsilon^k)^\top W r(\varepsilon^k)$ (WLS metric $W\succeq0$), then set
$\varepsilon^{k+1}=\varepsilon^k+p^k$ with trust-region or line-search safeguards.

\subsection{Bounding the remainder and defining validity windows}
Assume there exists $M_3>0$ such that $\|R_3(\varepsilon)\|\le \tfrac{M_3}{6}\|\varepsilon\|^3$
in a neighborhood of the origin, and let $\|\mathcal B\|$ denote the operator norm
of the bilinear form in \eqref{eq:quad_expansion}.
Then for the linear seed estimator $\varepsilon_{\text{lin}}:=A^{-1}\delta$
with exact data:
\begin{equation}
\|\varepsilon-\varepsilon_{\text{lin}}\|
\;\le\; \tfrac12\,\|A^{-1}\|\,\|\mathcal B\|\,\|\varepsilon\|^2
\;+\; \|A^{-1}\|\,\|R_3(\varepsilon)\|.
\label{eq:error_bound_linear}
\end{equation}
A practical \emph{validity radius} $r_\star$ for the linear model is the largest
$r$ such that
\begin{equation}
\tfrac12\,\|A^{-1}\|\,\|\mathcal B\|\,r \;+\; \tfrac{M_3}{6}\|A^{-1}\|\,r^2
\;\le\; \eta,
\label{eq:radius_condition}
\end{equation}
for a chosen tolerance $\eta\in(0,1)$ on the relative error
$\|\varepsilon-\varepsilon_{\text{lin}}\|/\max\{1,\|\varepsilon\|\}$.
Inside $B_{r_\star}(0)$, the linear approximation is deemed sufficient.

\subsection{Noise, weighting, and robustness}
With noisy $\delta$ and weight $W=\Sigma^{-1}$ (Ch.~15), define the weighted
quadratic loss
\[
V_2(\varepsilon)=\tfrac12\|A\varepsilon+\tfrac12 \mathcal B[\varepsilon,\varepsilon]-\delta\|_W^2.
\]
Gauss--Newton updates use $J(\varepsilon)=A+\mathcal B[\varepsilon,\cdot]$ as above.
If outliers/heavy tails are present, embed $V_2$ in IRLS (Ch.~15) by replacing $W$
with iteration-dependent diagonal weights; all operations remain local-only.

\subsection{Trust regions and acceptance rules}
Use a trust-region subproblem for $p$:
\[
\min_{\|p\|\le \Delta}\ \tfrac12\|Jp+r\|_W^2,
\]
update the radius $\Delta$ via the ratio of actual to predicted reduction.
\emph{Accept} steps that reduce $V_2$ and keep $\|\varepsilon\|$ inside the
estimated validity window; otherwise shrink $\Delta$ or revert to linear seed.

\subsection{Model selection: when to switch beyond linear}
Define the following falsifiable criteria:
\begin{enumerate}
\item \textbf{Curvature test:} compute $c:=\|\mathcal B[\varepsilon^{(0)},\varepsilon^{(0)}]\|/\|\delta\|$.
If $c\le c_{\min}$, stay linear; if $c\ge c_{\max}$, switch to quadratic.
\item \textbf{Residual split:} compare $V_{\text{lin}}=\tfrac12\|A\varepsilon^{(0)}-\delta\|_W^2$
and the one-shot corrected $V_{\text{1s}}=\tfrac12\|A\varepsilon^{(1)}+\tfrac12\mathcal B[\varepsilon^{(1)},\varepsilon^{(1)}]-\delta\|_W^2$.
\emph{Switch} if $V_{\text{1s}}$ drops by more than a preset fraction.
\item \textbf{Stability guard:} enforce $\|\varepsilon\|\le r_\star$ when using the
linear model; if violated, escalate to the quadratic solver.
\end{enumerate}

\subsection{Propagation to the field and phase layers}
Second-order corrections alter only the local residual map; the telegraph field
(Ch.~6--7) and phase layer (Ch.~8) remain unchanged. Consistency is maintained by
using the same admissibility and Energy--Coherence constraints (Ch.~9--10) and
feeding the improved $\varepsilon$ estimates into any residual-aware source terms
without introducing nonlocal dependencies.

\subsection{Computation of the curvature tensor}
For mother ratios that are rational monomials in the near-unity variables
(e.g., $\Pi_1=\kappa/\omega^2$, $\Pi_2=b/\omega$, $\Pi_3=\omega\rho$), the
entries $H^{(i)}_{ab}$ follow from differentiating $\log\Pi_i$ at unity.
All constants are dimensionless (Option~A). Store $\mathcal B$ as a set of four
symmetric $4\times4$ matrices $\{H^{(i)}\}$ if $m=4$, or as a rank-3 tensor for
general $m$; evaluate $\mathcal B[\varepsilon,\varepsilon]$ by symmetric products.

\subsection{Numerical checklist (local-only)}
\begin{enumerate}
\item Start with $\varepsilon^{(0)}=A^{-1}\delta$ (Ch.~4); compute $c$ and $r_\star$.
\item If the curvature test triggers, run Gauss--Newton/LM on $V_2$ with
line search or trust region (Ch.~14).
\item Use IRLS if robust losses are required (Ch.~15).
\item Report: $V_{\text{lin}}$, $V_2$, $\|\varepsilon\|$, $c$, and acceptance ratios.
\end{enumerate}

\paragraph{Takeaway.}
Second-order corrections furnish a controlled extension of the seed algebra:
they quantify curvature via a symmetric bilinear form, define explicit validity
windows, and provide practical algorithms (one-shot correction, Gauss--Newton/LM)
that remain dimensionless, local-only, and falsifiable.
% ===================== End Chapter 21 =====================

% ===================== Chapter 22 =====================
\section{Symmetry Breaking, Constraints, and Gauge Choices}

\subsection{Symmetry landscape (local-only)}
We recall the local scale map (Chs.~2--4)
\[
(\kappa,\omega,b,\rho)\ \mapsto\ (\lambda^2\kappa,\ \lambda\omega,\ \lambda b,\ \lambda^{-1}\rho),\qquad \lambda>0,
\]
under which the invariants $(\Pi_1,\Pi_2,\Pi_3)$ and the seed log-linearization
are preserved. At the field level, the telegraph core (Chs.~6--7) is invariant
under additive time shifts and, in the presence of a phase variable,
under local phase reparameterizations $\varphi\mapsto\varphi+\phi_0$ when
sources depend only on phase differences. These symmetries are \emph{local-only}
(no nonlocal kernels or global memory).

\subsection{Constraints vs.\ gauges: definitions}
\begin{itemize}
\item \textbf{Constraints} restrict feasible states (physical or calibration): 
box bounds on $\varepsilon$, equality relations among components, positivity of
weights $W$, etc. They alter the feasible set and can change the optimizer.
\item \textbf{Gauges} remove redundancies without changing physical predictions:
reference phase $\theta(\Psi)$, baseline offsets that cancel in observables,
or the choice of anchor units (rad/s vs.\ Hz). Fixing a gauge selects a
representative from an equivalence class.
\end{itemize}

\subsection{Residual layer with constraints (projection and multipliers)}
Given $r(\varepsilon)=G\varepsilon-\delta$ and weight $W\succeq0$ (Chs.~13--15),
constrained estimation solves
\begin{equation}
\min_{\varepsilon}\ \tfrac12\,\|r(\varepsilon)\|_W^2
\quad\text{s.t.}\quad
A_{\mathrm{eq}}\varepsilon=b_{\mathrm{eq}},\qquad \ell\le \varepsilon\le u.
\label{eq:residual_constrained}
\end{equation}
Two portable strategies:
\paragraph{(i) Projected metric.}
Let $\Pi_{\mathcal C}$ be the projection onto the box $[\ell,u]$; use
\[
\varepsilon^{k+1}=\Pi_{\mathcal C}\big(\varepsilon^k-\eta_k\,\Lambda_k\,G^\top W(G\varepsilon^k-\delta)\big)
\]
with $\Lambda_k\succeq0$ (Ch.~14). Equality constraints are enforced by projecting
the direction onto the nullspace of $A_{\mathrm{eq}}$.

\paragraph{(ii) Lagrange multipliers (KKT).}
Solve the saddle system for the linearized step $p$:
\[
\begin{bmatrix}
G^\top W G & A_{\mathrm{eq}}^\top\\
A_{\mathrm{eq}} & 0
\end{bmatrix}
\begin{bmatrix}
p\\ \lambda
\end{bmatrix}
=
-\begin{bmatrix}
G^\top W (G\varepsilon-\delta)\\
A_{\mathrm{eq}}\varepsilon-b_{\mathrm{eq}}
\end{bmatrix},
\]
then update $\varepsilon\leftarrow \Pi_{[\ell,u]}(\varepsilon+p)$. Both methods
are dimensionless and local-only.

\subsection{Gauge fixing in the phase/field layer}
The phase coupling (Ch.~8) uses a reference $\theta(\Psi)$ that enters only via
differences $\varphi-\theta(\Psi)$. To avoid drift in a neutrally stable phase
direction, choose a \emph{local gauge}:
\[
\theta(\Psi)=\arg(\Psi),\quad\text{or}\quad 
\theta(\Psi)=\mathcal G(\Psi)\ \text{with}\ \mathcal G\ \text{local, Lipschitz}.
\]
Gauge changes $\theta\mapsto \theta+\phi_0$ induce $\varphi\mapsto \varphi+\phi_0$
with no change to physical observables or the Energy--Coherence budget.
For the telegraph PDE, additive offsets in $\Psi$ are fixed by boundary/mean
conditions (periodic mean-zero or Dirichlet anchoring).

\subsection{Symmetry breaking: when and why}
Breaking a symmetry is legitimate when it:
(i) removes degeneracy that harms identifiability or numerical stability,
(ii) is realized by a gauge choice that leaves observables invariant, or
(iii) encodes a physical constraint (e.g., positivity or pinned boundary data).
Examples:
\begin{itemize}
\item \textbf{Phase pinning:} fix $\theta(\Psi)$ at a point or region to kill
global phase drift; $\rho_{\mathrm{lock}},\chi_{\mathrm{lock}}$ remain well-defined.
\item \textbf{Unit anchor:} adopt rad/s (Option A) globally; Hz appears only in
figures via $f_0=\Omega_0/(2\pi)$.
\item \textbf{Mean removal:} enforce $\int_\Omega \Psi\,dx=0$ (periodic), which
is a harmless gauge for dynamics that depend on gradients and time derivatives.
\end{itemize}

\subsection{Compatibility with Energy--Coherence and stability}
Any constraint or gauge must preserve the telegraph window and the energy law.
For the constrained residual problem, ensure the projected/augmented steps
remain descent directions in the metric induced by $W$:
\[
V(\varepsilon^{k+1})\le V(\varepsilon^k)\quad\text{with}\quad
V=\tfrac12\|G\varepsilon-\delta\|_W^2.
\]
For the field, gauge operations are algebraic and local; they do not alter
the dissipative terms (Ch.~10) nor the dispersion (Ch.~6).

\subsection{Local gauge for measurement maps (non-circular)}
If a measurement map admits an arbitrary baseline (e.g., a log-gain or offset),
fix it \emph{after} defining $\delta=\mathcal M(\mathcal O[\Psi])+\eta$ (Ch.~10),
never before. This prevents circular definitions linking $\delta$ to $\varepsilon$.
All calibrations (e.g., $W_0$) are reported explicitly and locally.

\subsection{Inequality constraints on phase and field}
Typical local constraints:
\[
|\varphi|\le \pi,\qquad \Psi_{\min}\le \Psi \le \Psi_{\max},\qquad
0\le J(\Psi)\le J_{\max}.
\]
Enforce them by projection in time-stepping (Ch.~12) or by proximal penalties:
\[
\min_{\Psi}\ \mathcal J(\Psi)
:= E_{\text{coh}}(\Psi) + \iota_{[\Psi_{\min},\Psi_{\max}]}(\Psi),
\]
where $\iota$ is the indicator of the feasible interval; implicit steps with
projection preserve the discrete energy decrease.

\subsection{Falsifiable checks (governance-compliant)}
\begin{enumerate}
\item \textbf{Gauge invariance:} run two simulations differing only by a constant
phase shift in $\theta$; show identical $(E_{\text{coh}},\rho_{\mathrm{lock}},\chi_{\mathrm{lock}})$
trajectories (up to the trivial shift).
\item \textbf{Constraint descent:} under box/equality constraints, verify monotone
decrease of $V$ for accepted steps; ablate projections to demonstrate failure.
\item \textbf{Mean removal neutrality:} with/without mean-zero gauge for periodic
BC, confirm identical dispersion and energy decay (differences appear only as a
constant offset).
\end{enumerate}

\subsection{Implementation notes (dimensionless)}
\begin{itemize}
\item Use rad/s anchors (Option A) throughout; do not mix Hz into algebra.
\item Prefer QR-based KKT solves for constrained WLS (avoid squaring the
condition number).
\item In time-stepping, apply projection after each implicit or IMEX step and
recompute the discrete energy to enforce the Energy--Coherence check.
\end{itemize}

\paragraph{Takeaway.}
Constraints change feasible sets; gauges remove redundancy. Both can be handled
locally—with projections, KKT systems, and phase pinning—without violating
telegraph stability, Energy--Coherence monotonicity, or non-circularity of the
measurement loop. The result is a symmetry-aware, dimensionless, K-free practice
for robust inference and evolution across the stack.
% ===================== End Chapter 22 =====================

% ===================== Chapter 23 =====================
\section{Stochastic Forcing \& Uncertainty Quantification in Residual/Field Layers}

\subsection{SPDE model (local-only, Option A)}
Let $\Psi(\mathbf x,t)$ satisfy the telegraph-stable core with admissible sources (Ch.~9--11) and additive local noise:
\begin{equation}
\tau\,\partial_t^2\Psi + \partial_t\Psi - c^2\nabla^2\Psi
= \mathsf S(\Psi,\varphi;\varepsilon,\delta) \;+\; \xi(\mathbf x,t),
\qquad \mathbb E[\xi]=0,
\label{eq:spde_core}
\end{equation}
where $\xi$ is white-in-time, $L^2(\Omega)$-valued, with spatial covariance operator $Q$:
\[
\mathbb E\big[\langle\xi(t),\phi\rangle \langle\xi(t'),\psi\rangle\big]
= 2\,\langle Q\phi,\psi\rangle\,\delta(t-t').
\]
All coefficients and statistics are \emph{local-only}, dimensionless, and respect Option~A (rad/s).

\subsection{Mild/semigroup formulation and energy in expectation}
Setting $Z=(\Psi,\partial_t\Psi)$ and $\mathcal A$ as in Ch.~11, the mild solution is
\[
Z(t)=e^{t\mathcal A}Z(0)+\int_0^t e^{(t-s)\mathcal A}\,\mathcal N(Z(s))\,ds
+\int_0^t e^{(t-s)\mathcal A}\,(0,\tau^{-1}\,dW_Q(s)),
\]
with $W_Q$ a $Q$-Wiener process and $\mathcal N(Z)=(0,\tau^{-1}\mathsf S)$.
Define the coherence energy $E_{\text{coh}}(t)$ as in Ch.~10. Under the admissibility rules (Ch.~9),
\begin{equation}
\mathbb E\,E_{\text{coh}}(t)
\;+\;\mathbb E\!\int_0^t \!\!\int_\Omega \big(\tau\,|\partial_t^2\Psi|^2+|\partial_t\Psi|^2\big)\,dx\,ds
\;\le\; C\Big(\mathbb E\,E_{\text{coh}}(0)+ t\,\mathrm{Tr}_{H} Q\Big),
\label{eq:exp_energy}
\end{equation}
so energy remains bounded in expectation with growth controlled by $\mathrm{Tr}\,Q$.

\subsection{Local stochastic models (additive, multiplicative, colored)}
While \eqref{eq:spde_core} uses additive noise, two local variants are common:
\begin{align}
\text{Multiplicative: }&
\tau\,\partial_t^2\Psi+\partial_t\Psi-c^2\nabla^2\Psi
= \mathsf S(\Psi,\varphi;\varepsilon,\delta)
+ \sigma(\Psi)\,\xi(\mathbf x,t),\label{eq:mult_noise}\\
\text{Colored-in-time: }&
\xi=\dot\eta,\quad \dot\eta=-\gamma_\eta\,\eta + \sqrt{2\gamma_\eta}\,\zeta,
\label{eq:colored}
\end{align}
where $\sigma(\Psi)$ is bounded/Lipschitz, $\gamma_\eta>0$, and $\zeta$ is white-in-time, local. Both preserve locality and, under small-gain bounds, the telegraph window $\Re s<0$.

\subsection{Residual-layer uncertainty: propagation and intervals}
Let $r=G\varepsilon-\delta$ with $W=\Sigma^{-1}\succeq0$ (Ch.~13--15).
\paragraph{Gaussian model (local covariance).}
If $\delta\sim \mathcal N(\bar\delta,\Sigma)$ and $\mathrm{rank}(G)=4$, the WLS estimator
\[
\hat\varepsilon=(G^\top W G)^{-1}G^\top W \delta
\]
has local covariance
\begin{equation}
\mathrm{Cov}(\hat\varepsilon)=(G^\top W G)^{-1}.
\label{eq:cov_eps}
\end{equation}
Hence $(\hat\varepsilon-\varepsilon)^\top (G^\top W G)(\hat\varepsilon-\varepsilon)$ is asymptotically $\chi^2_4$, producing local ellipsoidal intervals.

\paragraph{Tikhonov/LM adjustment.}
For $\varepsilon_\lambda=(G^\top W G+\lambda I)^{-1}G^\top W \delta$, use the sandwich form
\[
\widehat{\mathrm{Cov}}(\varepsilon_\lambda)\approx
(G^\top W G+\lambda I)^{-1}(G^\top W G)(G^\top W G+\lambda I)^{-1},
\]
or SVD-based filter-factor propagation (Ch.~13).

\paragraph{Robust (IRLS) surrogates.}
With IRLS weights $\hat W=\mathrm{diag}(w_i/\sigma_i^2)$, adopt the Godambe/sandwich estimator
\[
\widehat{\mathrm{Cov}}(\hat\varepsilon)\approx
(G^\top \hat W G)^{-1}\Big(G^\top \hat W\,\widehat S\,\hat W G\Big)(G^\top \hat W G)^{-1},
\quad \widehat S=\sum_i r_i^2 e_i e_i^\top .
\]

\subsection{Linearized covariance for the field (local, short-time)}
Around a reference trajectory $\bar\Psi$, linearize the source $\mathsf S$ and write the first-variation system
\[
\tau\,\partial_t^2 \tilde\Psi + \partial_t\tilde\Psi - c^2\nabla^2\tilde\Psi
= \mathcal L(t)\,\tilde\Psi + \xi,
\]
with bounded local operator $\mathcal L(t)$. Let $\mathcal C(t)=\mathbb E[\tilde Z\otimes\tilde Z]$ be the covariance of $\tilde Z=(\tilde\Psi,\partial_t\tilde\Psi)$. Then formally
\[
\dot{\mathcal C}=\mathcal A \mathcal C + \mathcal C \mathcal A^\ast + \mathcal B \mathcal Q \mathcal B^\ast
\]
(Lyapunov-type equation) on a truncated modal basis (e.g., eigenmodes or grid modes), where $\mathcal B$ injects the noise into the velocity component. This gives local, short-time covariance surrogates for $\Psi$.

\subsection{Ensemble UQ (local-only)}
Define an ensemble $\{\Psi^{(s)}\}_{s=1}^S$ by sampling local noise realizations consistent with $Q$ and (if needed) local priors for sources. Report ensemble statistics:
\[
\bar\Psi=\tfrac{1}{S}\sum_s \Psi^{(s)},\qquad
\mathrm{Var}[\Psi]=\tfrac{1}{S-1}\sum_s \big(\Psi^{(s)}-\bar\Psi\big)^2,
\]
and propagate to coherence energy and locking indices:
\[
\mathbb E[E_{\text{coh}}],\ \mathrm{Var}[E_{\text{coh}}],\quad
\mathbb E[\rho_{\mathrm{lock}}],\ \mathbb E[\chi_{\mathrm{lock}}].
\]
All sampling is local and dimensionless; seeds and windows are recorded for reproducibility.

\subsection{Local Bayesian view (hat-frame, non-circular)}
At a fixed space--time location (or short window), combine a local prior $p(\varepsilon)$ with the likelihood induced by $\delta\mid \varepsilon \sim \mathcal N(G\varepsilon,\Sigma)$ to obtain
\[
p(\varepsilon\mid \delta)\ \propto\ \exp\!\Big(-\tfrac12\|G\varepsilon-\delta\|_{W}^2\Big)\,p(\varepsilon).
\]
For Gaussian priors $\mathcal N(\mu_0,\Gamma_0)$, the posterior is Gaussian with
\[
\Gamma^{-1} = G^\top W G + \Gamma_0^{-1},\qquad
\mu= \Gamma\,(G^\top W \delta + \Gamma_0^{-1}\mu_0).
\]
All are computed locally; no global $K$ is introduced.

\subsection{Uncertainty on diagnostics and locking}
Report confidence intervals for
\[
E_{\text{coh}},\quad V=\tfrac12 r^\top W r,\quad
\rho_{\mathrm{lock}},\ \chi_{\mathrm{lock}},
\]
via ensemble quantiles or linearized variance propagation. In the locking regime, verify
\begin{equation}
\frac{d}{dt}\,\mathbb E[V]\ \le\ 0,
\qquad
\mathrm{Var}[V]\ \text{bounded on } [0,T],
\label{eq:locking_uq}
\end{equation}
modulo admissible parametric work (Ch.~10).

\subsection{Calibration to $\Omega_0$ and $W_0$ (Option A)}
Frequencies are reported in rad/s with anchor $\Omega_0$; use $f_0=\Omega_0/(2\pi)$ only for figures. Decompose
\[
W=W_0\,\widetilde W,\qquad \Lambda=\lambda_0\,\widetilde\Lambda,
\]
and choose $(W_0,\lambda_0)$ so that (i) whitened residuals satisfy $\mathbb E[\|W^{1/2}r\|_2^2]\approx m$ per window, and (ii) desired mean decay of $V$ is achieved.

\subsection{Falsifiable UQ tests (governance-compliant)}
\begin{enumerate}
\item \textbf{Energy in expectation:} with fixed $Q$, verify \eqref{eq:exp_energy} by measuring $\mathbb E[E_{\text{coh}}(t)]$ over ensembles; \emph{fail} if growth exceeds the predicted $C\,t\,\mathrm{Tr}\,Q$ scaling.
\item \textbf{Coverage of $\hat\varepsilon$:} for Gaussian data, check empirical coverage of ellipsoids from \eqref{eq:cov_eps}; \emph{fail} if under/over-coverage persists beyond tolerance.
\item \textbf{IRLS robustness:} contaminate a fraction of $\delta$; IRLS-based sandwich covariance should stabilize intervals (bounded influence); \emph{fail} if intervals collapse or inflate uncontrollably.
\item \textbf{Ensemble consistency:} increase $S$; monitor convergence of ensemble moments (LLN). \emph{Fail} if moments drift with $S$ beyond sampling error.
\end{enumerate}

\subsection{Reproducibility kit (local-only)}
For each UQ run, publish: noise seeds, window definitions, $Q$ (or its discretization), $(W_0,\lambda_0)$, solver options (CN/IMEX/Newmark), and time series of
\[
\{E_{\text{coh}},\,V,\,\rho_{\mathrm{lock}},\,\chi_{\mathrm{lock}}\}
\]
with ensemble means and quantiles.

\paragraph{Takeaway.}
Local stochastic models compatible with the telegraph core and the residual algebra provide quantitative, falsifiable uncertainty bounds. Expectation energy control, ensemble or linearized covariance, and robust residual statistics yield a portable UQ practice that remains dimensionless, K-free, and non-circular from data to fields.
% ===================== End Chapter 23 =====================

% ===================== Chapter 24 =====================
\section{Control \& Design in Residual Space: Targeting $\delta$ and Shaping Dynamics}

\subsection{Objectives and separation of roles (non-circular)}
We distinguish two local control channels at space--time point $(\mathbf x,t)$:
\begin{enumerate}
\item \textbf{Residual-space actuation} $u_r$: adjusts local data handling or actuator knobs that influence the \emph{measured} deviation $\delta$ (e.g., local gain/offset in $\mathcal M$, selection of features $\mathcal O[\Psi]$, or sensor-side dithers), without referencing $\varepsilon$ in the definition of $\delta$.
\item \textbf{Field-space actuation} $u_f$: modulates admissible source gains in the telegraph PDE (e.g., $\alpha_1,\alpha_3,\beta_{\mathrm{pm}},\alpha_{\mathrm{cpl}},D,\eta$), keeping small-gain/saturation bounds so that $\Re s<0$ and $\dot E_{\text{coh}}\le0$ (modulo source work).
\end{enumerate}
This separation upholds \emph{non-circularity}: $\delta=\mathcal M(\mathcal O[\Psi])+\eta$ depends on $\Psi$ only; $\varepsilon$ is inferred from $\delta$ by algebra (Chs.~4, 13).

\subsection{Residual-space control: target tracking for $r:=G\varepsilon-\delta$}
Let $r(\varepsilon)=G\varepsilon-\delta$, $V(\varepsilon)=\tfrac12 r^\top W r$, and fix a local reference $r_{\mathrm{ref}}$ (often $0$). We use a \emph{weighted} error
\[
e_V := r-r_{\mathrm{ref}},\qquad V_e:=\tfrac12\,e_V^\top W e_V.
\]
A continuous-time descent controller (cf.\ Ch.~5 \& 14) is
\begin{equation}
\dot\varepsilon = -\Lambda\,G^\top W\,e_V, \qquad \Lambda\succeq 0,
\label{eq:residual_feedback}
\end{equation}
yielding $\dot V_e = -e_V^\top W G \Lambda G^\top W e_V \le 0$.
Constraints (box/equality) on $\varepsilon$ are enforced by projection as in
Ch.~22. A discrete implementation uses projected/line-searched steps.

\paragraph{Actuation via $u_r$.}
If local actuation $u_r$ modifies the measurement map (e.g., a local gain or bias that is part of instrumentation), parameterize
\[
\delta=\mathcal M(\mathcal O[\Psi];u_r)+\eta
\]
with $u_r$ \emph{not} depending on $\varepsilon$. A simple descent on $u_r$ is
\begin{equation}
\dot u_r=-\Gamma_r\,\partial_{u_r}\big(\tfrac12\|G\varepsilon-\mathcal M(\mathcal O[\Psi];u_r)\|_W^2\big),
\qquad \Gamma_r>0,
\label{eq:ur_descent}
\end{equation}
which reduces $V$ while keeping the definition of $\delta$ non-circular.

\subsection{Field-space shaping: admissible telegraph modulation}
The telegraph core with sources reads (Ch.~6, 9)
\begin{equation}
\tau\,\partial_t^2\Psi+\partial_t\Psi-c^2\nabla^2\Psi
= \mathsf S(\Psi,\varphi;\varepsilon,\delta;u_f),
\label{eq:telegraph_controlled}
\end{equation}
where $u_f$ collects local gains subject to admissibility (Lipschitz, passivity).
A Lyapunov-compatible policy is a small-gain schedule
\begin{equation}
u_f(t)=\mathcal U\big(\Psi,\partial_t\Psi,\nabla\Psi;\ \rho_{\mathrm{lock}},\chi_{\mathrm{lock}}\big)
\quad\text{s.t.}\quad
\dot E_{\text{coh}}\le 0\ \text{and}\ \Re s<0.
\label{eq:uf_policy}
\end{equation}
Examples: (i) increase $\alpha_3$ when $|\Psi|$ grows (saturation); (ii) reduce
$\beta_{\mathrm{pm}}$ under large $|\partial_t\Psi|$ (anti-runaway); (iii) tune
$\alpha_{\mathrm{cpl}}$ to achieve phase locking (Ch.~8) without violating \eqref{eq:uf_policy}.

\subsection{Local MPC (finite-horizon) in the hat frame}
Define a local horizon $H>0$ and cost
\begin{equation}
\mathcal J = \int_{t}^{t+H}\!\!\Big(
\underbrace{\tfrac12\|r(\tau)-r_{\mathrm{ref}}\|_W^2}_{\text{residual}}
+\underbrace{\mu_E\,\mathscr D_E(\tau)}_{\text{field dissipation}}
+\underbrace{\mu_u\,\|u_f(\tau)\|^2}_{\text{effort}}
\Big)\,d\tau,
\qquad \mathscr D_E:= -\dot E_{\text{coh}}\ \ (\ge0),
\label{eq:mpc_cost}
\end{equation}
with weights $\mu_E,\mu_u\ge 0$. The MPC problem is
\[
\min_{u_f(\cdot),\,u_r(\cdot)} \mathcal J
\ \text{ subject to }\ 
\eqref{eq:telegraph_controlled},\ \text{admissibility (Ch.~9--11)},\
\text{constraints/gauges (Ch.~22)}.
\]
A \emph{local} solver uses the semi-discrete model (Ch.~12) and projected
gradient or SQP on the short horizon, then applies the first control slice.

\subsection{Passivity-based design: small-gain synthesis}
For the linearized telegraph operator with dissipative sources, the map
$(\mathsf S\mapsto \partial_t\Psi)$ is passive with respect to
$\langle \mathsf S, \partial_t\Psi\rangle$. Choose $u_f$ so that the active parts
are bounded by the intrinsic damping:
\begin{equation}
\|\mathsf S_{\text{active}}\|_{L^\infty}
\ \le\ \gamma_\star \|\Psi\|,
\qquad \gamma_\star < \gamma_{\min}(\tau,c),
\label{eq:small_gain_synthesis}
\end{equation}
ensuring $\dot E_{\text{coh}}\le 0$. This provides a constructive, falsifiable
recipe for $u_f$.

\subsection{Design under constraints and gauges}
Constraints (box/equality) on $\varepsilon$ and bounds on $u_f,u_r$ are enforced
by projection (Ch.~22). Gauges (e.g., reference phase $\theta(\Psi)$) do not
affect cost or stability since only differences appear. The Option~A rad/s
convention is maintained; Hz appears only in figures via $f_0=\Omega_0/(2\pi)$.

\subsection{Calibration of $(W_0,\lambda_0)$ for control}
Decompose $W=W_0\widetilde W$ and $\Lambda=\lambda_0\widetilde\Lambda$ (Ch.~10, 15).
Tuning rules:
\[
W_0:\ \mathbb E[\|W^{1/2}r\|_2^2]\approx m\ \ \Rightarrow\ \ \text{well-scaled residual penalty};
\qquad
\lambda_0:\ \text{sets the closed-loop decay rate of }V.
\]
Both are scalar and local; they do not change dimensionless algorithms.

\subsection{Algorithms (local templates)}
\paragraph{Projected residual--field coordination (one step).}
\begin{enumerate}
\item \textbf{Residual step:} update $\varepsilon\leftarrow \Pi_{\mathcal C}(\varepsilon-\eta\,\Lambda G^\top W e_V)$.
\item \textbf{Field step:} update $u_f$ by a projected descent on $\mathcal J$
or by rule \eqref{eq:small_gain_synthesis}; integrate \eqref{eq:telegraph_controlled}
one time step with CN/IMEX (Ch.~12).
\item \textbf{Actuator step:} update $u_r$ via \eqref{eq:ur_descent} (project to bounds).
\end{enumerate}

\paragraph{Local MPC (receding horizon).}
\begin{enumerate}
\item Linearize/discretize on $[t,t+H]$; form gradients of \eqref{eq:mpc_cost}.
\item Solve a projected SQP or gradient step; apply first control increment.
\item Advance the horizon and repeat.
\end{enumerate}

\subsection{Falsifiable tests (governance-compliant)}
\begin{enumerate}
\item \textbf{Lyapunov decrease:} in closed loop, verify
$\Delta V\le 0$ (residual) and $\dot E_{\text{coh}}\le 0$ (field) modulo admissible source work.
\item \textbf{Ablation:} remove saturation or violate small-gain;
expect energy growth or loss of $\Re s<0$; restore to pass.
\item \textbf{Constraint compliance:} log violations of bounds on $(\varepsilon,u_f,u_r)$;
\emph{pass} if none occur under the projected updates.
\item \textbf{Reference tracking:} set $r_{\mathrm{ref}}=0$; report convergence rate
vs.\ $(W_0,\lambda_0)$ tuning; \emph{fail} if rates contradict predictions.
\end{enumerate}

\subsection{Representative outcomes (qualitative)}
\begin{itemize}
\item \textbf{Targeted residuals:} $r\to r_{\mathrm{ref}}$ with monotone $V$.
\item \textbf{Stable field shaping:} admissible $u_f$ preserves $\Re s<0$ and enforces
bounded, coherent envelopes; discrete energy is nonincreasing.
\item \textbf{Locking-aware control:} when $\rho_{\mathrm{lock}},\chi_{\mathrm{lock}}$ are high,
residual energy decreases faster and smaller control effort is required.
\end{itemize}

\paragraph{Takeaway.}
By coordinating a residual-space descent with an admissible, passivity-aware
field actuation, one obtains a dimensionless, local-only control layer that
\emph{targets} the measured deviation $\delta$ and \emph{shapes} coherent
dynamics without violating telegraph stability, the Energy--Coherence Law,
or non-circularity of the measurement loop.
% ===================== End Chapter 24 =====================
